### 2017-01-20 hw01/hw01.rnw
Homework 1

\documentclass[11pt]{article}
\usepackage{fullpage,graphicx,float,amsmath,enumitem,fancyhdr}

\setlist{parsep=5.5pt}
\setlength{\parindent}{0pt}

\newcommand\assignment{Stat 534 Homework 1}
\newcommand\myname{Kenny Flagg}
\newcommand\duedate{January 18, 2017}

\pagestyle{fancy}
\lhead{\assignment}
\chead{\duedate}
\rhead{\myname}
\setlength{\headheight}{18pt}
\setlength{\headsep}{2pt}

\title{\assignment}
\author{\myname}
\date{\duedate}

\begin{document}
\maketitle

<<setup, echo = FALSE, message = FALSE, cache = FALSE>>=
library(knitr)
library(extrafont)
opts_chunk$set(echo = TRUE, comment = NA, message = FALSE,
               show.signif.stars = FALSE,
               fig.align = 'center', fig.width = 6.5, fig.height = 3,
               fig.pos = 'H', size = 'footnotesize', dev = 'pdf',
               dev.args = list(family = 'CM Roman', pointsize = 11))
knit_theme$set('print')

library(xtable)
options(xtable.floating = FALSE,
        xtable.sanitize.rownames.function = function(x) x,
        xtable.sanitize.colnames.function = function(x) x,
        width = 80, scipen = 2, show.signif.stars = FALSE)

library(spdep)
set.seed(-8125)
@

\begin{enumerate}

\item%1
{\it Our text implies and others state outright that the \(BB\), \(BW\), and
\(WW\) statistics reveal pretty much the same thing about spatial correlation.
The \texttt{joincount.mc} function will carry out Monte Carlo tests based on
the \(BB\) and \(WW\) statistics. We do not have an R formula for computing the
\(BW\) statistic but it is possible to carry out a \(BW\) joincount test of
spatial autocorrelation (or clustering) using Geary's \(c\).}

\begin{enumerate}

\item%a
{\it Show the relationship between Geary's \(c\) and \(BW\).}

\begin{align*}
BW&=\frac{1}{2}\sum_{i=1}^{n}\sum_{j=1}^{n}w_{ij}
(Z(\mathbf{s}_{i})-Z(\mathbf{s}_{j}))^{2} \\
&=S^{2}w_{..}\left(\frac{1}{2S^{2}w_{..}}\sum_{i=1}^{n}\sum_{j=1}^{n}w_{ij}
(Z(\mathbf{s}_{i})-Z(\mathbf{s}_{j}))^{2}\right) \\
&=S^{2}w_{..}c
\end{align*}

\item%b
{\it Carry out a test based on the \(BW\) statistics using \texttt{geary.mc}.
The data file \texttt{atrplx.dat} will be emailed to you at your math
department email addresses. The first 2 columns contain the spatial coordinates
and the fourth column contains the \(Z\) values you need. Use the R handout to
generate the necessary neighbors and list objects.}

<<prob1b, fig.height = 5>>=
# Read data.
atrplx <- read.table('atrplx.dat', col.names = c('X', 'Y', 'ignore', 'Z'))

# Get rook adjacency list.
atrplx.nb <- dnearneigh(as.matrix(atrplx[, c('X', 'Y')]), d1 = 0, d2 = 1)

# Augment with binary-style w[ij].
atrplx.lw <- nb2listw(atrplx.nb, style = 'B')

# Now do the test.
atrplx.c <- geary.mc(atrplx$Z, atrplx.lw, 999)
atrplx.c

# Make a plot.
hist(atrplx.c$res, breaks = 100, freq = FALSE,
     xlab = 'c', main = 'Permutation Distribution for Geary\'s c')
abline(v = atrplx.c$statistic, lwd = 2)
@

A permutation test with 1,000 permutations results in a p-value of
\Sexpr{atrplx.c$p.value}, which is not convincing evidence the the plants are
spatially clustered.

\item%c
{\it Using the output from \texttt{geary.mc} compute \(BW\) and \(E[BW]\). Do
you expect \(BW < E[BW]\) or \(BW > E[BW]\) in the presence of positive spatial
clustering of the plants? Why or why not?}

<<prob1c>>=
# Sample variance.
var(atrplx$Z)

# Neighbor information.
atrplx.lw

# Mean of the permutation distribution.
mean(atrplx.c$res)
@

We have \(S^{2}=\Sexpr{signif(var(atrplx$Z), 5)}\), and the number of nonzero
links tells us that \(w_{..}=\Sexpr{sum(card(atrplx.nb))}\). So
\begin{equation*}
BW=\Sexpr{signif(var(atrplx$Z), 5)}\times\Sexpr{sum(card(atrplx.nb))}
\times\Sexpr{signif(atrplx.c$statistic, 5)}
=\Sexpr{var(atrplx$Z) * sum(card(atrplx.nb)) * atrplx.c$statistic}.
\end{equation*}
Then the expected value is
\begin{equation*}
BW=\Sexpr{signif(var(atrplx$Z), 5)}\times\Sexpr{sum(card(atrplx.nb))}
\times\Sexpr{signif(mean(atrplx.c$res), 5)}
=\Sexpr{signif(var(atrplx$Z) * sum(card(atrplx.nb)) * mean(atrplx.c$res), 5)}.
\end{equation*}

Under positive spatial clustering I would expect \(BW < E[BW]\) because
clustering will place quadrats with plants together, leading to more
black-black and white-white joins, and fewer black-white joins than would
occur under if plants were placed completely randomly.

\item%d
{\it Reproduce the analysis I presented in class using the \emph{Atriplex}
data. Compare the results of the \(BW\) test to those of the \(BB\) and \(WW\)
test that were discussed in class. Do these statistics all seem to indicate the
same thing about spatial clustering of the plants?}

<<prob1d>>=
atrplx.join <- joincount.mc(factor(atrplx$Z), atrplx.lw, 999)
atrplx.join

par(mfrow = c(1, 2))
hist(atrplx.join[[1]]$res, breaks = 50, freq = FALSE,
     xlab = 'WW', main = 'Permutation Dist. for WW')
abline(v = atrplx.join[[1]]$statistic, lwd = 2)
hist(atrplx.join[[2]]$res, breaks = 50, freq = FALSE,
     xlab = 'BB', main = 'Permutation Dist. for BB')
abline(v = atrplx.join[[2]]$statistic, lwd = 2)
@

The \(WW\) test results in a p-value of \Sexpr{atrplx.join[[1]]$p.value},
no evidence of clustering. However, the \(BB\) test gives a p-value of
\Sexpr{atrplx.join[[1]]$p.value}, strong evidence of clustering. The p-value
of the \(BW\) test is in between these values. These tests do not indicate the
same thing about the distribution of the plants because the \(BB\) statistic
suggests they are clustered, by the \(WW\) and \(BW\) statistics do not.

\item%e
{\it For grins compute Moran's \(I\) and compare that result to those above.}

<<prob1e, fig.height = 5>>=
atrplx.i <- moran.mc(atrplx$Z, atrplx.lw, 999)
atrplx.i

hist(atrplx.i$res, breaks = 50, freq = FALSE,
     xlab = 'I', main = 'Permutation Dist. for Moran\'s I')
abline(v = atrplx.i$statistic, lwd = 2)
@

With 1,000 permutations, we get a p-value of \Sexpr{atrplx.i$p.value}, which
gives weak evidence of spatial clustering. This is not as convincing as the
result of the test based on \(BB\), and definitely inconsisitent with the
tests based on \(WW\) and \(BW\).

\end{enumerate}

\pagebreak
\item%2
{\it Categorize the following examples of spatial data as to their data type:}

\begin{enumerate}

\item%a
{\it Elevations in the foothills of the Allegheny mountains.}

\textbf{These are geostatistical data.}

\item%b
{\it Highest elevation within each state in the United States.}

\textbf{These are lattice data.}

\item%c
{\it Concentration of a mineral in soil.}

\textbf{These are geostatistical data.}

\item%d
{\it Plot yields in a uniformity trial.}

\textbf{These are lattice data.}

\item%e
{\it Crime statistics giving names of subdivisions where break-ins occurred in
the previous year and property loss values.}

If the list only includes the subdivisions where crimes occurred then the list
itself is random, so \textbf{these are marked point process data.}

\item%f
{\it Same as previous, but instead of the subdivisions, the individual
dwelling is identified.}

\textbf{These are marked point process data.}

\item%g
{\it Distribution of oaks and pines in a forest stand.}

\textbf{These are marked point process data.}

\end{enumerate}

\item%3
{\it Show that Moran's \(I\) is a scale-free statistic, i.e. \(Z(s)\) and
\(\lambda Z(s)\) yield the same value for any constant \(\lambda \neq 0\).}

First of all, the mean of \(\lambda Z(\mathbf{s})\) is
\begin{equation*}
\frac{\sum_{i=1}^{n}\lambda Z(\mathbf{s}_{i})}{n}=\lambda\overline{Z}
\end{equation*}
and the sample variance of \(\lambda Z(\mathbf{s})\) is
\begin{align*}
\frac{\sum_{i=1}^{n}(\lambda Z(\mathbf{s}_{i})-\lambda\overline{Z})^{2}}{n-1}
=\lambda^{2}\frac{\sum_{i=1}^{n}(Z(\mathbf{s}_{i})-\overline{Z})^{2}}{n-1}
=\lambda^{2}S^{2}.
\end{align*}
Then Moran's \(I\) for \(\lambda Z(\mathbf{s})\) is
\begin{align*}
\frac{n\sum_{i=1}^{n}\sum_{j=1}^{n}w_{ij}
(\lambda Z(\mathbf{s}_{i})-\lambda\overline{Z})
(\lambda Z(\mathbf{s}_{j})-\lambda\overline{Z})}
{(n-1)\lambda^{2}S^{2}w_{..}}
&=\frac{n\sum_{i=1}^{n}\sum_{j=1}^{n}w_{ij}\lambda
(Z(\mathbf{s}_{i})-\overline{Z})\lambda
(Z(\mathbf{s}_{j})-\overline{Z})}
{(n-1)\lambda^{2}S^{2}w_{..}} \\
&=\frac{n\lambda^{2}\sum_{i=1}^{n}\sum_{j=1}^{n}w_{ij}
(Z(\mathbf{s}_{i})-\overline{Z})
(Z(\mathbf{s}_{j})-\overline{Z})}
{(n-1)\lambda^{2}S^{2}w_{..}} \\
&=\frac{n\sum_{i=1}^{n}\sum_{j=1}^{n}w_{ij}
(Z(\mathbf{s}_{i})-\overline{Z})
(Z(\mathbf{s}_{j})-\overline{Z})}
{(n-1)S^{2}w_{..}} \\
&= I
\end{align*}
which is Moran's \(I\) for \(Z(\mathbf{s})\), so Moran's \(I\) is scale-free.

\item%4
{\it Let \(Y_{1},\dots, Y_{n}\) be normally distributed with unknown mean
\(\mu\) and known variance \(\sigma^{2}\). Let
\(Cov(Y_{i}, Y_{j}) = \sigma^{2}\rho\) for \(i \neq j\). We will further assume
that \(ρ > 0\).}

\begin{enumerate}

\item%a
{\it Show that} \(\displaystyle
Var\left(\overline{Y}\right)=\frac{\sigma^{2}}{n}\left[1+(n-1)\rho\right]\).

Note that there are \(n^{2}-n\) pairs \((i,j)\) such that \(i\neq j\). So,
\begin{align*}
Var\left(\overline{Y}\right)&=Var\left(\frac{\sum_{i=1}^{n}}{n}\right) \\
&=\frac{Var\left(\sum_{i=1}^{n}\right)}{n^{2}} \\
&=\frac{\sum_{i=1}^{n}\sum_{j=1}^{n}Cov(Y_{i},Y_{j})}{n^{2}} \\
&=\frac{\sum_{i=1}^{n}Var(Y_{i})+\sum_{i\neq j}Cov(Y_{i},Y_{j})}{n^{2}} \\
&=\frac{n\sigma^{2}+(n^{2}-n)\sigma^{2}\rho}{n^{2}} \\
&=\frac{\sigma^{2}}{n}[1+(n-1)\rho].
\end{align*}

\item%b
{\it Let \(n = 10\) and \(\rho = 0.26\). Compare and contrast a 95\% confidence
interval for \(\mu\) computed using the true standard deviation of
\(\overline{Y}\) and one computed assuming independence.}

\textbf{Using the true standard deviation:}
\begin{equation*}
\overline{Y}\pm 1.96\frac{\sigma^2}{10}\left[1+9\times 0.26\right]
=\overline{Y}\pm 0.65464\sigma^2
\end{equation*}

\textbf{Assuming independence:}
\begin{equation*}
\overline{Y}\pm 1.96\frac{\sigma^2}{10}
=\overline{Y}\pm 0.196\sigma^2
\end{equation*}

The confidence interval computed using the true standard deviation is quite a
bit wider than the confidence interval computed assuming independence. Ignoring
the correlation in this case means we will think our estimates are more
precise than they actually are.

\item%c
{\it Given independence, we know that \(\overline{Y}\) is the ``best''
estimator of \(\mu\). One nice property it has is that it is a consistent
estimator of the mean. Is \(\overline{Y}\) a consistent estimator of the mean
given the correlation structure above? Justify your answer.}

No, \(\overline{Y}\) is not consistent in this situation. Consistency is
convergence in probability to the quantity of interest, which, by definition
5.5.1 in Casella and Berger, is
\begin{equation*}
\lim_{n\to\infty}P\left(|\overline{Y}-\mu|\geq\epsilon\right)=0
\end{equation*}
for all \(\epsilon>0\). This is not possible because
\begin{equation*}
\lim_{n\to\infty}Var(\overline{Y})
=\lim_{n\to\infty}\frac{\sigma^{2}}{n}[1+(n-1)\rho]
=\frac{\sigma^{2}\rho}{n}
\end{equation*}
is greater than zero if \(\rho>0\).

\item%d
{\it Recall that effective sample size is a measure of the effect of
correlation on inference. An equation for the effective sample size under the
equicorrelation model is}
\begin{equation*}
n'=\frac{n}{1+(n-1)\rho}
\end{equation*}
{\it The effective sample size is defined to be the sample size \(n\) of
uncorrelated observations that provide the same information (in a sense) as a
sample of \(n\) correlated observations.}

\begin{enumerate}

\item%i
{\it Compute the effective sample size when \(n\) = 10, 100, and 1000 and
\(\rho\) = 0.05, 0.1, 0.25, and 0.5.}

<<prob4di, eval = FALSE>>=
n <- matrix(rep(c(10, 100, 1000), 4), ncol = 4,
            dimnames = list(paste('\\(n\\) =', c(10, 100, 1000)),
                            paste('\\(\\rho\\) =', c(0.05, 0.1, 0.25, 0.5))))
rho <- matrix(rep(c(0.05, 0.1, 0.25, 0.5), each = 3), ncol = 4,
              dimnames = list(paste('\\(n\\) =', c(10, 100, 1000)),
                              paste('\\(\\rho\\) =', c(0.05, 0.1, 0.25, 0.5))))
nprime <- n / (1 + (n - 1) * rho)
xtable(nprime, digits = 4, align = '|r|rrrr|')
@
{\centering
<<prob4di, echo = FALSE, results = 'asis'>>=
@
}

\item%ii
{\it Find \(\lim n'\) as \(n \to \infty\).}

Solution:
\begin{align*}
\lim_{n\to\infty}n'&=\lim_{n\to\infty}\frac{n}{1+(n-1)\rho} \\
&=\lim_{n\to\infty}\frac{n}{1+n\rho-\rho} \\
&=\frac{1}{\rho}
\end{align*}

\item%iii
{\it The effect is extreme here but we would not expect to see this type of
correlation structure in a spatial setting. Why not?}

Tobler's First Law of Geography: ``Everything is related to everything else,
but near things are more related than distant things.'' The setting in this
problem ignores distance, making distant locations just as correlated as nearby
locations. This is not how things usually work in reality.

\end{enumerate}

\end{enumerate}

\end{enumerate}

\end{document}

### 2017-01-26 hw02/hw02.rnw
hw2

\documentclass[11pt]{article}
\usepackage{fullpage,graphicx,float,amsmath,enumitem,fancyhdr}

\setlist{parsep=5.5pt}
\setlength{\parindent}{0pt}

\newcommand\assignment{Stat 534 Homework 2}
\newcommand\myname{Kenny Flagg}
\newcommand\duedate{January 26, 2017}

\pagestyle{fancy}
\lhead{\assignment}
\chead{\duedate}
\rhead{\myname}
\setlength{\headheight}{18pt}
\setlength{\headsep}{2pt}

\title{\assignment}
\author{\myname}
\date{\duedate}

\begin{document}
\maketitle

<<setup, echo = FALSE, message = FALSE, cache = FALSE>>=
library(knitr)
library(extrafont)
opts_chunk$set(echo = TRUE, comment = NA, message = FALSE,
               show.signif.stars = FALSE,
               fig.align = 'center', fig.width = 6.5, fig.height = 3,
               fig.pos = 'H', size = 'footnotesize', dev = 'pdf',
               dev.args = list(family = 'CM Roman', pointsize = 11))
knit_theme$set('print')

library(xtable)
options(xtable.floating = FALSE,
        xtable.sanitize.rownames.function = function(x) x,
        xtable.sanitize.colnames.function = function(x) x,
        width = 80, scipen = 2, show.signif.stars = FALSE)

set.seed(6815)
library(spatstat)
data(redwood)
@

\begin{enumerate}

\item%1
{\it It was pointed out in class that, conditional on \(n\) events, event
locations are uniformly distributed for a homogeneous Poisson process. We will
consider the simplest example of this. Consider a one-dimensional process on a
transect of length \(L\), \((0, L]\). Given that one event has occurred on the
interval \((0, L]\) what is the probability that it occurred in the subinterval
\((0, s]\) for \(s < L\)?}

Let \(X\) be the location of the event. Then
\(X|N((0, L]) = 1\sim\mathrm{Unif}(0, L)\). So
\begin{equation*}
P(0 < X \leq s|N((0, L]) = 1) = \int_{0}^{s}\frac{1}{L}ds = \frac{s}{L}.
\end{equation*}

\item%2
{\it We looked at an example of \texttt{quadrat.test} on the amacrine data set
in class. We will use it to analyze another data set, called \texttt{redwood}.
You can read about this in the \texttt{spatstat} help material. You will be
using the \texttt{quadrat.test} function. You can also read about this function
in the help material.}

\begin{enumerate}

\item%a
{\it Read the help pages on the \texttt{quadrat.test} function. What null
hypothesis do they claim to be testing?}

<<prob2a, eval = FALSE>>=
require(spatstat)
data(redwood)
help(redwood) # optional information on the data set
help(quadrat.test)
@

The help file says pretty clearly that ``we test the null hypothesis that the
data pattern is a realisation of Complete Spatial Randomness (the uniform
Poisson point process).''

\item%b
{\it Use \texttt{quadrat.test} on the redwood data set.}

<<prob2b1>>=
redwood.fit <- quadrat.test(redwood)
redwood.fit
@

The quadrat test on a \(5 \times 5\) grid yields a test statistic of
\(\chi^{2}_{\Sexpr{redwood.fit$parameter}}
= \Sexpr{sprintf('%.3f', redwood.fit$statistic)}\) with a p-value of
\(\Sexpr{format.pval(redwood.fit$p.value, 4, 0.0001)}\), very strong evidence
that the locations of the redwood trees do not follow complete spatial
randomness.

{\it The default partitioning of the grid is 5 \(\times\) 5. Does that appear
appropriate here? Justify your answer.}

R gave a warning that some of the grid cells had small expected counts, so the
\(5 \times 5\) grid is too fine for these data. Bigger grid cells would be
better.

\item%c
{\it Redo the analysis using a 3 \(\times\) 3 grid.}

<<prob2c>>=
redwood.fit <- quadrat.test(redwood, nx = 3, ny = 3)
redwood.fit
@

This time we don't get a warning message, so the expected cell counts in the
larger grid cells are big enough to use the \(\chi^{2}\) distribution. The
quadrat test on a \(3 \times 3\) grid yields a test statistic of
\(\chi^{2}_{\Sexpr{redwood.fit$parameter}}
= \Sexpr{sprintf('%.3f', redwood.fit$statistic)}\) with a p-value of
\(\Sexpr{sprintf('%.4f', redwood.fit$p.value)}\), still very strong evidence
that the locations of the redwood trees do not follow complete spatial
randomness.

\item%d
{\it Is the value of the test statistic \(X^{2}\) indicative of clustering,
CSR, or a regular pattern? Justify your answer. Note that I am only asking you
to compare the observed value of \(X^{2}\) to what you would expect under each
of these three patterns. You do not need to calculate a P-value just yet.}

Under the null hypothesis, the test statistic follows a
\(\chi^{2}_{\Sexpr{redwood.fit$parameter}}\) with mean
\Sexpr{redwood.fit$parameter}. The observed value of
\Sexpr{sprintf('%.3f', redwood.fit$statistic)} is larger, indicating greater
than excepted variability in quadrat counts which suggests clustering.

\item%e
{\it The investigator suspected a clustered pattern and the plot would seem to
be consistent with this. Rerun the test with \texttt{alternative="c"} in the
argument list. Give the p-value for the test and interpret the results. Does
the test provide evidence against CSR and for clustering? Justify your answer.}

<<prob2d>>=
redwood.fit <- quadrat.test(redwood, nx = 3, ny = 3, alternative = 'c')
redwood.fit
@

The quadrat test on a \(3 \times 3\) grid with a test statistic of
\(\chi^{2}_{\Sexpr{redwood.fit$parameter}}
= \Sexpr{sprintf('%.3f', redwood.fit$statistic)}\) has a one-sided p-value of
\(\Sexpr{sprintf('%.4f', redwood.fit$p.value)}\), very strong evidence that
the locations of the redwood trees are clustered.

\item%f
{\it We can plot the results of the fit.}

<<prob>>=
par(mar = c(0, 0, 0.5, 0))
plot(redwood.fit)
points(redwood, pch = 16, col = '#00000040') # Add semitransparent points for fun!
@

{\it You will see a plot of the 3 \(\times\) 3 grid. There are 3 numbers in
each cell: the observed count (upper left), expected count under CSR (upper
right), and a scaled residual (lower number). The sum of the scaled residuals
is the \(X^{2}\) statistic. Give the results of the test and using the plot
indicate where CSR seems to break down, if it does.}

The one-sided p-value of \(\Sexpr{sprintf('%.4f', redwood.fit$p.value)}\)
provides very strong evidence that trees are clustered. The upper-left and
center-right cells have fewer trees than expected under CSR, with
large-magnitude negative residuals. The center-left and upper-right cells
have large positive residuals, indicating that these cells contain relatively
dense clusters.

\pagebreak
\item%g
{\it Quadrat size can be important. Repeat the analysis using a 2 \(\times\) 2
grid. Give the results and compare to what we saw with the 3 \(\times\) 3
grid.}

<<prob2g>>=
redwood.fit <- quadrat.test(redwood, nx = 2, ny = 2)
redwood.fit
@

The quadrat test on a \(2 \times 2\) grid yields a test statistic of
\(\chi^{2}_{\Sexpr{redwood.fit$parameter}}
= \Sexpr{sprintf('%.3f', redwood.fit$statistic)}\) with a two-sided p-value of
\(\Sexpr{sprintf('%.4f', redwood.fit$p.value)}\), which is no evidence against
complete spatial randomness and contradicts the results of the test on the
\(3 \times 3\) grid.

\end{enumerate}

\item%3
{\it We will compare results from Monte Carlo procedures based on Poisson
sampling and based on conditioning on the number of observed points. We will
use the \texttt{cells} data set. The R code to accomplish that is shown below.
Compare the two procedures. What do they indicate about the spatial pattern and
why? Which procedure do you like best for this data set and why?}

<<prob3, fig.height = 5.5>>=
data(cells)
hbar <- mean(nndist(cells))
hbar
hbar.pois <- rep(0, 1000)
hbar.cond <- rep(0, 1000)
hbar.pois[1] <- hbar
hbar.cond[1] <- hbar
for(i in 2:1000){
  # Poisson Monte Carlo
  dat.pois <- rpoispp(42)
  hbar.pois[i] <- mean(nndist(dat.pois))
  # Conditional Monte Carlo
  dat.cond <- runifpoint(42)
  hbar.cond[i] <- mean(nndist(dat.cond))
}
par(mfrow = c(2, 1))
hist(hbar.pois, prob = TRUE, xlim = c(0.05, 0.13), main = 'Poisson Monte Carlo')
abline(v = hbar)
hist(hbar.cond, prob = TRUE, xlim = c(0.05, 0.13), main = 'Conditional Monte Carlo')
abline(v = hbar)
# Poisson P-value
2 * sum(hbar.pois >= hbar) / 1000
# Conditional P-value
2 * sum(hbar.cond >= hbar) / 1000
@

Both methods give similar p-values of about 0.002, very strong evidence against
complete spatial randomness. (The observed average nearest neighbor distance
is larger than expected under the null, implying regularity.) The null
distribution for the Poison Monte Carlo method is has a larger spread than the
null distribution for the conditional Monte Carlo method, reflecting
variability in the number of events. For this dataset, the conditional method
is more appropriate because the researcher chose the cells to look at and then
rescaled the viewing window to the unit square. Therefore it is appropriate to
consider permutations of these 42 cells rather than simulating new realizations
with random numbers of cells.

\item%4
{\it Below is the frequency distribution of the number of trees per quadrat in
a sample of 100 quadrats each of radius 6 m.}
\begin{center}\begin{tabular}{c|cccccc}
Trees per quadrat & 0 & 1 & 2 & 3 & 4 & \(\geq 5\) \\
\hline
Count & 34 & 33 & 17 & 7 & 3 & 6
\end{tabular}\end{center}
{\it The data were pooled for counts \(\geq 5\) to meet the assumptions of the
method. Carry out a Poisson goodness-of-fit test based on an assumption of CSR.
Discuss the results. The sample mean of the observed counts was 1.43.}

<<prob4>>=
# Observed number of quadrats in each bin.
Observed <- c(`0` = 34, `1` = 33, `2` = 17, `3` = 7, `4` = 3, `>=5` = 6)

# Expected number of quadrats in each bin.
Expected <- 100 * c(dpois(0:4, 1.43), ppois(4, 1.43, lower.tail = FALSE))

rbind(Observed, Expected)

X2 <- sum((Observed - Expected)^2 / Expected)
X2
pchisq(X2, 4, lower.tail = FALSE)
@

With a test statistic of \(\chi^{2}_{4}
= \Sexpr{sprintf('%.3f', X2)}\) and a p-value of
\Sexpr{sprintf('%.4f', pchisq(X2, 4, lower.tail = FALSE))}
there is very strong evidence that the quadrat counts do not come from a
\(\mathrm{Poisson}(1.43)\) distribution. The test result does not indicate
what type of spatial pattern the data follow, but we observed more counts of
zero and more counts \(\geq\) 5 than expected under CSR, which suggests
clustering.

\pagebreak
\item%5
{\it Suppose we have a realization of a spatial point process consisting of
\(N\) event locations \(\{s_{1}, s_{2},\dots, s_{N}\}\). Let \(H_{i}\) denote
the distance between the \(i\)th event and the nearest neighboring event. The
cumulative distribution function of \(H\) (the nearest event-event distance) is
the \(G\) function. (This problem will be continued on the next homework
assignment).}

\begin{enumerate}

\item%a
{\it What is the \(G\) function if the point process is CSR; i.e. what is
\(G(h) = P(H \leq h)\)?}

\begin{align*}
G(h) &= P(H \leq h) \\
&= P(\text{at least 1 event in a circle of radius }h) \\
&= 1 - P(\text{0 events in a circle of radius }h) \\
&= 1 - \frac{e^{-\lambda \pi h^{2}} \lambda^{0}}{0!} \\
&= 1 - e^{-\lambda \pi h^{2}},\quad h > 0
\end{align*}

\item%b
{\it Find the pdf of \(H\).}

\begin{align*}
f_{H}(h) = \frac{dG(h)}{dh} = 2\lambda \pi h e^{-\lambda \pi h^{2}},
\quad h > 0
\end{align*}

\item%c
{\it Find \(E(H)\) and \(Var(H)\). Hint: you found the pdf but before you start
evaluating a gnarly integral take a close look at that pdf and see if you
cannot identify the family of distributions it belongs to. If you can do that
then you can use that knowledge to find the mean and variance.}

\(H\) follows a Weibull distribution with (in the parameterization of Casella
and Berger) \(\gamma = 2\) and \(\displaystyle\beta = \frac{1}{\lambda\pi}\).
This distribution has mean
\begin{equation*}
E(H) = \beta^{\frac{1}{\gamma}}\Gamma\left(1+\frac{1}{\gamma}\right)
= \frac{1}{\sqrt{\lambda\pi}}\Gamma\left(\frac{3}{2}\right)
= \frac{1}{\sqrt{\lambda\pi}}\frac{1}{2}\Gamma\left(\frac{1}{2}\right)
= \frac{1}{\sqrt{\lambda\pi}}\frac{1}{2}\sqrt{\pi}
= \frac{1}{2\sqrt{\lambda}}
\end{equation*}
and variance
\begin{align*}
Var(H) &= \beta^{\frac{2}{\gamma}}
\left[\Gamma\left(1+\frac{2}{\gamma}\right)
-\Gamma^{2}\left(1+\frac{1}{\gamma}\right)\right] \\
&= \frac{1}{\lambda\pi}\left[\Gamma(2)
-\Gamma^{2}\left(\frac{3}{2}\right)\right] \\
&= \frac{1}{\lambda\pi}\left[1-\left(\frac{\sqrt{\pi}}{2}\right)^{2}\right] \\
&= \frac{1}{\lambda\pi}\left[1-\frac{\pi}{4}\right].
\end{align*}

\end{enumerate}

\end{enumerate}

\end{document}

### 2017-02-02 hw03/hw03.rnw
Hw3

\documentclass[11pt]{article}
\usepackage{fullpage,graphicx,float,amsmath,enumitem,fancyhdr}

\setlist{parsep=5.5pt}
\setlength{\parindent}{0pt}

\newcommand\assignment{Stat 534 Homework 3}
\newcommand\myname{Kenny Flagg}
\newcommand\duedate{February 3, 2017}

\pagestyle{fancy}
\lhead{\assignment}
\chead{\duedate}
\rhead{\myname}
\setlength{\headheight}{18pt}
\setlength{\headsep}{2pt}

\title{\assignment}
\author{\myname}
\date{\duedate}

\begin{document}
\maketitle

<<setup, echo = FALSE, message = FALSE, cache = FALSE>>=
library(knitr)
library(extrafont)
opts_chunk$set(echo = TRUE, comment = NA, message = FALSE,
               show.signif.stars = FALSE,
               fig.align = 'center', fig.width = 6.5, fig.height = 3,
               fig.pos = 'H', size = 'footnotesize', dev = 'pdf',
               dev.args = list(family = 'CM Roman', pointsize = 11))
knit_theme$set('print')

library(xtable)
options(xtable.floating = FALSE,
        xtable.sanitize.rownames.function = function(x) x,
        xtable.sanitize.colnames.function = function(x) x,
        width = 80, scipen = 2, show.signif.stars = FALSE)

set.seed(7915)
@

\begin{enumerate}

\item%1
{\it On the last homework there was some confusion about two problems. I took
off some points for one of those but am now giving you a chance to get them
back.}

\begin{enumerate}

\item%a
{\it It was pointed out in class that, conditional on n events, event
locations are uniformly distributed for a homogeneous Poisson process. Show
this result for a 1-\(d\) process. Hint: Consider a one-dimensional process on
a transect of length \(L\), \((0, L]\). Given that one event has occurred on
the interval \((0, L]\) what is the probability that it occurred in the
subinterval \((0, s]\) for \(s<L\)? Show that this is the cdf of a
\(\mathrm{Unif}(0, L)\) distribution.}

First, note that for a homogeneous Poisson process in one dimension with
intensity \(\lambda\), the number of events in an interval follows a Poisson
distribution with mean \(\lambda\) times the length of the interval, and the
numbers of events in disjoint intervals are independent. If we have a
homogeneous Poisson process on \((0, L]\) with intensity \(\lambda\) and one
event occurs at location \(S\), then, for \(0 < s < L\),
\begin{align*}
P(S \leq s) &= P(1\text{ event in }(0, s]|1\text{ event in }(0, L]) \\
&= \frac{P(1\text{ event in }(0, s]\text{ and }1\text{ event in }(0, L])}
{P(1\text{ event in }(0, L])} \\
&= \frac{P(1\text{ event in }(0, s])P(0\text{ events in }(s, L])}
{P(1\text{ event in }(0, L])} \\
&= \frac{e^{-\lambda s}\left(\lambda s\right)^{1}/1!
\times e^{-\lambda (L-s)}\left(\lambda (L-s)\right)^{0}/0!}
{e^{-\lambda L}\left(\lambda L\right)^{1}/1!} \\
&= e^{\lambda (L-s)}\frac{s}{L}e^{-\lambda (L-s)} \\
&= \frac{s}{L}
\end{align*}
is the cdf of \(S\), so \(S \sim \mathrm{Unif}(0, L)\).

\pagebreak
\item%b
{\it Suppose we have a realization of a spatial point process consisting of
\(N\) event locations \(\{s_{1},s_{2},\dots,s_{N}\}\). Let \(H_{i}\) denote
the distance between the \(i\)th event and the nearest neighboring event. The
cumulative distribution function of \(H\) (the nearest event-event distance)
is the \(G\) function. (This problem will be continued on the next homework
assignment). Derive the \(G\) function if the point process is CSR; i.e. what
is \(G(h)=P(H\leq h)\).}

The \(G\) function is
\begin{align*}
G(h) &= P(H \leq h) \\
&= P(\text{at least 1 event in a circle of radius }h) \\
&= 1 - P(\text{0 events in a circle of radius }h) \\
&= 1 - \frac{e^{-\lambda \pi h^{2}} \left(\lambda\pi h^{2}\right)^{0}}{0!} \\
&= 1 - e^{-\lambda \pi h^{2}},\quad h > 0.
\end{align*}
The pdf of \(H\) is
\begin{align*}
g(h) = \frac{dG(h)}{dh} = 2\lambda \pi h e^{-\lambda \pi h^{2}},
\quad h > 0.
\end{align*}

\end{enumerate}

\item%2
{\it We looked at one simple method of using nearest neighbor distances to
assess a null hypothesis of CSR. The method was based on using Monte Carlo
tests to evaluate the deviation of the mean distance from that expected
under CSR. We will look at another possible approach in this problem, one that
theoretically would allow us to use a test based on normal theory. A question
on Homework 2 asked you to find the probability density function of \(H\), the
distance between an event and the nearest neighboring event. If you worked
this problem correctly you got}
\begin{equation*}
g(h)=2\lambda\pi h\exp\left(-\lambda\pi h^{2}\right)
\end{equation*}
{\it where \(\lambda > 0\). This is a Weibull distribution parametrized as}
\begin{equation*}
g(h)=\frac{\beta}{\theta^{\beta}}h^{\beta-1}
\exp\left(-\frac{h}{\theta}\right)^{\beta}
\end{equation*}
{\it and with parameters \(\beta = 2\) and
\(\theta = \left(\lambda\pi\right)^{-1/2}\). We will be working with a
homogeneous Poisson process with intensity \(\lambda = 30\).}

\begin{enumerate}

\item%a
{\it What are the mean and variance of \(\overline{H} = (1/30)\sum H_{i}\)
when \(\lambda = 30\), i.e. both the sample size and the intensity equal 30?}

Under CSR,
\begin{equation*}
E(H_{i}) = \theta \Gamma\left(1 + \frac{1}{\beta}\right)
= \frac{1}{\sqrt{30\pi}} \Gamma\left(\frac{3}{2}\right)
= \Sexpr{signif(gamma(1 + 1/2) / sqrt(30*pi), 6)}
\end{equation*}
and
\begin{align*}
Var(H_{i}) &= \theta^{2} \left(\Gamma\left(1 = \frac{2}{\beta}\right)
- \Gamma\left(1+\frac{1}{\beta}\right)^{2}\right) \\
&= \left(\frac{1}{\sqrt{30\pi}}\right)^{2} \left(\Gamma\left(2\right)
- \Gamma\left(\frac{3}{2}\right)^{2}\right) \\
&= \Sexpr{signif(1/(sqrt(30*pi)^2) * (gamma(1 + 2/2) - (gamma(1 + 1/2))^2), 6)}
.\end{align*}
If the events are independent (as they are under CSR) then the \(H_{i}\) are
also independent. So
\begin{equation*}
E\left(\overline{H}\right) = \frac{1}{n} \sum_{i}E(H_{i})
= \frac{n \times \Sexpr{signif(gamma(1 + 1/2) / sqrt(30*pi), 6)}}{n}
= \Sexpr{signif(gamma(1 + 1/2) / sqrt(30*pi), 6)}
\end{equation*}
and
\begin{equation*}
Var\left(\overline{H}\right) = \frac{1}{n^{2}} \sum_{i} Var(H_{i})
= \frac{n \times \Sexpr{signif(1/(sqrt(30*pi)^2) * (gamma(1 + 2/2) -
                               (gamma(1 + 1/2))^2), 6)}}{n^{2}}
= \frac{\Sexpr{signif(1/(sqrt(30*pi)^2) * (gamma(1 + 2/2) -
                      (gamma(1 + 1/2))^2), 4)}}{30}
= \Sexpr{signif(1/(30 * sqrt(30*pi)^2) * (gamma(1 + 2/2) -
                (gamma(1 + 1/2))^2), 6)}.
\end{equation*}

\item%b
{\it What is the approximate sampling distribution of}
\begin{equation*}
\frac{\overline{H}-E\left[\overline{H}\right]}
{\sqrt{Var\left[\overline{H}\right]}}
\end{equation*}
{\it under CSR and how do you know this?}

For a large sample this is approximately standard normal because
\(\overline{H}\) is a sample mean so the Central Limit Theorem applies.

\item%c
{\it Simulate 1000 realizations of complete spatial randomness in the unit
square with 30 events in each realization. For each realization}

\begin{enumerate}

\item%i
{\it Calculate the distance between each event and its nearest-neighboring
event (\(H_{i}\) for the \(i\)th event in the realization)}

\item%ii
{\it Calculate and store the mean distance.}

\item%iii
{\it Calculate and store the values of}
\begin{equation*}
Z=\frac{\overline{H}-E\left[\overline{H}\right]}
{\sqrt{Var\left[\overline{H}\right]}}
\end{equation*}
{\it using the mean and variance from part (a) above.}

\end{enumerate}

<<prob2c, message = FALSE>>=
library(spatstat, quietly = TRUE)

# Simulation parameters.
lambda <- 30
n_sim <- 1000

# Simulate.
sims <- runifpoint(n = lambda, nsim = n_sim)
results <- data.frame(t(sapply(sims, function(x){

  # Nearest neighbor distances.
  H <- nndist(x)

  # Mean nearest neighbor distance.
  Hbar <- mean(H)

  # Z-score.
  Z <- (Hbar - 0.0912871) / sqrt(0.0000759)

  return(c(Hbar = Hbar, Z = Z))
})))
@

\item%d
{\it Compute the mean and standard deviation of the 1000 simulated
\(\overline{H}\) values and compare them to what would be expected under CSR.
Are they higher or lower than expected? What could explain this result?}

<<prob2d>>=
mean(results$Hbar)
sd(results$Hbar)
var(results$Hbar)
@

The mean and variance of the simulated \(\overline{H}\) are both larger than
what would be expected under CSR. This could be due to edge effects where
points near the boundary have larger nearest neighbor distances than points in
the interior of the region, resulting in a skewed distribution of the
\(H_{i}\) values and therefore inflating the mean and variance of
\(\overline{H}\).

\item%e
{\it Produce a qqplot of the \(z\) scores. Comment.}

<<prob2e, fig.height = 5>>=
qqnorm(results$Z)
@

The points form a straight line so there is little doubt that the distribution
of simulated \(\overline{H}\) values is normal. However, the sample quantiles
range from about \(-3\) to about 4 while the theoretical quantiles go from
\(-3\) to 3, so it appears the distribution is not exactly \emph{standard}
normal.

\item%f
{\it Use the following formulas for the expected value and variance of \(H\):}
\begin{equation*}
E\left[\overline{H}\right]=0.5\sqrt{A/n}+0.051P/n+0.041P/n^{3/2}
\end{equation*}
\begin{equation*}
Var\left[\overline{H}\right]=0.0703A/n^{2}+0.037P\sqrt{A/n^{5}}
\end{equation*}
{\it where \(A\) is the area and \(P\) is the perimeter of the spatial domain
(the unit square). Compare the mean and standard deviation from these formulas
to those you computed from the simulations above. Does this modification seem
to help?}

\begin{equation*}
E\left(\overline{H}\right)
= 0.5\sqrt{\frac{1}{30}}+0.051\frac{4}{30}+0.041\frac{4}{30^{3/2}}
= \Sexpr{signif(0.5 / sqrt(30) + 0.051 * 4 / 30 + 0.041 * 4 / (30^1.5), 6)}
\end{equation*}
\begin{equation*}
Var\left(\overline{H}\right)=0.0703\frac{1}{30^{2}}
+0.037 \times 4\sqrt{\frac{1}{30^{5}}}
= \Sexpr{signif(0.0703 / (30^2) + 0.037 * 4 / sqrt(30^5), 6)}
\end{equation*}
Yes, the mean and variance computed using these formulas are closer to the
simulated mean and variance.

\item%g
{\it The above procedure is called the Clark-Evans test. Use it to test the
null hypothesis of CSR for the cells and redwood data sets. Interpret the
results of each test. Also, compute approximate large sample 95\% confidence
intervals for the mean distance and interpret.}

\textbf{Cells}

<<prob2g1>>=
data(cells)
cells # Checking that the window is the unit square.
Hbar <- mean(nndist(cells))
Hbar
@

The observed point pattern has \Sexpr{cells[['n']]} trees with a sample mean
nearest neighbor distance of \(\overline{H} = \Sexpr{signif(Hbar, 6)}\). Under
the null hypothesis of complete spatial randomness, we expect
\(\overline{H}\) to follow a normal distribution with mean
\begin{equation*}
E\left(\overline{H}\right)
= 0.5\sqrt{\frac{1}{42}}+0.051\frac{4}{42}+0.041\frac{4}{42^{3/2}}
= \Sexpr{signif(0.5 / sqrt(42) + 0.051 * 4 / 42 + 0.041 * 4 / (42^1.5), 6)}
\end{equation*}
and variance
\begin{equation*}
Var\left(\overline{H}\right)=0.0703\frac{1}{42^{2}}
+0.037 \times 4\sqrt{\frac{1}{42^{5}}}
= \Sexpr{signif(0.0703 / (42^2) + 0.037 * 4 / sqrt(42^5), 6)}.
\end{equation*}
<<prob2g2>>=
2 * pnorm(0.128973, 0.0826113, sqrt(0.0000528), lower.tail = FALSE)
@
This results in a p-value \(<\) 0.0001, very strong evidence that the
locations of the cells are not completely spatially random. A 95\% confidence
interval for the mean nearest neighbor distance is
\Sexpr{paste(signif(qnorm(c(0.025, 0.975), 0.128974, sqrt(0.0000528)), 3),
             collapse = ' to ')}; this interval suggests that cells are farther
from their nearest neighbors than expected under CSR, implying that the cells
have a regular pattern.

\textbf{Redwood}

<<prob2g3>>=
data(redwood)
redwood
Hbar <- mean(nndist(redwood))
Hbar
@

The observed point pattern has \Sexpr{redwood[['n']]} trees with a sample mean
nearest neighbor distance of \(\overline{H} = \Sexpr{signif(Hbar, 6)}\). Under
the null hypothesis of complete spatial randomness, we expect
\(\overline{H}\) to follow a normal distribution with mean
\begin{equation*}
E\left(\overline{H}\right)
= 0.5\sqrt{\frac{1}{62}}+0.051\frac{4}{62}+0.041\frac{4}{62^{3/2}}
= \Sexpr{signif(0.5 / sqrt(62) + 0.051 * 4 / 62 + 0.041 * 4 / (62^1.5), 6)}
\end{equation*}
and variance
\begin{equation*}
Var\left(\overline{H}\right)=0.0703\frac{1}{62^{2}}
+0.037 \times 4\sqrt{\frac{1}{62^{5}}}
= \Sexpr{signif(0.0703 / (62^2) + 0.037 * 4 / sqrt(62^5), 6)}.
\end{equation*}
<<prob2g4>>=
2 * pnorm(0.0392843, 0.0671263, sqrt(0.0000232))
@
This results in a p-value \(<\) 0.0001, very strong evidence that the
locations of the trees are not completely spatially random. A 95\% confidence
interval for the mean nearest neighbor distance is
\Sexpr{paste(signif(qnorm(c(0.025, 0.975), 0.0392843, sqrt(0.0000232)), 3),
             collapse = ' to ')}; this interval suggests that trees are closer
to their nearest neighbors than expected under CSR, implying that the trees
are clustered.

\end{enumerate}

\item%3
{\it I am sending you a copy of a paper by Peter Diggle on the use of \(K\)
and cross \(K\) functions in the analysis of spatial point patterns. The data
he is referring to are in the amacrine data set in the spatstat library in R.
Read the paper and reproduce the analysis. The data are in \texttt{spatstat}
(use the command \texttt{data(amacrine)}. You do not have to carry out the
significance tests he refers to but I would like for you to take the same
approach I took on the analysis of the \texttt{betacells} data set we
discussed in class. Write up a summary of your analysis. Pay attention to the
distinction between the independence and random labeling hypotheses.}

<<prob3_1>>=
data(amacrine)
par(mar = c(0, 0, 0.5, 0))
plot(amacrine)
@

As Diggle did, I will assess the support for two hypotheses, H1: the on cells
and off cells result from independent processes, and H2: the on cells and off
cells result from a single process and are differentiated by random labeling.
I begin by using the \(L\) function for each cell type to examine the
second-order structure of the observed patterns.

<<prob3_2>>=
amacrine_split <- split(amacrine)
par(mfrow = c(1, 2), mar = c(4, 5, 2, 1))
plot(envelope(amacrine_split$off, fun = Lest, verbose = FALSE), .-r~r,
     legend = FALSE, ylab = expression(hat(L)[off](r)-r),
     main = 'L Function for Off Cells')
plot(envelope(amacrine_split$on, fun = Lest, verbose = FALSE), .-r~r,
     legend = FALSE, ylab = expression(hat(L)[on](r)-r),
     main = 'L Function for On Cells')
@

The \(L\) functions are generally similar, both taking values less than \(r\)
and falling outside the pointwise confidence band for \(r\) less than 0.15 or
so. This indicates regularity for both cell types. I will use the D function
to assess equality of the \(L\) functions more formally.

<<prob3_3, fig.width = 3.25>>=
library(splancs, quietly = TRUE)
amacrine_box <- bboxx(rbind(Window(amacrine)$xrange, Window(amacrine)$xrange))
r <- Lest(amacrine)$r
k_off <- khat(as.points(amacrine_split$off), amacrine_box, r)
k_on <- khat(as.points(amacrine_split$on), amacrine_box, r)
D_hat <- k_off - k_on
Denv <- Kenv.label(as.points(amacrine_split$off), as.points(amacrine_split$on),
        amacrine_box, nsim = 99, r, quiet = TRUE)

# Use spatstat's plotting system for a consistent look.
Dfv <- fv(data.frame(r = r, Dhat = D_hat, lo = Denv$lower, hi = Denv$upper),
          valu = 'Dhat', yexp = expression(hat(D)(r)))
fvnames(Dfv, '.s') <- c('lo', 'hi')
par(mar = c(4, 5, 2, 1))
plot(Dfv, main = 'D Function', legend = FALSE)
segments(y0 = 0, x0 = min(r), x1 = max(r), col = 'red', lty = 2)
@

The \(D\) function stays near 0 for all \(r\), so we have no reason to believe
that the \(K\) (or \(L\)) functions differ between the cell types. Similarity
of the \(L\) functions is consistent with both independence and random
labeling, so I'll investigate further using the \(L\) function for all cells
of both types and the cross \(L\) function.

<<prob3_4>>=
par(mfrow = c(1, 2), mar = c(4, 5, 2, 1))
plot(envelope(amacrine, fun = Lest, verbose = FALSE), .-r~r,
     legend = FALSE, ylab = expression(hat(L)(r)-r),
     main = 'L Function Ignoring Marks')
plot(envelope(amacrine, fun = Lcross, verbose = FALSE), .-r~r,
     legend = FALSE, ylab = expression(hat(L)[list(off,on)](r)-r),
     main = 'Cross L Function')
@

\pagebreak
The \(L\) function ignoring cell type looks about the same as the \(L\)
functions for the separate cell types, but it does not deviate quite as far
from zero. As a single point pattern, the locations of the cells have a
regular pattern on a scale up to about 0.10 to 0.15 units. The cross \(L\)
function stays within the pointwise confidence bounds, providing no evidence of
dependence between the two point patterns. However, the cross \(L\) function is
clearly different from the \(L\) function of either cell type, which rules out
random labeling.

I like Diggle's plot showing all four \(K\) functions, so I recreated it with
the \(L\) functions. Random labeling requires all four of these functions to
be equal, which is not the case here. The cross \(L\) function is not very
different from \(r\) for all \(r\), so there is no reason to believe the
locations of cells of different types are not independent. The evidence is
consistent with H1, that the on cells and off cells are placed in regular
patterns by independent processes.

<<prob3_5, fig.height = 5>>=
par(mar = c(4, 5, 2, 1))
plot(Lest(amacrine), iso -r ~ r, ylab = expression(hat(L)(r) - r),
     main = 'All Four L Functions',
     ylim = c(-0.035, 0.025), legend = FALSE, lty = 1)
plot(Lest(amacrine_split$off), iso - r ~ r, lty = 2, add = TRUE)
plot(Lest(amacrine_split$on), iso - r ~ r, lty = 3, add = TRUE)
plot(Lcross(amacrine), iso - r ~ r, lty = 4, add = TRUE)
legend(0, 0.025, lty = 1:4, legend = expression(hat(L)(r), hat(L)[off](r),
                                                hat(L)[on](r), hat(L)[list(off,on)](r)))
@

\end{enumerate}

\end{document}

### 2017-02-10 hw04/hw04.rnw
Homework 4

\documentclass[11pt]{article}
\usepackage{fullpage,graphicx,float,amsmath,enumitem,fancyhdr}

\setlist{parsep=5.5pt}
\setlength{\parindent}{0pt}

\newcommand\assignment{Stat 534 Homework 4}
\newcommand\myname{Kenny Flagg}
\newcommand\duedate{February 10, 2017}

\pagestyle{fancy}
\lhead{\assignment}
\chead{\duedate}
\rhead{\myname}
\setlength{\headheight}{18pt}
\setlength{\headsep}{2pt}

\title{\assignment}
\author{\myname}
\date{\duedate}

\begin{document}
\maketitle

<<setup, echo = FALSE, message = FALSE, cache = FALSE>>=
library(knitr)
library(extrafont)
opts_chunk$set(echo = TRUE, comment = NA, message = FALSE,
               show.signif.stars = FALSE,
               fig.align = 'center', fig.width = 6.5, fig.height = 3,
               fig.pos = 'H', size = 'footnotesize', dev = 'pdf',
               dev.args = list(family = 'CM Roman', pointsize = 11))
knit_theme$set('print')

library(xtable)
options(xtable.floating = FALSE,
        xtable.sanitize.rownames.function = function(x) x,
        xtable.sanitize.colnames.function = function(x) x,
        width = 80, scipen = 2, show.signif.stars = FALSE)

set.seed(2753)
@

\begin{enumerate}

\item%1
{\it For \(\lambda = 30\) generate 9 realizations of CSR on the unit square.
For each realization, construct a kernel estimate of \(\lambda(s)\). How do
the estimated intensity functions compare to the constant intensity under CSR?
What precautions does this exercise suggest with regard to interpreting
estimates of intensity from a single realization (or data set)? The following
R code will simultaneously produce the data and plots.}
<<prob1, fig.height = 4>>=
library(spatstat, quiet = TRUE)
par(mfrow = c(3, 3), mar = c(0, 0, 1, 1), cex = 1)
for(i in 1:9) plot(density(rpoispp(30)), zlim = c(0, 140)) # zlim sets the color scale.
@
The estimated intensity functions do not look constant. Most of them have some
hotspots where the estimated intensity is at least 60, and most also have some
cold spots where there are few points so the estimated intensity is close to
zero. This suggests we shouldn't make much out of our estimated intensity
surface unless we have other information suggesting the process really is
heterogeneous, and even then we should be aware that there may be a lot of
uncertainty around the estimate.

\item%2
{\it In class we looked at a heterogeneous Poisson process on the unit square
with with intensity function}
\begin{equation*}
\lambda(x,y) = \exp(5x + 2y).
\end{equation*}

\begin{enumerate}

\item%a
{\it Simulate a realization of the process using the following R code. Plot
the results and comment.}
<<prob2a, fig.height = 2.5>>=
sim.dat <- rpoispp(function(x, y){exp(5 * x + 2 * y)})
par(mar = c(0, 0, 0.5, 0))
plot(sim.dat)
@
The plot looks about how I expected it to look. There are few events at the
bottom left, many events at the top right, a noticeable increase in event
intensity from left to right, and a subtle increase in intensity from bottom to
top (easily seen at the right edge).

\item%b
{\it Plot simulation envelopes for the \(K\) function (or some suitable
modification of it) and comment.}
<<prob2b>>=
par(mar = c(4, 4, 1, 4))
plot(envelope(sim.dat, Kest, verbose = FALSE), legend = FALSE)
@
The empirical \(K\) function stays far outside of the simulation envelope for
all distances except very small ones. There tend to be more events within a
given distance than expected under CSR, but this is not the same pattern we've
seen for clustered processes, where the \(K\) function is large for small
distances but similar to the CSR \(K\) function for larger distances. This
\(K\) function is close to the CSR \(K\) function at small distances, but moves
away from the value expected under CSR for larger distances because, for most
locations, any region other than a small neighborhood will include points with
higher intensity.

\item%c
{\it Fit a trend model to your data using ppm. Provide me with the parameter
estimates and associated standard errors.}
<<prob2c>>=
sim.fit <- ppm(sim.dat, ~ x + y, correction = 'isotropic')
sim.fit
@
The coefficient estimates, \(\widehat{\beta}_{0} =
\Sexpr{signif(coef(sim.fit)['(Intercept)'], 3)}\) (SE =
\Sexpr{signif(sqrt(vcov(sim.fit)['(Intercept)', '(Intercept)']), 3)}),
\(\widehat{\beta}_{x} = \Sexpr{signif(coef(sim.fit)['x'], 3)}\)
(SE = \Sexpr{signif(sqrt(vcov(sim.fit)['x', 'x']), 3)}) and
\(\widehat{\beta}_{y} = \Sexpr{signif(coef(sim.fit)['y'], 3)}\) (SE =
\Sexpr{signif(sqrt(vcov(sim.fit)['y', 'y']), 3)}), are pretty close to the
true values, with the true values being in the 95\% confidence intervals.

\item%d
{\it Check the fit using \texttt{quadrat.test}. Use
\texttt{method="MonteCarlo"} instead of the large sample chi-squared test.
Plot the results. Discuss.}
<<prob2d, fig.width = 5>>=
sim.qtest <- quadrat.test(sim.fit, method = 'MonteCarlo')
sim.qtest
par(mar = c(0, 0, 1, 0))
plot(sim.qtest)
points(sim.dat, pch = 19, col = '#00000040')
@
The p-value of \Sexpr{sim.qtest$p.value} is small enough to make you think, but
provides at best weak evidence that the model is a poor fit. The only standardized residuals with absolute value \(>\) 2 are in the leftmost column of quadrats
and one right in the center. I wouldn't worry about the left edge because the
expected counts are so close to zero that even 1 event would have a big
residual. In the center, the cluster of points with the standardized residual
of 3 isn't all that unusual because any points in a low-intensity region are
going to stand out. If I encountered these data in the field I would see if a
more complicated surface would fit better, but overall the model fits pretty
well. With 25 quadrats, some of them are bound to be in the tails of the
distribution.

\item%e
{\it Compare these results from this model with those to a model fit under an
assumption of CSR. Summarize the results. Provide me the model comparison
results (AIC comparisons are fine).}
<<prob2e>>=
sim.csr <- ppm(sim.dat, ~ 1, correction = 'isotropic')
sim.csr
AIC(sim.csr)
AIC(sim.fit)
@
The CSR model (intercept-only) assumes a constant intensity, estimated to be
\(\widehat{\lambda} = \exp(\Sexpr{signif(coef(sim.csr), 3)}) =
\Sexpr{signif(exp(coef(sim.csr)), 3)}\) with a 95\%
confidence interval of \(\exp(4.28)=\Sexpr{signif(exp(4.280881), 3)}\) to
\(\exp(4.70)=\Sexpr{signif(exp(4.696392), 3)}\). The heterogeneous model
fits much better, with an improvement in AIC of
\(\Sexpr{signif(AIC(sim.csr)-AIC(sim.fit), 5)}\).

\item%f
{\it Plot a nonparametric estimate of the intensity function. Compare the
fitted surface you got using \texttt{ppm} with the nonparametric (kernel
density estimate) surface in some suitable way.}

<<prob2f, fig.height = 2.5, results = 'hide'>>=
par(mfrow = c(1, 2), mar = c(0, 0, 1, 0))
plot(density(sim.dat))
diagnose.ppm(sim.fit, which = 'smooth')
@
The plot on the left shows a kernel density estimate of the intensity surface,
and the plot on the right shows the smoothed residuals, computed as the fitted
intensity surface minus a kernel density estimate. The residuals have bands
alternating positive and negative, indicating that the model missed some
of the patterns in the data.

\end{enumerate}

\item%3
{\it Recall the use of the \texttt{nncorr} statistic in the Finland Pines data
set. The distribution of heights (the marks) was of interest. We saw that the
nearest neighbor correlation between heights was −0.1839798. We questioned
whether or not this was unusual. Carry out a randomization test to assess
this. You can use the \texttt{rlabel} command to scramble the marks if you
want. Provide me with a histogram of the randomization distribution and a
p-value. Discuss \textbf{BRIEFLY} your results. Provide me with your R-code,
also.}
<<prob3, fig.width = 5.75>>=
finpines.ht <- finpines
marks(finpines.ht) <- marks(finpines)$height
obs_corr <- nncorr(finpines.ht)['correlation']
perm_corr <- c(obs_corr, replicate(999, nncorr(rlabel(finpines.ht))['correlation']))
hist(perm_corr, breaks = 100, freq = FALSE, xlab = 'Correlation',
     main = 'Permutation Distribution of Nearest Neighbor Correlation')
abline(v = obs_corr)
pval <- 2 * mean(perm_corr <= obs_corr)
pval
@
999 permutations yield an approximately symmetric permutation distributon, so
doubling the left-tail p-value results in a two-sided p-value of
\Sexpr{pval} giving no evidence that the heights or nearest neighbors are
correlated.

\item%4
{\it Let's derive a \(K\) function for something other than a CSR process. We
will assume a Neyman-Scott process with the following properties.}
\begin{enumerate}[label=\roman*.]
\item {\it The parent process is a homogeneous Poisson process with intensity
\(\lambda\).}
\item {\it The number of offspring produced by each parent (\(N\)) is
homogeneous Poisson with intensity \(\mu\).}
\item {\it The position of each offspring is determined by a bivariate normal
distribution with mean \((0, 0)\) (i.e. it is centered over the parent) and
variance-covariance matrix \(\sigma^{2}\mathbf{I}\). Note that this implies
that the \(x\) and \(y\) coordinates are determined independently of one
another with the same variance.}
\end{enumerate}
{\it Consider 2 offspring from the same parent located at \((X_{1},Y_{1})\)
and \((X_{2},Y_{2})\).}

\begin{enumerate}

\item%a
{\it What is the distribution of}
\begin{equation*}
W=\frac{(X_{1}-X_{2})^{2}}{2\sigma^{2}}+\frac{(Y_{1}-Y_{2})^{2}}{2\sigma^{2}}?
\end{equation*}
\(X_{1}\), \(X_{2}\), \(Y_{1}\), and \(Y_{2}\) are all independent
\(\mathrm{N}(0, \sigma^{2})\), so \(X_{1}-X_{2}\) and \(Y_{1}-Y_{2}\) are
independent \(\mathrm{N}(0, 2\sigma^{2})\). Then
\(\dfrac{(X_{1}-X_{2})^{2}}{2\sigma^{2}}\) and
\(\dfrac{(Y_{1}-Y_{2})^{2}}{2\sigma^{2}}\) are independent \(\chi^{2}_{1}\).
Therefore, \(W \sim \chi^{2}_{2}\).

\item%b
{\it Note that the Euclidean distance between the 2 points is}
\begin{equation*}
H=\left[(X_{1}-X_{2})^{2}+(Y_{1}-Y_{2})^{2})\right]^{1/2}
=\left(2\sigma^{2}W\right)^{1/2}.
\end{equation*}
{\it Derive the cdf of H.}
\begin{align*}
F(h) &= P(H \leq h) \\
&= P\left(\left(2\sigma^{2}W\right)^{1/2} \leq h\right) \\
&= P\left(W \leq \frac{h^{2}}{2\sigma^{2}}\right) \\
&= \int_{0}^{\frac{h^{2}}{2\sigma^{2}}} \frac{t^{0}e^{-\frac{t}{2}}}
{\Gamma\left(1\right)2^{1}} \mathrm{d}t \\
&= \int_{0}^{\frac{h{2}}{2\sigma^{2}}} \frac{1}{2}e^{-\frac{t}{2}}
\mathrm{d}t \\
&= \left[-e^{-\frac{t}{2}}\right]_{t=0}^{\frac{h^{2}}{2\sigma^{2}}} \\
&= 1 - \exp\left(-\frac{h^{2}}{4\sigma^{2}}\right)
\end{align*}

\item%c
{\it Recall that we were told that the \(K\) function for Neyman-Scott
processes with homogeneous Poisson parent processes and radially symmetric
\(f(h)\) is}
\begin{equation*}
K(h)=\pi h^{2} + \frac{E(N(N-1))}{\lambda E(N)^2}F(h).
\end{equation*}
{\it Use the results from above to find the K function for the described
process.}
\begin{align*}
K(h) &= \pi h^{2} + \frac{E(N(N-1))}{\lambda E(N)^{2}}
\left(1-\exp\left(-\frac{h^{2}}{4}\right)\right) \\
&= \pi h^{2} + \frac{E(N^{2})-E(N)}{\lambda E(N)^{2}}
\left(1-\exp\left(-\frac{h^{2}}{4}\right)\right) \\
&= \pi h^{2} + \frac{Var(N)+E(N)^{2}-E(N)}{\lambda E(N)^{2}}
\left(1-\exp\left(-\frac{h^{2}}{4}\right)\right) \\
&= \pi h^{2} + \frac{\mu+\mu^{2}-\mu}{\lambda\mu^{2}}
\left(1-\exp\left(-\frac{h^{2}}{4}\right)\right) \\
&= \pi h^{2} + \frac{1}{\lambda}
\left(1-\exp\left(-\frac{h^{2}}{4}\right)\right) \\
\end{align*}

\end{enumerate}

\end{enumerate}

\end{document}

### 2017-02-17 hw05/hw05.rnw
Hw5

\documentclass[11pt]{article}
\usepackage{fullpage,graphicx,float,amsmath,enumitem,fancyhdr}

\setlist{parsep=5.5pt}
\setlength{\parindent}{0pt}

\newcommand\assignment{Stat 534 Homework 5}
\newcommand\myname{Kenny Flagg}
\newcommand\duedate{February 17, 2017}

\pagestyle{fancy}
\lhead{\assignment}
\chead{\duedate}
\rhead{\myname}
\setlength{\headheight}{18pt}
\setlength{\headsep}{2pt}

\title{\assignment}
\author{\myname}
\date{\duedate}

\begin{document}
\maketitle

<<setup, echo = FALSE, message = FALSE, cache = FALSE>>=
library(knitr)
library(extrafont)
opts_chunk$set(echo = TRUE, comment = NA, message = FALSE,
               show.signif.stars = FALSE,
               fig.align = 'center', fig.width = 6.5, fig.height = 4,
               fig.pos = 'H', size = 'footnotesize', dev = 'pdf',
               dev.args = list(family = 'CM Roman', pointsize = 11))
knit_theme$set('print')

library(xtable)
options(xtable.floating = FALSE,
        xtable.sanitize.rownames.function = function(x) x,
        xtable.sanitize.colnames.function = function(x) x,
        xtable.include.rownames = FALSE,
        width = 80, scipen = 2, show.signif.stars = FALSE)

set.seed(123)#set.seed(25277)
@

\begin{enumerate}

\item%1
{\it Let \(\gamma(\mathbf{s}_{i}, s_{j}) = \gamma(\mathbf{h}_{ij})\) be a
semivariogram for a second-order stationary spatial process.}

\begin{enumerate}

\item%a
{\it Show that \(\gamma(\mathbf{h}_{ij})=C(\mathbf{0})-C(\mathbf{h}_{ij})\).}
\begin{align*}
\gamma(\mathbf{h}_{ij})
&= \frac{1}{2} Var\left[Z(\mathbf{s}_{i}) - Z(\mathbf{s}_{j})\right] \\
&= \frac{1}{2} \left[Var(Z(\mathbf{s}_{i})) + Var(Z(\mathbf{s}_{j}))
- 2 Cov(Z(\mathbf{s}_{i}), Z(\mathbf{s}_{j}))\right] \\
&= \frac{1}{2} Var(Z(\mathbf{s}_{i})) + \frac{1}{2} Var(Z(\mathbf{s}_{j}))
- Cov(Z(\mathbf{s}_{i}), Z(\mathbf{s}_{j})) \\
&= \frac{1}{2} C(\mathbf{0}) + \frac{1}{2} C(\mathbf{0})
- C(\mathbf{s}_{i} - \mathbf{s}_{j}) \\
&= C(\mathbf{0}) - C(\mathbf{h}_{ij})
\end{align*}
\item%b
{\it Show that \(\displaystyle
\sum_{i=1}^{n}\sum_{j=1}^{n} a_{i}a_{j}\gamma(\mathbf{h}_{ij}) \leq 0\)
for any sites \(\mathbf{s}_{i}\), \(i = 1,\dots,n\) and for constants
\(a_{i}\), \(i = 1,\dots,n\) with \(\sum_{i=1}^{n} a_{i} = 0\).}
\begin{align*}
\sum_{i=1}^{n}\sum_{j=1}^{n} a_{i}a_{j}\gamma(\mathbf{h}_{ij})
&= \sum_{i=1}^{n}\sum_{j=1}^{n}a_{i}a_{j}[C(\mathbf{0})-C(\mathbf{h}_{ij})] \\
&= C(\mathbf{0})\sum_{i=1}^{n} a_{i} \sum_{j=1}^{n} a_{j}
- \sum_{i=1}^{n}\sum_{j=1}^{n}a_{i}a_{j}
 Cov(Z(\mathbf{s}_{i}), Z(\mathbf{s}_{j})) \\
&= 0 - \sum_{i=1}^{n}\sum_{j=1}^{n}a_{i}a_{j}
\left[E(Z(\mathbf{s}_{i})Z(\mathbf{s}_{j}))
- E(Z(\mathbf{s}_{i}))E(Z(\mathbf{s}_{j}))\right] \\
&= -\sum_{i=1}^{n}\sum_{j=1}^{n}a_{i}a_{j}
E(Z(\mathbf{s}_{i})Z(\mathbf{s}_{j}))
+ \sum_{i=1}^{n}\sum_{j=1}^{n}a_{i}a_{j}
E(Z(\mathbf{s}_{i}))E(Z(\mathbf{s}_{j})) \\
&= -E\left(\sum_{i=1}^{n}a_{i}Z(\mathbf{s}_{i})
\sum_{j=1}^{n}a_{j}Z(\mathbf{s}_{j})\right)
+ \sum_{i=1}^{n}\sum_{j=1}^{n}a_{i}a_{j} \mu^{2} \\
&= -E\left(\sum_{i=1}^{n}a_{i}Z(\mathbf{s}_{i})\right)^{2} + 0 \\
&\leq 0
\end{align*}

\end{enumerate}

\item%2
{\it Matheron's semivariogram estimator is}
\begin{equation*}
\widehat{\gamma}(\mathbf{h}) = \frac{1}{2|N(\mathbf{h})|}
\sum_{N(\mathbf{h})} \left\{Z(\mathbf{s}_{i}) - Z(\mathbf{s}_{j})\right\}^{2}
\end{equation*}
{\it where \(N(\mathbf{h}) = \{(\mathbf{s}_{i}, \mathbf{s}_{j}):
\mathbf{h} = \mathbf{s}_{i} - \mathbf{s}_{j}\}\) and \(|N(\mathbf{h})|\) is
the number of pairs in the set \(N(\mathbf{h})\).}

\begin{enumerate}

\item%a
{\it Let \(Z(\mathbf{s}) = \mu + e(\mathbf{s})\) where \(E[e(\mathbf{s})]=0\)
with \(\gamma_{Z}(\mathbf{h}) = \gamma_{e}(\mathbf{h})\) (adding a constant to
the random error terms does not change the variance/covariance properties of
the process). Show that \(\widehat{\gamma}(\mathbf{h})\) is unbiased for
\(\gamma_{Z}(\mathbf{h})\). That is, show \(E[\widehat{\gamma}(\mathbf{h})]
= \gamma_{Z}(\mathbf{h})\). Hint: Show that under an assumption of a constant
mean}
\begin{equation*}
\gamma_{Z}(\mathbf{h})
= \frac{1}{2}E\left[Z(\mathbf{s}_{i}) - Z(\mathbf{s}_{j})\right]^{2}
\end{equation*}

First,
\begin{align*}
\gamma_{Z}(\mathbf{h})
&= \frac{1}{2} Var\left[Z(\mathbf{s}_{i}) - Z(\mathbf{s}_{j})\right] \\
&= \frac{1}{2} \left(E\left[Z(\mathbf{s}_{i}) - Z(\mathbf{s}_{j})\right]^{2}
- \left(E[Z(\mathbf{s}_{i}) - Z(\mathbf{s}_{j})]\right)^{2}\right) \\
&= \frac{1}{2} \left(E\left[Z(\mathbf{s}_{i}) - Z(\mathbf{s}_{j})\right]^{2}
- \left(E[Z(\mathbf{s}_{i})] - E[Z(\mathbf{s}_{j})]\right)^{2}\right) \\
&= \frac{1}{2} \left(E\left[Z(\mathbf{s}_{i}) - Z(\mathbf{s}_{j})\right]^{2}
- \left(\mu - \mu\right)^{2}\right) \\
&= \frac{1}{2} E\left[Z(\mathbf{s}_{i}) - Z(\mathbf{s}_{j})\right]^{2}.
\end{align*}
Then,
\begin{align*}
E\left[\widehat{\gamma}(\mathbf{h})\right]
&= E\left[\frac{1}{2|N(\mathbf{h})|} \sum_{N(\mathbf{h})}
\left\{Z(\mathbf{s}_{i}) - Z(\mathbf{s}_{j})\right\}^{2}\right] \\
&= \frac{1}{|N(\mathbf{h})|} \sum_{N(\mathbf{h})}\frac{1}{2}E\left[
Z(\mathbf{s}_{i}) - Z(\mathbf{s}_{j})\right]^{2} \\
&= \frac{1}{|N(\mathbf{h})|} \sum_{N(\mathbf{h})}\gamma_{Z}(\mathbf{h}_{ij}) \\
&= \frac{1}{|N(\mathbf{h})|} |N(\mathbf{h})|\gamma_{Z}(\mathbf{h}) \\
&= \gamma_{Z}(\mathbf{h})
\end{align*}
so \(\widehat{\gamma}(\mathbf{h})\) is unbiased for \(\gamma_{Z}(\mathbf{h})\).

\pagebreak
\item%b
{\it We pointed out in class that Matheron's Estimator is biased in the
presence of trend. Let \(Z(\mathbf{s}) = \mu(\mathbf{s}) + e(\mathbf{s})\)
with \(E[e(\mathbf{s})] = 0\) and \(\gamma_{Z}(\mathbf{h})
= \gamma_{e}(\mathbf{h})\). Show}
\begin{equation*}
E\left[\widehat{\gamma}(\mathbf{h})\right]
= \gamma_{e}(\mathbf{h}) + \frac{1}{2|N(\mathbf{h})|}
\sum_{N(\mathbf{h})}\left[\mu(\mathbf{s}_{i})-\mu(\mathbf{s}_{j})\right]^{2}.
\end{equation*}

First note that
\begin{equation*}
\gamma_{e}(\mathbf{h})
= \frac{1}{2}E\left[e(\mathbf{s}_{i})-e(\mathbf{s}_{j})\right]^{2}
\end{equation*}
because the mean of \(e(\mathbf{s})\) is constant. Now,
\begin{align*}
E\left[\widehat{\gamma}(\mathbf{h})\right]
&= E\left[\frac{1}{2|N(\mathbf{h})|} \sum_{N(\mathbf{h})}
\left\{Z(\mathbf{s}_{i}) - Z(\mathbf{s}_{j})\right\}^{2}\right] \\
&= E\left[\frac{1}{2|N(\mathbf{h})|} \sum_{N(\mathbf{h})}
\left\{\mu(\mathbf{s}_{i})+e(\mathbf{s}_{i})
- \mu(\mathbf{s}_{j})-e(\mathbf{s}_{j})\right\}^{2}\right] \\
&= E\left[\frac{1}{2|N(\mathbf{h})|} \sum_{N(\mathbf{h})}
\left\{[e(\mathbf{s}_{i}) - e(\mathbf{s}_{j})]^{2}
+ [\mu(\mathbf{s}_{i}) - \mu(\mathbf{s}_{j})]^{2}
- [\mu(\mathbf{s}_{i}) - \mu(\mathbf{s}_{j}))
(e(\mathbf{s}_{i}) - e(\mathbf{s}_{j})]\right\}\right] \\
&= \frac{1}{2|N(\mathbf{h})|} \sum_{N(\mathbf{h})}\left\{
E[e(\mathbf{s}_{i}) - e(\mathbf{s}_{j})]^{2}
+ E[\mu(\mathbf{s}_{i}) - \mu(\mathbf{s}_{j})]^{2}
- E[\mu(\mathbf{s}_{i}) - \mu(\mathbf{s}_{j}))
(e(\mathbf{s}_{i}) - e(\mathbf{s}_{j})]\right\} \\
&= \frac{1}{2|N(\mathbf{h})|} \sum_{N(\mathbf{h})}\left\{
E[e(\mathbf{s}_{i}) - e(\mathbf{s}_{j})]^{2}
+ [\mu(\mathbf{s}_{i}) - \mu(\mathbf{s}_{j})]^{2}
- [\mu(\mathbf{s}_{i}) - \mu(\mathbf{s}_{j})]
E[e(\mathbf{s}_{i}) - e(\mathbf{s}_{j})]\right\} \\
&= \frac{1}{2|N(\mathbf{h})|} \sum_{N(\mathbf{h})}\left\{
E[e(\mathbf{s}_{i}) - e(\mathbf{s}_{j})]^{2}
+ [\mu(\mathbf{s}_{i}) - \mu(\mathbf{s}_{j})]^{2}
- [\mu(\mathbf{s}_{i}) - \mu(\mathbf{s}_{j})]
(0)\right\} \\
&= \frac{1}{2|N(\mathbf{h})|} \sum_{N(\mathbf{h})}
E[e(\mathbf{s}_{i}) - e(\mathbf{s}_{j})]^{2}
+ \frac{1}{2|N(\mathbf{h})|} \sum_{N(\mathbf{h})}
[\mu(\mathbf{s}_{i}) - \mu(\mathbf{s}_{j})]^{2} \\
&= \gamma_{e}(\mathbf{h})
+ \frac{1}{2|N(\mathbf{h})|} \sum_{N(\mathbf{h})}
[\mu(\mathbf{s}_{i}) - \mu(\mathbf{s}_{j})]^{2}
\end{align*}

\pagebreak
\item%c
{\it Consider the (very) simple model \(Z_{i} = 10 + e_{i}\) where the
\(e_{i}\) are independent normally distributed error terms with variance
\(\sigma^{2} = 81\). We have a pure nugget effect model
\(\gamma_{Z}(\mathbf{h}) = \gamma_{e}(\mathbf{h}) = 81\). Simulate 100
observations of \(Z_{i}\) and calculate the empirical semivariogram assuming
the observations are on a one-dimensional transect.}
<<prob2c>>=
library(geoR)
set.seed(25277)
Zdat <- 10 + rnorm(100, 0, 9)
i <- 1:100
xycoord <- cbind(rep(1, 100), i)
Zvgram <- variog(coords = xycoord, data = Zdat)
par(mar = c(4, 4, 1, 1))
plot(Zvgram)
abline(h = 81)
@
{\it Compare what you see in the plot to the true \(\gamma_{Z}(\mathbf{h})\).
Is the result consistent with part (a) above? Why or why not?}

The empirical semivariogram is nearly horizontal and generally stays near the
true value of \(\gamma(\mathbf{h}) = 81\), which is consistent with the
theoretical unbiasedness shown in part (a). The estimates at different lags
tend to be higher than the true semivariogram, but that is not surprising
because they use the same observed values and are correlated. This is only
one observation of a spatial process; Matheron's estimator is unbiased over
all realizations of the process.

\pagebreak
\item%d
{\it Redo the above calculations based on the (still) simple model
\(Z_{i} = 10 + 10i + e_{i}\), i.e. there is now a linear trend and the process
is no longer stationary.}
<<prob2d>>=
Zdat2 <- 10 + 10 * i + rnorm(100, 0, 9)
Zvgram2 <- variog(coords = xycoord, data = Zdat2)
par(mar = c(4, 4, 1, 1))
plot(Zvgram2)
@
{\it Compare the empirical semivariogram \texttt{Zvgram2} to \text{Zvgram}.
Are the results consistent with part (b) above? Justify your answer.}

The data in (c) came from a stationary process so that semivariogram is roughly
constant. but this empirical semivariogram has a parabolic shape because of
son-stationarity. In this situation, the trend in the mean is linear, with
\(\mu(s_{i})-\mu(s_{j}) = 10(s_{i}-s_{j})\), so the parabolic pattern is
consistent with part (b) which shows that the bias in Matheron's estimator is
quadratic in the difference in means.

\pagebreak
\item%e
{\it Fit a linear model to the data in (d), extract the residuals, and compute
the empirical semivariogram for the residuals. Note that what you are doing is
removing the trend.}
<<prob2e>>=
e.resid <- residuals(lm(Zdat2 ~ i))
evgram <- variog(coords = xycoord, data = e.resid)
par(mar = c(4, 4, 1, 1))
plot(evgram)
abline(h = 81)
@
{\it Compare the 3 empirical semivariograms.}

The residuals have a constant mean of zero, so the empirical semivariogram of
the residuals does not have a pattern of increasing bias like the one seen in
part (d). Like in part (c), the empirical semivariogram is an overestimate but
that is probably because of sampling variability (this is another sample of
size one, after all).

\end{enumerate}

\pagebreak
\item%3
{\it Attached is a data set containing the carbon nitrogen values used in the
carbon/nitrogen data set. The first 2 columns contain the coordinates, total
nitrogen is in the third column, total carbon is in the 4th column and the
ratio is in the last column. We will work with the total carbon data. Use
\texttt{geoR} for the analysis. It will be easiest if you convert the data
into a \texttt{geodata} object as follows.}
<<prob3>>=
CN.dat <- read.table('CN.dat', header = TRUE)
TC.geodata <- as.geodata(CN.dat, coords.col = 1:2, data.col = 4)
@

\begin{enumerate}

\item%a
{\it Calculate the empirical semivariogram. Give initial eyeball estimates of
the nugget effect, sill, and (effective) range.}

<<prob3a>>=
TC.vgram <- variog(TC.geodata)
par(mar = c(4, 4, 1, 1))
plot(TC.vgram)
@

The sill is around 0.015, the effective range looks to be about 150, and the
nugget is probably somewhere near 0.005.

\pagebreak
\item%b
{\it Fit an exponential semivariogram to the carbon data using \(OLS\),
\(WLS\), \(MLE\), and \(REML\) methods. Specify a nugget effect in each case,
i.e. you do not need to consider models without a nugget. Plot the fitted
functions and comment on which one you like best.}

<<prob3b, cache = TRUE, results = 'hide'>>=
tc.e.ols <- variofit(TC.vgram, ini.cov.pars = c(0.015, 150/3), nugget = 0.005,
                     fix.nugget = FALSE, cov.model = 'exponential',
                     weights = 'equal')
tc.e.wls <- variofit(TC.vgram, ini.cov.pars = c(0.015, 150/3), nugget = 0.005,
                     fix.nugget = FALSE, cov.model = 'exponential',
                     weights = 'cressie')
tc.e.ml <- likfit(TC.geodata, ini.cov.pars = c(0.015, 150/3), nugget = 0.005,
                  fix.nugget = FALSE, cov.model = 'exponential',
                  lik.method = 'ML')
tc.e.reml <- likfit(TC.geodata, ini.cov.pars = c(0.015, 150/3), nugget = 0.005,
                    fix.nugget = FALSE, cov.model = 'exponential',
                    lik.method = 'RML')

par(mar = c(4, 4, 1, 1))
plot(TC.vgram)
lines(tc.e.ols, lty = 1)
lines(tc.e.wls, lty = 2)
lines(tc.e.ml, lty = 3)
lines(tc.e.reml, lty = 4)
legend('bottomright', lty = 1:4, legend = c('OLS', 'WLS', 'ML', 'REML'))
@

All of the estimates are similar, but the OLS and WLS estimates look best
because they come closest to \(\widehat{\gamma}\) at the smallest lags. I
would use WLS just because I like the theoretical idea of downweighting the
large lags even though the result is barely distinguishable from the OLS
result here.

\item%c
{\it Redo part (b) by fitting a spherical model.}

<<prob3c, cache = TRUE, results = 'hide'>>=
tc.s.ols <- variofit(TC.vgram, ini.cov.pars = c(0.015, 150), nugget = 0.005,
                     fix.nugget = FALSE, cov.model = 'spherical',
                     weights = 'equal')
tc.s.wls <- variofit(TC.vgram, ini.cov.pars = c(0.015, 150), nugget = 0.005,
                     fix.nugget = FALSE, cov.model = 'spherical',
                     weights = 'cressie')
tc.s.ml <- likfit(TC.geodata, ini.cov.pars = c(0.015, 150), nugget = 0.005,
                  fix.nugget = FALSE, cov.model = 'spherical',
                  lik.method = 'ML')
tc.s.reml <- likfit(TC.geodata, ini.cov.pars = c(0.015, 150), nugget = 0.005,
                    fix.nugget = FALSE, cov.model = 'spherical',
                    lik.method = 'RML')

par(mar = c(4, 4, 1, 1))
plot(TC.vgram, ylim = c(0, 0.022))
lines(tc.s.ols, lty = 1)
lines(tc.s.wls, lty = 2)
lines(tc.s.ml, lty = 3)
lines(tc.s.reml, lty = 4)
legend('bottomright', lty = 1:4, legend = c('OLS', 'WLS', 'ML', 'REML'))
@

Again, I would choose WLS. This time, ML and REML both overestimate the sill,
but WLS and OLS match \(\widehat{\gamma}\) pretty well.

\pagebreak
\item%d
{\it Summarize your results in table format. Compare the results and discuss.}

<<prob3d, eval = FALSE>>=
compare <- do.call(rbind,
                   lapply(list(tc.e.ols, tc.e.wls, tc.e.ml, tc.e.reml,
                               tc.s.ols, tc.s.wls, tc.s.ml, tc.s.reml),
                          function(x){
                            return(data.frame(
                                     Model = x$cov.model,
                                     Method = x$method,
                                     Nugget = x$nugget,
                                     Sill = x$cov.pars[1],
                                     Range = x$cov.pars[2] *
                                       ifelse(x$cov.model == 'exponential', 3, 1)))
                          }))
xtable(compare, digits = 5)
@
\begin{center}
<<prob3d, echo = FALSE, results = 'asis'>>=
@
\end{center}
I multiplied the exponential range parameter by three so that the table
actually shows the effective range. I tried a few different initial range
values, and as long as the initial range was somewhere in the flat section of
the semivariogram, the OLS and WLS range estimate was close to the initial
value (for both models). For a given model, all four estimation methods gave
about the same nugget estimate, though the nugget estimates for the spherical
model were almost 2.5 times as big as the nugget estimates for the exponential
model. OLS and WLS gave similar sill values, and ML and REML also found
similar sill estimates; for the exponential model all four methods found
pretty similar sills, but for the spherical model OLS/WLS and ML/REML yielded
noticeably different sill values. The conclusion I can make from all this is
that fitting a semivariogram model involves many choices that affect the end
result, so you need to check the fitted model against the empirical
semivariogram to see if it looks reasonable. For this dataset, I would use the
exponential model fit by WLS.

\end{enumerate}

\end{enumerate}

\end{document}

### 2017-03-08 hw06/hw06.rnw
Got a little behind.

\documentclass[11pt]{article}
\usepackage{fullpage,graphicx,float,amsmath,enumitem,fancyhdr}

\setlist{parsep=5.5pt}
\setlength{\parindent}{0pt}

\newcommand\assignment{Stat 534 Homework 6}
\newcommand\myname{Kenny Flagg}
\newcommand\duedate{February 24, 2017}

\pagestyle{fancy}
\lhead{\assignment}
\chead{\duedate}
\rhead{\myname}
\setlength{\headheight}{18pt}
\setlength{\headsep}{2pt}

\title{\assignment}
\author{\myname}
\date{\duedate}

\begin{document}
\maketitle

<<setup, echo = FALSE, message = FALSE, cache = FALSE>>=
library(knitr)
library(extrafont)
opts_chunk$set(echo = TRUE, comment = NA, message = FALSE,
               show.signif.stars = FALSE,
               fig.align = 'center', fig.width = 6.5, fig.height = 4,
               fig.pos = 'H', size = 'footnotesize', dev = 'pdf',
               dev.args = list(family = 'CM Roman', pointsize = 11))
knit_theme$set('print')

library(geoR)
library(xtable)
options(xtable.floating = FALSE,
        xtable.sanitize.rownames.function = function(x) x,
        xtable.sanitize.colnames.function = function(x) x,
        xtable.include.rownames = FALSE,
        width = 80, scipen = 2, show.signif.stars = FALSE)

set.seed(789623)
@

\begin{enumerate}

\item%1
{\it Refer to the 7 point ordinary kriging example we discussed in class.
Repeat the example with the following changes. We move point
\(\mathbf{s}_{3}\) to (22, 22) and point \(\mathbf{s}_{5}\) to (15, 20).
Compare and contrast the results to the example being sure to address the
following. You need to be careful with the nugget effect and range parameters.}

{\it Recall that Model D is a pure nugget effect model. You do not need the
above function to fit that one. A table showing predictions, kriging
variances, and weights will help.}

<<prob1_1, fig.width = 3.5>>=
prob1 <- data.frame(x = c(  5, 20, 22,  8, 15, 35, 38),
                    y = c( 20,  2, 22, 39, 20, 20, 10),
                    z = c(100, 70, 60, 90, 50, 80, 40))
plot(y ~ x, data = prob1, pch = as.character(1:7),
     main = expression('Locations with Relocated '*bold(s)[3]*' and '*bold(s)[5]))
points(20, 20, pch = 19)
@

<<prob1_2, eval = FALSE>>=
# Eighth row/column are for s_0.
dists <- as.matrix(dist(rbind(prob1[,1:2], c(20, 20))))

# A function to automate all this.
prob1_function <- function(model, eff_range, sill, nugget){
  if(model == 'exponential')
    eff_range <- eff_range / 3 # Adjust exponential range by 3.
  else if(model == 'gaussian')
    eff_range <- eff_range / sqrt(3) # Adjust gaussian range by sqrt(3).
  if(model == 'nugget')
    Sigma <- diag(rep(nugget, 8))
  else
    Sigma <- cov.spatial(dists, cov.model = model,
                         cov.pars = c(sill - nugget, eff_range)) +
             diag(rep(nugget, 8))
  sig <- Sigma[8, 1:7]
  SigmaStar <- Sigma
  SigmaStar[8, 1:7] <- 1
  SigmaStar[1:7, 8] <- 1
  SigmaStar[8, 8] <- 0
  sigStar <- c(sig, 1)
  lambdaStar <- solve(SigmaStar) %*% sigStar
  p_ok <- t(lambdaStar[-8]) %*% prob1$z
  if(model == 'nugget')
    sig2_ok <- nugget - t(lambdaStar) %*% sigStar
  else
    sig2_ok <- sill - t(lambdaStar) %*% sigStar
  return(c(p_ok, sig2_ok, lambdaStar[-8]))
}

prob1_table <- cbind(data.frame(Model = c(LETTERS[1:6], 'Distance')),
                     as.data.frame(rbind(
                       prob1_function('exponential', 20, 10, 0),
                       prob1_function('exponential', 10, 10, 0),
                       prob1_function('exponential', 20, 10, 5),
                       prob1_function('nugget', 0, 0, 10),
                       prob1_function('exponential', 20, 20, 0),
                       prob1_function('gaussian', 20, 10, 0),
                       c(NA, NA, dists[8, 1:7])
                     )))
colnames(prob1_table)[-1] <- c('\\(p_{ok}\\)', '\\(\\sigma^{2}_{ok}\\)',
                                paste0('\\(\\lambda_{', 1:7, '}\\)'))
print(xtable(prob1_table, align = 'ccrrrrrrrrr'), hline.after = c(0, 6))
@
<<prob1_2, echo = FALSE, results = 'asis'>>=
@

\begin{enumerate}

\item%a
{\it Do the same conclusions regarding the effects of the sill, range, and
nugget effect still hold?}

Yes, the sill still just rescales the variance, and decreasing the range or
increasing the nugget still reduces the variability of the weights.

It's interesting that the weights for \(\mathbf{s}_{3}\) and
\(\mathbf{s}_{5}\) are closer to the weights of the other points in Model B
(effective range of 10) than they are in Model A (effective range of 20). The
range doesn't act like a magic threshold that separates highly correlated
points from points with low correlation; it simply adjusts how quickly the
correlation decays over distance. Even though \(\mathbf{s}_{3}\) and
\(\mathbf{s}_{5}\) are the closest points to \(\mathbf{s}_{0}\), under Model B
they are still less correlated with \(\mathbf{s}_{0}\) than they are under
Model A, so under Model B they get weights that are closer to the weights of
the farthest points.

\item%b
{\it Which points are screening?}

\(\mathbf{s}_{1}\) is screened by \(\mathbf{s}_{5}\) and \(\mathbf{s}_{6}\)
is screened by \(\mathbf{s}_{3}\). The screening is apparent from the table
because, for most models, \(\lambda_{1}\) and \(\lambda_{6}\) are smaller than
\(\lambda_{2}\) even though \(\mathbf{s}_{2}\) is a farther from
\(\mathbf{s}_{0}\). It is especially obvious for the gaussian model (F) where
\(\lambda_{1}\) and \(\lambda_{6}\) are negative.

\item%c
{\it What happens to the kriging variance?}

For all of the models except the pure nugget model, the kriging variance is
lower. This makes sense because moving some correlated observations closer to
the location of the prediction provides more information about
\(Z(\mathbf{s}_{0})\) and should reduce the amount of uncertainty in the
prediction.

\item%d
{\it In which models does the predicted value change and in which does it stay
the same?}

The predicted value stays the same for the pure nugget model, and it changes
for all the other models. For the models with spatial dependence, the
predicted value is pulled towards the values \(Z(\mathbf{s}_{3}) = 60\)
and \(Z(\mathbf{s}_{5}) = 50\) because those nearby points have the largest
weights.

\end{enumerate}

\pagebreak
\item%2
{\it We are going to use ordinary kriging to predict values of total carbon
over a grid. You should still have the carbon/nitrogen data set from the last
homework but let me know if you need to have it emailed to you. Choose your
grid to be the same as in the CN ratio example I worked in class. Set up R
code is below.}
<<prob2>>=
CN.dat <- read.table('../hw05/CN.dat', header = TRUE)
pred.grid <- expand.grid(seq(-50, 550, l = 100), seq(-15, 330, l = 100))
TC.geodat <- as.geodata(CN.dat, coords.col = 1:2, data.col = 4)
@
{\it We will work with total carbon (column labeled TC in the data set) in
this problem. You can use either the \texttt{ksline} or \texttt{krige.conv}
functions. Be a bit patient; it should take only a minute
or so. Don’t worry about anisotropy.}

\begin{enumerate}

\item%a
{\it Plot the data using}
<<prob2a, fig.height = 5>>=
plot(TC.geodat, breaks = 20)
@
{\it Do you see any evidence of trend? Outliers? Other potential anomalies?}

There appears to be a slight trend of decreasing total carbon to the north.
There is a possible low outlier in the northwest corner, and a possible high
outlier near \((200, 150)\), but neither of these look very extreme in the
histogram.

\pagebreak
\item%b
{\it Predict using global ordinary kriging. You estimated a semivariogram for
these data on the last homework and you can use those results. Prepare contour
plots of the predictions and kriging standard errors.}

On the previous assignment, I selected an exponential model (estimated by
Cressie's WLS method) with a nugget of 0.00241, partial sill of 0.01418, and
effective range of 154.62223.

<<prob2b, fig.height = 6, fig.width = 4, results = 'hide', cache = TRUE>>=
control <- krige.control(cov.model = 'exponential',
                         cov.pars = c(0.01418, 154.62223/3),
                         nugget = 0.00241)
TC_pred <- krige.conv(TC.geodat, locations = pred.grid, krige = control)
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
contour(TC_pred, main = 'Kriging Predictions')
contour(TC_pred, val = sqrt(TC_pred$krige.var), main = 'Kriging Standard Errors')
@

\item%c
{\it Evaluate the predictions using cross validation (CV2). You can use the R
code I provided in class notes.}

<<prob2c1, results = 'hide', cache = TRUE>>=
CV2_term <- function(i, geodata = TC.geodat, ...){
  pred <- krige.conv(coords = geodata$coords[-i,],
                     data = geodata$data[-i],
                     locations = rbind(geodata$coords[i,]), ...)
  return((geodata$data[i] - pred$predict) / sqrt(pred$krige.var))
}
CV2 <- sqrt(mean(sapply(seq_along(TC.geodat$data), CV2_term, krige = control)^2))
@
<<prob2c2>>=
CV2
@

\item%d
{\it Briefly discuss your results.}

The plots of the observed values in part (a) shows a random-looking arrangement
of colors and shapes, indicating that some large total carbon values occur
small values, but some banding is visible, suggesting that there is a trend.
Unsurprisingly, the kriging prediction surface is noisy because of the
variation between nearby measurements. The predictions tend to be higher in
the southern half of the site than in the northern half, so the carbon values
may be non-stationary and it would be worthwhile to fit a linear trend model.
However, the \(CV_{2}\) value is close to 1, meaning that the model accurately
predicts the observed values when one at a time is omitted. The standard error
surface shows small dips at the observed locations and larger standard errors
in the vertical gaps where no measurements were recorded.

\end{enumerate}

\end{enumerate}

\end{document}

### 2017-03-08 hw07/hw07.rnw
Got a little behind.

\documentclass[11pt]{article}
\usepackage{fullpage,graphicx,float,amsmath,enumitem,fancyhdr}

\setlist{parsep=5.5pt}
\setlength{\parindent}{0pt}

\newcommand\assignment{Stat 534 Homework 7}
\newcommand\myname{Kenny Flagg}
\newcommand\duedate{March 6, 2017}

\pagestyle{fancy}
\lhead{\assignment}
\chead{\duedate}
\rhead{\myname}
\setlength{\headheight}{18pt}
\setlength{\headsep}{2pt}

\title{\assignment}
\author{\myname}
\date{\duedate}

\begin{document}
\maketitle

<<setup, echo = FALSE, message = FALSE, cache = FALSE>>=
library(knitr)
library(extrafont)
opts_chunk$set(echo = TRUE, comment = NA, message = FALSE,
               show.signif.stars = FALSE,
               fig.align = 'center', fig.width = 6.5, fig.height = 4,
               fig.pos = 'H', size = 'footnotesize', dev = 'pdf',
               dev.args = list(family = 'CM Roman', pointsize = 11))
knit_theme$set('print')

library(geoR)
library(xtable)
options(xtable.floating = FALSE,
        xtable.sanitize.rownames.function = function(x) x,
        xtable.sanitize.colnames.function = function(x) x,
        width = 80, scipen = 2, show.signif.stars = FALSE)

set.seed(218923)
@

\begin{enumerate}

\item%1
{\it A data set (\texttt{wheat.txt}) has been sent to you. The data set
contains yields of wheat recorded at spatial coordinates. Note that the header
is x, y, and z with z being the yields. We will need a couple of different data
object types. Pay attention to the R code below. Do not worry about
anisotropy.}

<<prob1_1>>=
wheat.dat <- read.table('wheat.txt', header = TRUE)
wheat.geodat <- as.geodata(wheat.dat, coords.col = 1:2, data.col = 3)
wheat.grid <- expand.grid(seq(0, 50, l = 25), seq(0, 30, l = 25))
@

\begin{enumerate}

\item%a
{\it Plot the data and comment on the results.}

<<prob1a1>>=
plot(wheat.geodat, breaks = 30)
@

There is a trend in wheat yield, with the lowest yields occuring in the bottom
left of the region and the highest yields in the top right.

\item%b
{\it Produce a plot of the empirical semivariogram of the wheat yields. Can
this plot be trusted for estimation of semivariogram parameters to be used in
kriging. Why or why not?}

<<prob1b1, results = 'hide', fig.height = 5>>=
yield.semivar <- variog(wheat.geodat)
plot(yield.semivar, main = 'Empirical Semivariogram of Yield')
@

No, this plot cannot be trusted because the process is non-stationary so the
empirical semivariogram is biased.

\item%c
{\it We will use the \texttt{surf.ls} function in the spatial library to fit a
quadratic trend model to the yields by ordinary least squares and plot the
empirical semivariogram of the residuals.}
<<prob1c1>>=
library(spatial, quietly = TRUE)
wheat.ls <- surf.ls(2, wheat.dat) # fits a second order polynomial trend surface
resid.dat <- cbind(wheat.dat$x, wheat.dat$y, residuals(wheat.ls))
resid.geodat <- as.geodata(resid.dat, coords.col = 1:2, data.col = 3)
@
{\it Fit an appropriate semivariogram model to the semivariogram using your
method of choice. Justify your final selection.}

Based on the empirical semivariogram (plot on next page) I chose initial
values of 17 for the nugget, 10 for the (effective) range, and 25 for the
total sill (implying a partial sill of 8). I fit exponential and spherical
models by Cressie's WLS and by REML.

<<prob1c2, cache = TRUE, results = 'hide', fig.height = 5>>=
resid.semivar <- variog(resid.geodat)
exp.wls <- variofit(resid.semivar, ini.cov.pars = c(8, 10/3), nugget = 17,
                    fix.nugget = FALSE, cov.model = 'exponential', weights = 'cressie')
exp.reml <- likfit(resid.geodat, ini.cov.pars = c(8, 10/3), nugget = 17,
                   fix.nugget = FALSE, cov.model = 'exponential', lik.method = 'rml')
sph.wls <- variofit(resid.semivar, ini.cov.pars = c(8, 10), nugget = 17,
                    fix.nugget = FALSE, cov.model = 'spherical', weights = 'cressie')
sph.reml <- likfit(resid.geodat, ini.cov.pars = c(8, 10), nugget = 17,
                   fix.nugget = FALSE, cov.model = 'spherical', lik.method = 'rml')
plot(resid.semivar, main = 'Semivariogram of Residuals')
lines(exp.wls, lty = 1)
lines(exp.reml, lty = 2)
lines(sph.wls, lty = 3)
lines(sph.reml, lty = 4)
legend('bottomright', lty = 1:4,
       legend = c('Exponential, WLS', 'Exponential, REML', 'Spherical, WLS', 'Spherical, REML'))
@
<<prob1c3, eval = FALSE>>=
estimates <- rbind(
  `Exponential, WLS` = c(exp.wls$nugget, exp.wls$cov.pars * c(1, 3)),
  `Exponential, REML` = c(exp.reml$nugget, exp.reml$cov.pars * c(1, 3)),
  `Spherical, WLS` = c(sph.wls$nugget, sph.wls$cov.pars),
  `Spherical, REML` = c(sph.reml$nugget, sph.reml$cov.pars)
)
colnames(estimates) <- c('Nugget', 'Partial Sill', 'Range')
xtable(estimates)
@
\begin{center}
<<prob1c3, echo = FALSE, results = 'asis'>>=
@
\end{center}

On the plot, all four estimated models look very similar but the table above
shows some disagreement about the parameters. (Note that I multiplied the
exponential range parameter by 3 so the table shows the effective range for
the exponential models and the range for the spherical models.)

I ultimately chose the spherical model estimated by WLS because WLS does not
depend on normality and the spherical models fit the points at the lower lags
a bit better than the exponential models do.

\item%d
{\it Predict yields using universal kriging and ordinary kriging. Use the
parameter estimates from the residual semivariogram when you do ordinary
kriging. Plot the results along with a plot of the kriging standard errors.
Remember to be careful of that range parameter -- what you enter depends on
which semivariogram model you used.}
<<prob1d, cache = TRUE, dependson = 'prob1c2', results = 'hide', fig.height = 5>>=
# I'll just get the parameters straight out of the semivariogram model object.
wheat.uk <- ksline(wheat.geodat, locations = wheat.grid, cov.model = 'spherical',
                   cov.pars = sph.wls$cov.pars, nugget = sph.wls$nugget,
                   trend = 2, m0 = 'kt')
wheat.ok <- ksline(wheat.geodat, locations = wheat.grid, cov.model = 'spherical',
                   cov.pars = sph.wls$cov.pars, nugget = sph.wls$nugget)

par(mfrow = c(2, 2), mar = c(4, 3, 2, 1))
contour(wheat.uk, main = 'Universal Kriging Predicitons')
contour(wheat.uk, val = sqrt(wheat.uk$krige.var), main = 'Universal Kriging SEs')
contour(wheat.ok, main = 'Ordinary Kriging Predicitons')
contour(wheat.ok, val = sqrt(wheat.ok$krige.var), main = 'Ordinary Kriging SEs')
@
{\it Compare the results and comment.}

The predicted values from both methods are pretty similar, but the prediction
surface from universal kriging smoother than the prediction surface from the
ordinary kriging. The standard errors from universal kriging are larger
because they incorporate uncertainty from estimating the mean surface.

\end{enumerate}

\pagebreak
\item%2
{\it We looked at this example in class. We have a one-dimensional process with
point to point covariance function}
\begin{equation*}
C(s_{i}, s_{j}) = \exp\left(-\frac{3\left|s_{i}-s_{j}\right|}{5}\right).
\end{equation*}
{\it The nugget effect is 0, the sill is 1 and the practical range is 5. The
region \(B\) is defined to be the interval \(B = (2, 4)\) with \(|B| = 2\). The
point to block covariance function is}
\begin{equation*}
\sigma(B, s) = Cov(Z(B), Z(s)) = \frac{1}{2}\int_{2}^{4}
\exp\left(-\frac{3\left|u-s\right|}{5}\right)du.
\end{equation*}

\begin{enumerate}

\item%a
{\it Find the covariance function.}

There are three cases to consider: \(s<2\), \(s>4\), and \(s \in B\).

If \(s<2\) then \(u-s>0\) \(\forall u \in B\), so
\begin{align*}
\frac{1}{2}\int_{2}^{4}\exp\left(-\frac{3|u-s|}{5}\right)du
&= \frac{1}{2}\int_{2}^{4}\exp\left(-\frac{3(u-s)}{5}\right)du \\
&= \frac{1}{2}\left[-\frac{5}{3}\exp\left(-\frac{3(u-s)}
{5}\right)\right]_{u=2}^{4} \\
&= \frac{5}{6} \left[\exp\left(\frac{3}{5}(s-2)\right)
- \exp\left(\frac{3}{5}(s-4)\right)\right].
\end{align*}
If \(s>4\) then \(u-s<0\) \(\forall u \in B\), so
\begin{align*}
\frac{1}{2}\int_{2}^{4}\exp\left(-\frac{3|u-s|}{5}\right)du
&= \frac{1}{2}\int_{2}^{4}\exp\left(\frac{3(u-s)}{5}\right)du \\
&= \frac{1}{2}\left[\frac{5}{3}\exp\left(\frac{3(u-s)}
{5}\right)\right]_{u=2}^{4} \\
&= \frac{5}{6} \left[\exp\left(\frac{3}{5}(4-s)\right)
- \exp\left(\frac{3}{5}(2-s)\right)\right].
\end{align*}
Finally, if \(s \in B\) then
\begin{align*}
\frac{1}{2}\int_{2}^{4}\exp\left(-\frac{3|u-s|}{5}\right)du
&= \frac{1}{2}\int_{2}^{s}\exp\left(-\frac{3|u-s|}{5}\right)du
+ \frac{1}{2}\int_{s}^{4}\exp\left(-\frac{3|u-s|}{5}\right)du \\
&= \frac{1}{2}\int_{2}^{s}\exp\left(\frac{3(u-s)}{5}\right)du
+ \frac{1}{2}\int_{s}^{4}\exp\left(-\frac{3(u-s)}{5}\right)du \\
&= \frac{1}{2}\left[\frac{5}{3}\exp\left(\frac{3(u-s)}
{5}\right)\right]_{u=2}^{s}
+ \frac{1}{2}\left[-\frac{5}{3}\exp\left(-\frac{3(u-s)}
{5}\right)\right]_{u=s}^{4} \\
&= \frac{5}{6} \left[\exp(0) - \exp\left(\frac{3}{5}(2-s)\right)
- \exp\left(\frac{3}{5}(s-4)\right) + \exp(0)\right] \\
&= \frac{5}{6} \left[2 - \exp\left(\frac{3}{5}(2-s)\right)
- \exp\left(\frac{3}{5}(s-4)\right)\right].
\end{align*}
So the point-to-block covariance function is
\begin{equation*}
\sigma(B,s) = \frac{5}{6}\begin{cases}
\exp\left(\frac{3}{5}(s-2)\right) - \exp\left(\frac{3}{5}(s-4)\right) &
s<2 \\
2 - \exp\left(\frac{3}{5}(2-s)\right) - \exp\left(\frac{3}{5}(s-4)\right) &
2 \leq s \leq 4 \\
\exp\left(\frac{3}{5}(4-s)\right) - \exp\left(\frac{3}{5}(2-s)\right) &
s>4
\end{cases}.
\end{equation*}
% \(\sigma(B,3)=0.7519806\) here.
% Numerical integration is the notes got 0.7519603.

\item%b
{\it Find \(\sigma(B, B)\).}

The block-to-block covariance is %variance of the block mean?
\begin{align*}
\sigma(B, B) &= \frac{1}{4}\int_{2}^{4}\int_{2}^{4}\sigma(u,s)duds \\
&= \frac{1}{2}\int_{2}^{4}\sigma(B,s)ds \\
&= \frac{1}{2} \int_{2}^{4} \frac{5}{6}\left(2
- \exp\left(\frac{3}{5}(2-s)\right)
- \exp\left(\frac{3}{5}(s-4)\right)\right)ds \\
&= \frac{5}{12} \left[2s + \frac{5}{3}\exp\left(\frac{3}{5}(2-s)\right)
- \frac{5}{3}\exp\left(\frac{3}{5}(s-4)\right)\right]_{s=2}^{4} \\
&= \frac{5}{12} \left[8 + \frac{5}{3}\exp\left(-\frac{6}{5}\right)
- \frac{5}{3}\exp(0) - 4 - \frac{5}{3}\exp(0)
+ \frac{5}{3}\exp\left(-\frac{6}{5}\right)\right] \\
&= \frac{5}{12} \left[4 - \frac{10}{3}
+ \frac{10}{3}\exp\left(-\frac{6}{5}\right)\right] \\
&= \frac{5}{12} \left[\frac{2}{3}
+ \frac{10}{3}\exp\left(-\frac{6}{5}\right)\right] \\
&= \frac{5}{18} + \frac{25}{18}\exp\left(-\frac{6}{5}\right).
\end{align*}
% \(\sigma(B,B)=0.6961031\) here.
% Numerical integration is the notes got 0.6958757.

\end{enumerate}

\end{enumerate}

\end{document}

### 2017-03-12 exam1/exam1.rnw
midterm

\documentclass[11pt]{article}
\usepackage{fullpage,graphicx,float,amsmath,enumitem,fancyhdr}

\setlist{parsep=5.5pt}
\setlength{\parindent}{0pt}

\newcommand\assignment{Stat 534 Exam 1}
\newcommand\myname{Kenny Flagg}
\newcommand\duedate{March 10, 2017}

\pagestyle{fancy}
\lhead{\assignment}
\chead{\duedate}
\rhead{\myname}
\setlength{\headheight}{18pt}
\setlength{\headsep}{2pt}

\title{\assignment}
\author{\myname}
\date{\duedate}

\begin{document}
\maketitle

<<setup, echo = FALSE, message = FALSE, cache = FALSE>>=
library(knitr)
library(extrafont)
opts_chunk$set(echo = TRUE, comment = NA, message = FALSE,
               show.signif.stars = FALSE,
               fig.align = 'center', fig.width = 6.5, fig.height = 4,
               fig.pos = 'H', size = 'footnotesize', dev = 'pdf',
               dev.args = list(family = 'CM Roman', pointsize = 11))
knit_theme$set('print')

library(geoR)
library(nlme)
library(xtable)
options(xtable.floating = FALSE,
        xtable.sanitize.rownames.function = function(x) x,
        xtable.sanitize.colnames.function = function(x) x,
        width = 80, scipen = 2, show.signif.stars = FALSE)

set.seed(87624)
@

\begin{enumerate}

\item%1
{\it Suppose we have an intrinsically stationary process with semivariogram}
\begin{equation*}
\left(\frac{1}{2}\right)Var\left(Z(\mathbf{s}_{i})-Z(\mathbf{s}_{j})\right)
= \gamma\left(\mathbf{s}_{i}-\mathbf{s}_{j}\right)
= \gamma\left(\mathbf{h}_{ij}\right)
\end{equation*}
{\it On an earlier homework you showed that}
\begin{equation*}
\sum_{i=1}^{n}\sum_{j=1}^{n}a_{i}a_{j}\gamma\left(\mathbf{h}_{ij}\right)
\leq 0
\end{equation*}
{\it for any sites \(\mathbf{s}_{i}\), \(i=1,\dots,n\) and for any constants
\(a_{i}\), \(i=1,\dots,n\) with \(\sum_{i=1}^{n}a_{i}=0\) but you did it under
an assumption of second order stationarity. We will now establish it in
general.}

\begin{enumerate}
\item%a
{\it First show that \(\displaystyle
-\left(\frac{1}{2}\right)\left\{\sum_{i=1}^{n}\sum_{j=1}^{n}
a_{i}a_{j}\left(Z(\mathbf{s}_{i})-Z(\mathbf{s}_{j})\right)^{2}\right\}
= \left\{\sum_{i=1}^{n}a_{i}Z(\mathbf{s}_{j})\right\}^{2}\).}

Proof:
\begin{align*}
-\left(\frac{1}{2}\right)\left\{\sum_{i=1}^{n}\sum_{j=1}^{n}
a_{i}a_{j}\left(Z(\mathbf{s}_{i})-Z(\mathbf{s}_{j})\right)^{2}\right\}
&= -\left(\frac{1}{2}\right)\left\{\sum_{i=1}^{n}\sum_{j=1}^{n}
a_{i}a_{j}\left(Z(\mathbf{s}_{i})^{2}
- 2Z(\mathbf{s}_{i})Z(\mathbf{s}_{j})
+ Z(\mathbf{s}_{j})^{2}\right)\right\} \\
&= -\frac{1}{2}\sum_{i=1}^{n}\sum_{j=1}^{n}a_{i}a_{j}Z(\mathbf{s}_{i})^{2}
+ \sum_{i=1}^{n}\sum_{j=1}^{n}a_{i}a_{j}Z(\mathbf{s}_{i})Z(\mathbf{s}_{j}) \\
&\qquad - \frac{1}{2} \sum_{i=1}^{n}\sum_{j=1}^{n}a_{i}a_{j}
Z(\mathbf{s}_{j})^{2} \\
&= -\frac{1}{2}\sum_{i=1}^{n}a_{i}Z(\mathbf{s}_{i})^{2}\sum_{j=1}^{n}a_{j}
+ \sum_{i=1}^{n}a_{i}Z(\mathbf{s}_{i})\sum_{j=1}^{n}a_{j}Z(\mathbf{s}_{j}) \\
&\qquad - \frac{1}{2}\sum_{j=1}^{n}a_{j}Z(\mathbf{s}_{j})^{2}
\sum_{i=1}^{n}a_{i} \\
&= 0 + \sum_{i=1}^{n}a_{i}Z(\mathbf{s}_{i})
\sum_{j=1}^{n}a_{j}Z(\mathbf{s}_{j}) + 0 \\
&= \left\{\sum_{i=1}^{n}a_{i}Z(\mathbf{s}_{i})\right\}^{2}.
\end{align*}

\item%b
{\it Now take expectations of both sides to establish the result.}

\begin{align*}
\sum_{i=1}^{n}\sum_{j=1}^{n}a_{i}a_{j}\gamma\left(\mathbf{h}_{ij}\right)
&= \left\{\sum_{i=1}^{n}\sum_{j=1}^{n}a_{i}a_{j}\left(\frac{1}{2}\right)
\left(E\left[\left(Z(\mathbf{s}_{i})-Z(\mathbf{s}_{j})\right)^{2}
\right]-\left[E(Z(\mathbf{s}_{i})-Z(\mathbf{s}_{j}))\right]^{2}
\right)\right\} \\
&= E\left[\left(\frac{1}{2}\right)\left\{\sum_{i=1}^{n}\sum_{j=1}^{n}
a_{i}a_{j}\left(Z(\mathbf{s}_{i})-Z(\mathbf{s}_{j})\right)^{2}
\right\}\right]
- \frac{1}{2}\sum_{i=1}^{n}\sum_{j=1}^{n}a_{i}a_{j}
(\mu(\mathbf{s}_{i})-\mu(\mathbf{s}_{j}))^{2} \\
&= E\left[-\left\{\sum_{i=1}^{n}a_{i}Z(\mathbf{s}_{j})\right\}^{2}\right]
- \frac{1}{2}\sum_{i=1}^{n}\sum_{j=1}^{n}a_{i}a_{j}\mu(\mathbf{s}_{i})^{2}
+ \sum_{i=1}^{n}\sum_{j=1}^{n}a_{i}a_{j}
\mu(\mathbf{s}_{i})\mu(\mathbf{s}_{j}) \\
&\qquad - \frac{1}{2}\sum_{i=1}^{n}\sum_{j=1}^{n}a_{i}a_{j}
\mu(\mathbf{s}_{j})^{2} \\
&= -E\left[\left\{\sum_{i=1}^{n}a_{i}Z(\mathbf{s}_{j})\right\}^{2}\right] 
- \frac{1}{2}\sum_{i=1}^{n}a_{i}\mu(\mathbf{s}_{i})^{2}\sum_{j=1}^{n}a_{j}
+ \sum_{i=1}^{n}a_{i}\mu(\mathbf{s}_{i})
\sum_{j=1}^{n}a_{j}\mu(\mathbf{s}_{j}) \\
&\qquad - \frac{1}{2}\sum_{i=1}^{n}a_{i}\sum_{j=1}^{n}a_{j}
\mu(\mathbf{s}_{j})^{2}\\
&\leq 0.
\end{align*}

\end{enumerate}

\item%2
{\it Let \(X_{0} \sim Gamma(\alpha, \beta)\) with the parameterization}
\begin{equation*}
f(x)=\frac{1}{\Gamma(\alpha)\beta^{\alpha}}x^{\alpha-1}
\exp(-x/\beta); \qquad x > 0
\end{equation*}
{\it and 0 elsewhere. Let \(X_{i} \sim Gamma(\alpha_{i}, \beta)\) for
\(i=1,\dots,n\). We construct a one-dimensional regularly spaced random field
at locations \(i=1,\dots,n\)}
\begin{equation*}
Z(s_{i}) = X_{0} + X_{i}; \qquad i=1,\dots,n.
\end{equation*}
{\it You can assume that \(X_{0},X_{1},\dots,X_{n}\) are independent.}

\begin{enumerate}

\item%a
{\it What is the distribution of \(Z(s_{i})\)?}

Because \(X_{0}\) and \(X_{i}\) are independent Gamma random variables,
\begin{equation*}
Z(s_{i}) = X_{0} + X_{i} \sim \mathrm{Gamma}(\alpha + \alpha_{i}, \beta).
\end{equation*}

\item%b
{\it Find \(E(Z(s_{i}))\) and \(Var(Z(s_{i}))\). You can use known properties
of the Gamma distribution to answer this question, i.e. you can just write
down the answer if you know it or can find it.}

\begin{equation*}
E(Z(s_{i})) = \frac{\alpha + \alpha_{i}}{\beta}
\qquad\text{ and }\qquad
Var(Z(s_{i})) = \frac{\alpha + \alpha_{i}}{\beta^{2}}.
\end{equation*}

%\pagebreak
\item%c
{\it Find \(Cov(Z(s_{i}), Z(s_{j}))\) for \(i \neq j\).}

If \(i \neq j\),
\begin{align*}
Cov(Z(s_{i}), Z(s_{j})) &= Cov(X_{0} + X_{i}, X_{0} + X_{j}) \\
&= Var(X_{0}) + Cov(X_{0}, X_{j}) + Cov(X_{i}, X_{0}) + Cov(X_{i}, X_{j}) \\
&= \frac{\alpha}{\beta^{2}} + 0 + 0 + 0 \\
&= \frac{\alpha}{\beta^{2}}.
\end{align*}

\item%d
{\it Is this a second-order stationary process? Justify your answer.}

This process is not second order stationary because the mean and variance of
\(Z(s_{i})\) vary by location.

\end{enumerate}

\item%3
{\it The \texttt{lansing} data set in the \texttt{spatstat} package contains
spatial locations of several different species of trees. We will be looking
and comparing the distributions of black oaks and maples.}
<<prob3_1, cache = TRUE>>=
require(spatstat)
data(lansing)
blackoak <- split(lansing)$blackoak
maple <- split(lansing)$maple
@
{\it The rectangular region is a unit square. Use the isotropic edge
corrected version when applicable below. Answer the following questions. You
will be computing several simulation envelopes below. Be patient and keep
nsim=99, the default.}

\begin{enumerate}

\item%a
{\it What does the \(K\) function measure?}

The \(K\) function measures the average number of events within a radius
\(h\) of a randomly-selected event, scaled by the intensity over the whole
region.

\item%b
{\it It is often easier to interpret the \(L\) function than the \(K\)
function. Based on the \(L\) function do the black oaks appear to be clustered
or do they appear to be regularly distributed? Do the maples appear to be
clustered or do they appear to be regularly distributed? Justify your answer.
Simulation envelopes will help you give a better answer to this question.}

<<prob3b1, cache = TRUE, dependson = 'prob3_1', results = 'hide'>>=
par(mfrow = c(1, 2))
Lblackoak <- envelope(blackoak, fun = Lest, correction = 'iso')
plot(Lblackoak,.-r ~ r, legend = FALSE)
Lmaple <- envelope(maple, fun = Lest, correction = 'iso')
plot(Lmaple,.-r ~ r, legend = FALSE)
@
The black oaks appear to be clustered because the \(L\) function is larger
than expected under complete spatial randomness, meaning trees tend to be
closer together than expected. Likewise, the maples also appear to be
clustered because their \(L\) function is larger than expected under CSR.

\item%c
{\it Compare the two \(L\) functions and discuss whether or not the 2
processes appear to be the same. You can use the results from (a) but you
should also look at the difference more formally using the following also
provided in an attached script file.}
<<prob3c1, cache = TRUE, dependson = 'prob3_1', results = 'hide'>>=
require(splancs)
# specify radii
h <- seq(0, 0.5, l = 100)
# get coordinates
tree.poly <- list(x = c(blackoak$x, maple$x), y = c(blackoak$y, maple$y))
# recompute the K functions
kblackoak <- khat(as.points(blackoak), bboxx(bbox(as.points(tree.poly))), h)
kmaple <- khat(as.points(maple), bboxx(bbox(as.points(tree.poly))), h)
# get the differences
k.diff <- kblackoak - kmaple
# generate the envelope
env <- Kenv.label(as.points(blackoak), as.points(maple),
                  bboxx(bbox(as.points(tree.poly))), nsim = 99, s = h)
# plot the results
plot(h, seq(-0.15, 0.05, l = length(h)), type = 'n', ylab = 'Kdiff',
     main = 'Envelopes for Kdiff')
lines(h, k.diff)
lines(h, env$low, lty = 2)
lines(h, env$up, lty = 2)
abline(h = 0)
@

At a glance, the plots in part (b) indicate that the processes are similar.
For both, \(L(h)-h\) increases quickly until about \(h=0.15\) units and then
levels off. However, the \(L\) function for black oaks is a bit larger than
the \(L\) function for maples. We formally test equality of the \(L\) or \(K\)
functions with a simulation envelope for \(K_{oak}(h) - K_{maple}(h)\). The
observed difference in \(K\) functions is not contained entirely in the
envelope, so we have evidence that the locations of the two tree species result
from two separate processes.

\pagebreak
\item%d
{\it Plot \(L_{ij} - h\) versus \(h\) for black oaks and maples.}
<<prob3d1, cache = TRUE, dependson = 'prob3_1', results = 'hide'>>=
Kplot <- envelope(lansing, Kcross, i = 'blackoak', j = 'maple')
plot(Kplot, sqrt(./pi)-r ~ r, ylab = expression(L[ij] - h),
     main = 'Cross L Function', legend = FALSE)
@
{\it What type of relationship between the point patterns of the two species
of trees is indicated by this plot? Justify your answer.}

The cross \(L\) function is smaller than expected under independence,
indicating that fewer trees of one species than expected are found within a
given distance of a tree of the other species. This suggests an inhibitory
process where trees of the same species tend to cluster together rather than
growing near trees of the other species.

\item%e
{\it Based on the above, comment on the null hypotheses of independence and
random labeling.}

In part (c) we found evidence that the processes differ between the species,
so we reject the hypothesis of random labeling. In part (d) we found evidence
that trees of one species tend not to grow near trees of the other species,
so we reject the hypothesis of independence.

\end{enumerate}

\pagebreak
\item%4
{\it You were sent the wheat data set on a previous homework assignment. You
want to predict the value of \(Z\) (yield) at an arbitrary location. Assume a
pure nugget effect model.}

\begin{enumerate}

\item%a
{\it What are the kriging weights and what is the predicted value?}

<<prob4a1>>=
wheat <- read.table('wheat.txt', header = TRUE)
nrow(wheat)
mean(wheat$z)
@

There are \(n = \Sexpr{nrow(wheat)}\) observations so the kriging weights are
\(\lambda_{i} = \frac{1}{\Sexpr{nrow(wheat)}}\) for all \(i\). The predicted
value is simply \(\bar{Z} = \Sexpr{sprintf('%.2f', mean(wheat$z))}\).

\item%b
{\it What is the estimate of the sill?}

<<prob4b1>>=
var(wheat$z)
@

The estimate of the sill is the sample variance, \(\widehat{\sigma}^{2}
= \Sexpr{sprintf('%.2f', var(wheat$z))}\).

\item%c
{\it What is the kriging standard error (note that this is a prediction
error)?}

The kriging standard error is \(\displaystyle
\sqrt{\widehat{Var}(Z) + \widehat{Var}(\bar{Z})}
= \sqrt{\widehat{\sigma}^{2} + \frac{\widehat{\sigma}^{2}}{n}}
= \Sexpr{sprintf('%.2f', var(wheat$z) * (1 + 1/nrow(wheat)))}\).

\end{enumerate}

\item%5
{\it Carbon-Nitrogen data example: We looked at estimating the semivariogram
of the residuals from a simple linear regression model of total carbon on
total nitrogen in class. We used \texttt{gls} to do this (as part of
incorporating a spatial covariance structure into the regression) specifying
an exponential covariance model and estimating the parameters using both
maximum likelihood and REML. Let's check to see what the \texttt{likfit}
function in the \texttt{geoR} package would return as parameter estimates
(nugget, practical range, and partial sill) and see if the results are
comparable. Some of the relevant R code is included in the attached script
file. Compare the estimates on page 11 of the Spatial Regression notes and
the estimates you get out of \texttt{likfit}. Use the same starting values.}

<<prob5_1, cache = TRUE, results = 'hide'>>=
# Get the CN data.
CN.dat <- read.table('CN.dat', header = TRUE)
names(CN.dat) <- c('x', 'y', 'tn', 'tc', 'cn')
CN.lm <- lm(tc ~ tn, data = CN.dat)
resids <- residuals(CN.lm)

# Convert to a geodata object.
require(geoR)
resids.dat <- cbind(CN.dat$x, CN.dat$y, resids)
resids.dat <- data.frame(resids.dat)
names(resids.dat) <- c('x', 'y', 'resids')
resids.geodat <- as.geodata(resids.dat, coords.col = 1:2, data.col = 3)

# Use likfit() on the residuals.
CNlikexp.ml <- likfit(resids.geodat, cov.model = 'exponential',
                      ini.cov.pars = c(0.00159, 15),
                      fix.nugget = TRUE, nugget = 0,
                      lik.method = 'ML')
CNlikexp.reml <- likfit(resids.geodat, cov.model = 'exponential',
                        ini.cov.pars = c(0.00159, 15),
                        fix.nugget = TRUE, nugget = 0,
                        lik.method = 'REML')
CNlikexp.ml.nugget <- likfit(resids.geodat, cov.model = 'exponential',
                             ini.cov.pars = c(0.00159, 15),
                             fix.nugget = FALSE, nugget = 0.4 * 0.00159,
                             lik.method = 'ML')
CNlikexp.reml.nugget <- likfit(resids.geodat, cov.model = 'exponential',
                               ini.cov.pars = c(0.00159, 15),
                               fix.nugget = FALSE, nugget = 0.4 * 0.00159,
                               lik.method = 'REML')

# Now fit gls the models.
CNglsexp.ml <- gls(tc ~ tn, data = CN.dat, method = 'ML',
                   correlation = corExp(15, form = ~x+y))
CNglsexp.reml <- gls(tc ~ tn, data = CN.dat, method = 'REML',
                     correlation = corExp(15, form = ~x+y))
CNglsexp.ml.nugget <- gls(tc ~ tn, data = CN.dat, method = 'ML',
                          correlation = corExp(c(15, 0.4), form = ~x+y,
                                                nugget = TRUE))
CNglsexp.reml.nugget <- gls(tc ~ tn, data = CN.dat, method = 'REML',
                            correlation = corExp(c(15, 0.4), form = ~x+y,
                                                 nugget = TRUE))
@
<<prob5_2, eval = FALSE>>=
# Put all this in a nice table.
compare <- data.frame(
  Model = paste('Exponential', rep(c('REML', 'ML'), 2)),
  'Nugget (gls)' = c(NA, NA,
                     coef(CNglsexp.reml.nugget$modelStruct$corStruct, unconstrained = FALSE)['nugget'],
                     coef(CNglsexp.ml.nugget$modelStruct$corStruct, unconstrained = FALSE)['nugget']),
  'Nugget (likfit)' = c(NA, NA,
                        CNlikexp.reml.nugget$nugget /
                          (CNlikexp.reml.nugget$sigmasq + CNlikexp.reml.nugget$nugget),
                        CNlikexp.ml.nugget$nugget /
                          (CNlikexp.ml.nugget$sigmasq + CNlikexp.ml.nugget$nugget)),
   'Sill (gls)' = c(CNglsexp.reml$sigma^2,
                    CNglsexp.ml$sigma^2,
                    CNglsexp.reml.nugget$sigma^2,
                    CNglsexp.ml.nugget$sigma^2),
   'Sill (likfit)' = c(CNlikexp.reml$sigmasq,
                       CNlikexp.ml$sigmasq,
                       CNlikexp.reml.nugget$sigmasq + CNlikexp.reml.nugget$nugget,
                       CNlikexp.ml.nugget$sigmasq + CNlikexp.ml.nugget$nugget),
   'Range (gls)' = 3 * c(coef(CNglsexp.reml$modelStruct$corStruct, unconstrained = FALSE)['range'],
                         coef(CNglsexp.ml$modelStruct$corStruct, unconstrained = FALSE)['range'],
                         coef(CNglsexp.reml.nugget$modelStruct$corStruct, unconstrained = FALSE)['range'],
                         coef(CNglsexp.ml.nugget$modelStruct$corStruct, unconstrained = FALSE)['range']),
   'Range (likfit)' = c(CNlikexp.reml$practicalRange,
                        CNlikexp.ml$practicalRange,
                        CNlikexp.reml.nugget$practicalRange,
                        CNlikexp.ml.nugget$practicalRange),
   check.names = FALSE
)
print(xtable(compare, digits = 5), include.rownames = FALSE, size = 'footnotesize')
@
<<prob5_2, echo = FALSE, results = 'asis'>>=
@

Getting the output to be comparable took some work because \texttt{gls}'s
``nugget'' is the proportion of the total sill and its ``sill'' is the total
sill, while \texttt{likfit}'s ``nugget'' is the actual nugget and its
``sill'' is the partial sill. To create the table above I adjusted the
\texttt{likfit} output, adding the nugget to the partial sill so the ``sill''
column shows the total sill, and dividing the nugget by the total sill so the
nugget column is the proportion.

The practical range estimates from \texttt{likfit} are a little smaller than
those from \texttt{gls} but the sill and nugget estimates are essentially the
same.

\end{enumerate}

\end{document}

### 2017-03-12 exam1/Exam1Rcode.R

@@ -1,0 +1,56 @@
+# Some (but not all) R code that may be useful. 
+# Problem 3
+require(spatstat)
+
+blackoak<-split(lansing)$blackoak
+maple<-split(lansing)$maple
+
+Lblackoak<-envelope(blackoak,fun=Lest,correction="iso")
+plot(Lblackoak,.-r~r,legend=F)
+Lmaple<-envelope(maple,fun=Lest,correctin="iso")
+plot(Lmaple,.-r~r,legend=F)
+
+require(splancs)
+# specify radii
+h<-seq(0,.5,l=100)
+# get coordinates
+tree.poly<-list(x=c(blackoak$x,maple$x),y=c(blackoak$y,maple$y))
+# recompute the K functions
+kblackoak<-khat(as.points(blackoak),bboxx(bbox(as.points(tree.poly))),h)
+kmaple<-khat(as.points(maple),bboxx(bbox(as.points(tree.poly))),h)
+# get the differences
+k.diff<-kblackoak - kmaple
+# generate the envelope
+env<-Kenv.label(as.points(blackoak),as.points(maple),
+bboxx(bbox(as.points(tree.poly))),nsim=99,s=h)
+# plot the results
+plot(h,seq(-0.15,0.05,l=length(h)),type="n",ylab="Kdiff",
+main="Envelopes for Kdiff")
+lines(h,k.diff)
+lines(h,env$low,lty=2)
+lines(h,env$up,lty=2)
+abline(h=0)
+
+Kenv<-envelope(lansing,Kcross, i="maple",j="blackoak")
+plot(Kenv,sqrt(./pi)-r~r,ylab="Lij - h",main="Cross L Function - Maple
+and Black Oak",legend=F)
+
+plot(density(blackoak))
+plot(density(maples))
+
+# Problem 5
+
+# get the CN data
+names(CN.dat)<-c("x","y","tn","tc","cn")
+attach(CN.dat)
+CN.lm<-lm(tc~tn)
+resids<-residuals(CN.lm)
+# convert to a geodata object 
+require(geoR)
+resids.dat<-cbind(CN.dat$x,CN.dat$y,resids)
+resids.dat<-data.frame(resids.dat)
+names(resids.dat)<-c("x","y","resids")
+resids.geodat<-as.geodata(resids.dat,coords.col=1:2,data.col=3)
+# now fit the models. 
+
+

### 2017-03-12 project/paper.rnw
Project skeleton.

\documentclass[11pt]{article}
\usepackage{fullpage,graphicx,float,amsmath,enumitem,fancyhdr}

\setlist{parsep=5.5pt}
\setlength{\parindent}{0pt}
\setlength{\parskip}{5.5pt}

\usepackage[backend=bibtex,style=authoryear,citestyle=authoryear-comp]{biblatex}
\addbibresource{references.bib}

\newcommand\assignment{Stat 534 Project: Point Process Intensity Surface Models}
\newcommand\myname{Kenny Flagg}
\newcommand\duedate{\today}

\pagestyle{fancy}
\lhead{Stat 534 Project}
\chead{\duedate}
\rhead{\myname}
\setlength{\headheight}{18pt}
\setlength{\headsep}{2pt}

\title{\assignment}
\author{\myname}
\date{\duedate}

\begin{document}
\maketitle

<<setup, echo = FALSE, message = FALSE, cache = FALSE>>=
library(knitr)
library(extrafont)
opts_chunk$set(echo = FALSE, comment = NA, message = FALSE,
               fig.path = 'proj_figure/', cache.path = 'proj_cache/',
               show.signif.stars = FALSE,
               fig.align = 'center', fig.width = 6.5, fig.height = 3,
               fig.pos = 'H', size = 'footnotesize', dev = 'pdf',
               dev.args = list(family = 'CM Roman', pointsize = 11))
knit_theme$set('print')

#library(xtable)
options(xtable.table.placement = 'H', width = 80, scipen = 2,
        show.signif.stars = FALSE)
@

\section{Introduction}

\section{Surface Fitting}

\textcite{baddeleycoeurjolly14,baddeleyturner,diggleetal,bermanturner}

\section{Simple Example}

\section{Simulation}

\printbibliography

\end{document}

### 2017-03-13 project/paper.rnw
More structure.

@@
 \documentclass[11pt]{article}
-\usepackage{fullpage,graphicx,float,amsmath,enumitem,fancyhdr}
+\usepackage{fullpage,graphicx,float,amsmath,enumitem,fancyhdr,hyperref}
 
@@
 opts_chunk$set(echo = FALSE, comment = NA, message = FALSE,
-               fig.path = 'proj_figure/', cache.path = 'proj_cache/',
+               fig.path = 'paper_figure/', cache.path = 'paper_cache/',
                show.signif.stars = FALSE,
@@
         show.signif.stars = FALSE)
+
+library(spatstat)
+
+set.seed(83534)
 @
@@
 
+unobserved point processes -- only observe events in subregions selected by
+practitioner but want to make inference about inhomogeneous intensity over
+entire region
+
+know something is in the vicinity but want to know where specifically --
+examples: endangered clustered plant species in underbrush of thick forest,
+conservationists want to know where the clusters are; subsurface geomagnetic
+anomalies that could be unexploded ordnance -- concerned with mapping the
+intensity of this realization, not estimating parameters of the process
+
 \section{Surface Fitting}
 
-\textcite{baddeleycoeurjolly14,baddeleyturner,diggleetal,bermanturner}
+\textcite{wp,digglebook,baddeleycoeurjolly14,baddeleyturner,diggleetal,bermanturner}
 
-\section{Simple Example}
+\section{Simple Examples}
 
-\section{Simulation}
+loglinear example from HW4
+
+<<hw4_1>>=
+observe_win <- setminus.owin(owin(c(0, 1), c(0, 1)),
+                             owin(c(0.5, 0.8), c(0.5, 0.8)))
+predict_win <- owin(c(-0.5, 1.5), c(-0.5, 1.5))
+hw4 <- rpoispp(function(x, y){exp(5 * x + 2 * y)}, win = observe_win)
+par(mar = c(0, 0, 0.5, 0))
+plot(hw4)
+@
+
+<<hw4_2>>=
+hw4_fit <- ppm(hw4, ~ x + y, correction = 'isotropic')
+hw4_pred <- predict(hw4_fit, window = predict_win, se = TRUE)
+par(mfrow = c(1, 2), mar = c(0, 0, 0.5, 2))
+plot(hw4_pred$estimate)
+plot(observe_win, add = TRUE, col = '#ffffff40', border = NA)
+points(hw4, pch = '.', col = 'white')
+plot(hw4_pred$se)
+plot(observe_win, add = TRUE, col = '#ffffff40', border = NA)
+points(hw4, pch = '.', col = 'white')
+@
+
+easy site, realization 2,000, 390 ft spacing
+
+\section{Simulation Study}
+
+\section{Discussion and Conclusion}
+
+\appendix
+\section{R Code Appendix}
 

### 2017-03-14 project/paper.rnw
Some plots and equations for the first example.

@@
 
-#library(xtable)
+library(xtable)
 options(xtable.table.placement = 'H', width = 80, scipen = 2,
+        xtable.sanitize.rownames.function = function(x){return(x)},
         show.signif.stars = FALSE)
@@
 
-\textcite{wp,digglebook,baddeleycoeurjolly14,baddeleyturner,diggleetal,bermanturner}
+\textcite{wp,digglebook,bermanturner}
 
@@
 
-loglinear example from HW4
+log-linear example from HW4
 
-<<hw4_1>>=
+simulate from
+\begin{equation*}
+\log(\lambda(x,y)) = 5x + 2y; \quad (x, y) \in (0, 1)^{2}
+\end{equation*}
+but do not observe any events in \((0.5, 0.8)^{2}\). Then fit
+\begin{equation*}
+\log(\lambda(x,y)) = \beta_{0} + \beta_{1}x + \beta_{2}y
+\end{equation*}
+and predict on \((-0.5, 1.5)^{2}\).
+
+<<hw4_sim>>=
 observe_win <- setminus.owin(owin(c(0, 1), c(0, 1)),
@@
 hw4 <- rpoispp(function(x, y){exp(5 * x + 2 * y)}, win = observe_win)
-par(mar = c(0, 0, 0.5, 0))
-plot(hw4)
 @
 
-<<hw4_2>>=
+<<hw4_plot, fig.cap = 'One realization of the loglinear intensity on a subset of the unit square.'>>=
+par(mar = c(0, 0, 1, 0))
+plot(hw4, main = 'Event Locations')
+@
+
+<<hw4_fit>>=
 hw4_fit <- ppm(hw4, ~ x + y, correction = 'isotropic')
 hw4_pred <- predict(hw4_fit, window = predict_win, se = TRUE)
-par(mfrow = c(1, 2), mar = c(0, 0, 0.5, 2))
-plot(hw4_pred$estimate)
-plot(observe_win, add = TRUE, col = '#ffffff40', border = NA)
+# By default hw4_pred is on a 128 by 128 grid.
+@
+
+<<hw4_fitcoefs, results = 'asis'>>=
+hw4_coef <- summary(hw4_fit)$coefs.SE.CI[,c('Estimate', 'S.E.')]
+rownames(hw4_coef) <- c('\\(\\widehat{\\beta_{0}}\\)',
+                        '\\(\\widehat{\\beta_{1}}\\)',
+                        '\\(\\widehat{\\beta_{2}}\\)')
+xtable(hw4_coef, label = 'hw4est',
+       caption = 'Estimated coefficients for the log-linear trend model.')
+@
+
+<<hw4_fitplot, fig.cap = 'Estimated log-intensity surface and log-scale prediction standard errors.'>>=
+par(mfrow = c(1, 2), mar = c(0, 0, 1.2, 2), las = 2)
+plot(log(hw4_pred$estimate), zlim = c(-4.5, 10.5),
+     main = expression(log(hat(lambda)(italic(list(x,y))))))
+plot(observe_win, add = TRUE, border = '#ffffff40')
 points(hw4, pch = '.', col = 'white')
-plot(hw4_pred$se)
-plot(observe_win, add = TRUE, col = '#ffffff40', border = NA)
+plot(log(hw4_pred$se), zlim = c(-4.5, 10.5),
+     main = expression(log(SE(hat(lambda)(italic(list(x,y)))))))
+plot(observe_win, add = TRUE, border = '#ffffff40')
 points(hw4, pch = '.', col = 'white')

### 2017-03-14 project/paper.rnw
Something like an intro.

--- 2017-03-12 project/paper.rnw
@@
 
-\newcommand\assignment{Stat 534 Project: Point Process Intensity Surface Models}
+\newcommand\assignment{Stat 534 Project: \\
+Extrapolation from Poisson Process Intensity Surface Models}
 \newcommand\myname{Kenny Flagg}
@@
 
-#library(xtable)
+library(xtable)
 options(xtable.table.placement = 'H', width = 80, scipen = 2,
+        xtable.sanitize.rownames.function = function(x){return(x)},
         show.signif.stars = FALSE)
+
+library(spatstat)
+
+set.seed(83534)
 @
@@
 
+One little-studied use of inhomogeneous point process intensity estimation
+is the use of an intensity surface model fit to events in a subregion to
+predict the intensity over the entire region of interest. This procedure would
+be relevant whenver it is known or suspected that some type of plant, animal,
+or other item occurs in the region, and the goal of the analysis is to map the
+trend in where these tend to be located rather than estimate parameters of
+some process at work across hypothetical replicates of similar regions. It may
+be prohibitively expensive or difficult to observe all events over the entire
+region, so a sample of subregions is taken. For example, conservationists want
+to study the spatial distribution of an endangered plant across a large region
+of thick jungle so that they can establish a preserve where the plant is
+protected. They cannot search the entire jungle, so they take a simple random
+sample of quadrats and record the locations of the plants in the quadrats.
+They will then want to fit a model describing the trends in intensity of these
+plants extrapolated over the entire jungle. In this setting, the objective is
+to map the realized spatial intensity of the plant over this jungle so that
+these plants can be protected, not to estimate parameters of the process that
+arranges plants of this species at jungles like this one.
+
+Another situation where it is useful to extrapolate point pattern intensity
+outside of the observed region is in mapping subsurface geomagnetic anomalies,
+such as the munitions debris found at former military test ranges. This is
+frequently done in the early stages of an unexploded ordnance (UXO)
+remediation, where the intensity surface of inert munitions fragments is
+used to identify the locations of targets so that the search for UXO can be
+focused on the sections of the site most likely to contain it. The anomalies
+are only observed when detection equipment passes directly over them, but the
+project leaders are concerened with finding UXO and do not want to waste
+resources finding every inert fragment that could be found by metal detectors.
+The most common data collection method is to take a systematic sample of
+ straight-line transects and observe anomalies in rectangular regions centered
+along the transects. Frequently, moving averages of the intensity are computed
+in circular windows and an intensity map is produced using ordinary kriging to
+predict the intensity in a grid of these windows, but this assumes
+stationarity when the moving averages are believed to be non-stationary,
+ignores the point process nature of the data, and is very sensitive to the
+window size~\parencite{vspguide,wp}. In this project, I explore the use of
+polynomial trend surface models for the log-intensity at a simulated UXO site.
+
 \section{Surface Fitting}
 
-\textcite{baddeleycoeurjolly14,baddeleyturner,diggleetal,bermanturner}
+\textcite{digglebook,bermanturner}
 
-\section{Simple Example}
+\section{Simple Examples}
 
-\section{Simulation}
+log-linear example from HW4
+
+simulate from
+\begin{equation*}
+\log(\lambda(x,y)) = 5x + 2y; \quad (x, y) \in (0, 1)^{2}
+\end{equation*}
+but do not observe any events in \((0.5, 0.8)^{2}\). Then fit
+\begin{equation*}
+\log(\lambda(x,y)) = \beta_{0} + \beta_{1}x + \beta_{2}y
+\end{equation*}
+and predict on \((-0.5, 1.5)^{2}\).
+
+<<hw4_sim>>=
+simulate_win <- owin(c(0, 1), c(0, 1))
+observe_win <- setminus.owin(simulate_win, owin(c(0.5, 0.8), c(0.5, 0.8)))
+predict_win <- owin(c(-0.5, 1.5), c(-0.5, 1.5))
+hw4 <- rpoispp(function(x, y){exp(5 * x + 2 * y)}, win = simulate_win)
+@
+
+<<hw4_plot, fig.cap = 'One realization of a Poisson process with log-linear intensity on the unit square.'>>=
+par(mar = c(0, 0, 1, 0))
+plot(hw4, main = 'Event Locations')
+@
+
+<<hw4_fit_all>>=
+hw4_fit_a <- ppm(hw4 ~ x + y, correction = 'isotropic')
+hw4_pred_a <- predict(hw4_fit_a, window = predict_win, se = TRUE)
+# By default predictions are on a 128 by 128 grid.
+@
+<<hw4_fit_unobserved>>=
+hw4_fit_u <- ppm(hw4[observe_win] ~ x + y, correction = 'isotropic')
+hw4_pred_u <- predict(hw4_fit_u, window = predict_win, se = TRUE)
+@
+
+<<hw4_fitcoefs_a, results = 'asis'>>=
+hw4_coef_a <- summary(hw4_fit_a)$coefs.SE.CI[,c('Estimate', 'S.E.')]
+rownames(hw4_coef_a) <- c('\\(\\widehat{\\beta_{0}}\\)',
+                          '\\(\\widehat{\\beta_{1}}\\)',
+                          '\\(\\widehat{\\beta_{2}}\\)')
+xtable(hw4_coef_a, label = 'hw4esta',
+       caption = 'Estimated coefficients for the log-linear trend model fit using the full region.')
+@
+
+<<hw4_fitcoefs_u, results = 'asis'>>=
+hw4_coef_u <- summary(hw4_fit_u)$coefs.SE.CI[,c('Estimate', 'S.E.')]
+rownames(hw4_coef_u) <- c('\\(\\widehat{\\beta_{0}}\\)',
+                          '\\(\\widehat{\\beta_{1}}\\)',
+                          '\\(\\widehat{\\beta_{2}}\\)')
+xtable(hw4_coef_u, label = 'hw4estu',
+       caption = 'Estimated coefficients for the log-linear trend model fit using a subset of the region.')
+@
+
+<<hw4_fitplot, fig.pos = 'p', fig.height = 8, fig.cap = 'Estimated log-intensity surface, log-scale prediction standard errors, and prediction coeffiecient of variation from models fit using all events in the full simulation region (left) and events in a subset of the simulation region (right).
+The regions and event locations are overlaid in white.'>>=
+par(mfrow = c(3, 2), mar = c(0, 0, 1.2, 2), las = 2, cex = 1)
+
+plot(log(hw4_pred_a$estimate), zlim = c(-4.5, 10.5),
+     main = expression(log(hat(lambda)[f](italic(list(x,y))))))
+plot(simulate_win, add = TRUE, border = '#ffffff40')
+points(hw4, pch = '.', col = 'white')
+
+plot(log(hw4_pred_u$estimate), zlim = c(-4.5, 10.5),
+     main = expression(log(hat(lambda)[s](italic(list(x,y))))))
+plot(observe_win, add = TRUE, border = '#ffffff40')
+points(hw4[observe_win], pch = '.', col = 'white')
+
+plot(log(hw4_pred_a$se), zlim = c(-4.5, 10.5),
+     main = expression(log(SE(hat(lambda)[f](italic(list(x,y)))))))
+plot(simulate_win, add = TRUE, border = '#ffffff40')
+points(hw4, pch = '.', col = 'white')
+
+plot(log(hw4_pred_u$se), zlim = c(-4.5, 10.5),
+     main = expression(log(SE(hat(lambda)[s](italic(list(x,y)))))))
+plot(observe_win, add = TRUE, border = '#ffffff40')
+points(hw4[observe_win], pch = '.', col = 'white')
+
+plot(hw4_pred_a$se/hw4_pred_a$estimate, zlim = c(0, 1),
+     main = expression(SE(hat(lambda)[f](italic(list(x,y)))) /
+                         hat(lambda)[f](italic(list(x,y)))))
+plot(simulate_win, add = TRUE, border = '#ffffff40')
+points(hw4, pch = '.', col = 'white')
+
+plot(hw4_pred_u$se/hw4_pred_u$estimate, zlim = c(0, 1),
+     main = expression(SE(hat(lambda)[s](italic(list(x,y)))) /
+                         hat(lambda)[s](italic(list(x,y)))))
+plot(observe_win, add = TRUE, border = '#ffffff40')
+points(hw4[observe_win], pch = '.', col = 'white')
+@
+
+easy site, realization 2,000, 390 ft spacing
+
+use \texttt{polynom()} not \texttt{poly()}!
+
+\section{Simulation Study}
+
+\section{Discussion and Conclusion}
+
+smallish SEs outside observed region because of faith in the model form --
+need model checking
+
+how much of the region do you need to observe to trust your model?
+
+\appendix
+\section{R Code Appendix}
 

### 2017-03-15 project/paper.rnw
Theoretical background is about done.

@@
 
-One little-studied use of inhomogeneous point process intensity estimation
-is the use of an intensity surface model fit to events in a subregion to
-predict the intensity over the entire region of interest. This procedure would
-be relevant whenver it is known or suspected that some type of plant, animal,
-or other item occurs in the region, and the goal of the analysis is to map the
-trend in where these tend to be located rather than estimate parameters of
-some process at work across hypothetical replicates of similar regions. It may
-be prohibitively expensive or difficult to observe all events over the entire
-region, so a sample of subregions is taken. For example, conservationists want
-to study the spatial distribution of an endangered plant across a large region
-of thick jungle so that they can establish a preserve where the plant is
-protected. They cannot search the entire jungle, so they take a simple random
-sample of quadrats and record the locations of the plants in the quadrats.
-They will then want to fit a model describing the trends in intensity of these
-plants extrapolated over the entire jungle. In this setting, the objective is
-to map the realized spatial intensity of the plant over this jungle so that
-these plants can be protected, not to estimate parameters of the process that
-arranges plants of this species at jungles like this one.
+One little-studied application of inhomogeneous point process intensity
+estimation is the use of an intensity surface estimated from events in a
+subregion to predict the intensity over the entire region of interest. This
+procedure would be relevant whenver it is known or suspected that some type of
+plant, animal, or other item occurs in the region, and the goal of the
+analysis is to map the trend in where these tend to be located rather than
+estimate parameters of some process at work across hypothetical replicates of
+similar regions. It may be prohibitively expensive or difficult to observe all
+events over the entire region, so a sample of subregions is taken. For
+example, conservationists want to study the spatial distribution of an
+endangered plant across a large region of thick jungle so that they can
+establish a preserve where the plant is protected. They cannot search the
+entire jungle, so they take a simple random sample of quadrats and record the
+locations of the plants in those quadrats. They will then want to fit a model
+describing the trends in intensity of these plants extrapolated over the
+entire jungle. In this setting, the objective is to map the realized spatial
+intensity of the plant over this jungle so that these plants can be protected,
+not to estimate parameters of the process that arranges plants of this species
+at jungles like this one.
 
@@
 in circular windows and an intensity map is produced using ordinary kriging to
-predict the intensity in a grid of these windows, but this assumes
-stationarity when the moving averages are believed to be non-stationary,
-ignores the point process nature of the data, and is very sensitive to the
-window size~\parencite{vspguide,wp}. In this project, I explore the use of
-polynomial trend surface models for the log-intensity at a simulated UXO site.
+predict the intensity in a grid of these windows. There are several problems
+with this approach: it assumes stationarity when the moving averages are
+believed to be non-stationary, it ignores the point process nature of the
+data, and it is very sensitive to the window size~\parencite{vspguide,wp}. In
+this project, I explore the use of polynomial trend surface models for the
+log-intensity at a simulated UXO site.
 
@@
 
-\textcite{digglebook,bermanturner}
+It is theoretically possible to use maximum likelihood or maximum
+pseudo-likelihood methods to fit trend surface models or regression models
+(with covariates) for the intensity of an inhomogeneous point process, but
+these are not widely used because, in the most general cases, the problems
+are ``notoriously intractable''~\parencite{digglebook}. However, for spatial
+Poisson processes, trend surface models are tenable with the help of some
+numerical methods.
+
+The log-likelihood of an unmarked Poisson process on a region \(D\) with
+intensity \(\lambda(\mathbf{s})\) is found by conditioning on the number of
+events in the following manner. The number of events in \(D\) follows a
+Poisson distribution with mean \(\mu=\int_{D}\lambda(\mathbf{s})d\mathbf{s}\).
+Given that \(n\) events occured, their locations \(\mathbf{s}_{1},\dots,
+\mathbf{s}_{n}\) are independent and identically distributed with density
+\(\lambda(\mathbf{s})/\mu\). Then the log-likelihood of the intensity is
+\begin{align*}
+\ell(\lambda) &= \left(-\mu + n\log(\mu) - \log(n!)\right)
++ \left(\sum_{i=1}^{n}\left[\log\left(\lambda(\mathbf{s}_{i})\right)
+- \log(\mu)\right]\right) \\
+&= \sum_{i=1}^{n}\log\left(\lambda(\mathbf{s}_{i})\right)
+-\int_{D}\lambda(\mathbf{s})d\mathbf{s} - \log(n!).
+\end{align*}
+A natural way to model the intensity is to define a log-linear model
+\begin{equation*}
+\log(\lambda(\mathbf{s})) = \mathbf{x}(\mathbf{s})^{T}\boldsymbol{\beta}
+\end{equation*}
+where \(\mathbf{x}(\mathbf{s})\) can, in principle, include functions of the
+spatial coordinates and also covariates. Unfortunately, the likelihood becomes
+\begin{equation*}
+\sum_{i=1}^{n}\mathbf{x}(\mathbf{s})^{T}\boldsymbol{\beta}
+-\int_{D}\exp\left(\mathbf{x}(\mathbf{s})^{T}
+\boldsymbol{\beta}\right)d\mathbf{s} - \log(n!)
+\end{equation*}
+so \(\mathbf{x}(\mathbf{s})\) must be known (or predicted) either across the
+entire region \(D\) or at enough locations to approximate the integral
+numerically. Thus, covariates require some extra data collection or modeling
+effort to incorporate, but trend surfaces are feasible.
+
+\textcite{bermanturner} proposed a practical method for fitting these models
+with generalized linear models software using quadratic approximations to the
+likelihood, even generalizing to other link functions besides the log link.
+Their method is based on partitioning \(D\) such that each subset in the
+partition contains at most one event, and then using a binomial or Poission
+distribution for the count of events in each subset. This was followed up by
+\textcite{baddeleycoeurjolly14}, who developed a logistic regression-based
+approach that avoids the quadratic approximation, thereby reducing bias in the
+coefficient estimates. Both methods are available in the \texttt{spatstat}
+package~\parencite{spatstat} for R~\parencite{baser}. The \texttt{spatstat}
+function \texttt{ppm} implements these methods using the base R \texttt{glm}
+function to maximize the likelihood; I will use the Berman and Turner method
+because it is the default in \texttt{ppm} and therefore better reflects the
+experience of a na\"{i}ve practitioner who is not an expert on spatial point
+processes.
 
@@
 
+What to compare? MSPE for this method vs the moving average kriging? I'd
+like to cross-validate (could leave one transect out) but I don't know how to
+contruct a criterion. Maybe compare modeled intensity within the rectangle to
+a kernel intensity estimate?
+
 \section{Discussion and Conclusion}

### 2017-03-17 project/paper.rnw
Finished the toy example. The next task is to figure out the fun stuff.

@@
                show.signif.stars = FALSE,
-               fig.align = 'center', fig.width = 6.5, fig.height = 3,
+               fig.align = 'center', fig.width = 6.5, fig.height = 2.5,
                fig.pos = 'H', size = 'footnotesize', dev = 'pdf',
@@
         show.signif.stars = FALSE)
+@
 
+<<packages, message = FALSE, cache = FALSE>>=
 library(spatstat)
@@
 data, and it is very sensitive to the window size~\parencite{vspguide,wp}. In
-this project, I explore the use of polynomial trend surface models for the
+this project, I evaluate the use of polynomial trend surface models for the
 log-intensity at a simulated UXO site.
 
-\section{Surface Fitting}
+\section{Surface Fitting by Maximum Likelihood}
 
@@
 \begin{align*}
-\ell(\lambda) &= \left(-\mu + n\log(\mu) - \log(n!)\right)
-+ \left(\sum_{i=1}^{n}\left[\log\left(\lambda(\mathbf{s}_{i})\right)
-- \log(\mu)\right]\right) \\
+\ell(\lambda) &= \left\{-\mu + n\log(\mu) - \log(n!)\right\}
++ \sum_{i=1}^{n}\left\{\log\left(\lambda(\mathbf{s}_{i})\right)
+- \log(\mu)\right\} \\
 &= \sum_{i=1}^{n}\log\left(\lambda(\mathbf{s}_{i})\right)
@@
 \textcite{bermanturner} proposed a practical method for fitting these models
-with generalized linear models software using quadratic approximations to the
-likelihood, even generalizing to other link functions besides the log link.
+with generalized linear model (GLM) software using quadratic approximations to
+the likelihood, even generalizing to other link functions besides the log link.
 Their method is based on partitioning \(D\) such that each subset in the
-partition contains at most one event, and then using a binomial or Poission
-distribution for the count of events in each subset. This was followed up by
-\textcite{baddeleycoeurjolly14}, who developed a logistic regression-based
-approach that avoids the quadratic approximation, thereby reducing bias in the
-coefficient estimates. Both methods are available in the \texttt{spatstat}
-package~\parencite{spatstat} for R~\parencite{baser}. The \texttt{spatstat}
-function \texttt{ppm} implements these methods using the base R \texttt{glm}
-function to maximize the likelihood; I will use the Berman and Turner method
-because it is the default in \texttt{ppm} and therefore better reflects the
-experience of a na\"{i}ve practitioner who is not an expert on spatial point
-processes.
+partition contains at most one event, and then maximizing a weighted
+pseudolikelihood based upon a binomial or Poission distribution for the count
+of events in each subset. They comment that the covariance matrices produced
+by standard GLM software under this method are reasonable approximations to
+the correct covariance matrices that would be derived from the true likelihood
+as long as the quadrature approximation of the likelihood converges to the
+true likelihood as the resolution of the partition is increased.
+This was followed up by \textcite{baddeleycoeurjolly14}, who developed a
+logistic regression-based approach that avoids the quadratic approximation,
+thereby reducing bias in the coefficient estimates. Both methods are available
+in the \texttt{spatstat} package~\parencite{spatstat} for R~\parencite{baser}.
+The \texttt{spatstat} function \texttt{ppm} implements these methods using the
+base R \texttt{glm} function to maximize the likelihood; I will use the Berman
+and Turner method because it is the default in \texttt{ppm} and therefore
+better reflects the experience of a na\"{i}ve practitioner who is not an
+expert on spatial point processes.
 
-\section{Simple Examples}
+\section{Examples}
 
-log-linear example from HW4
+To illustrate model fitting and prediction, I present two examples before
+proceeding to the simulation study. The first is a toy example where I
+simulate from a Poisson process with a true log-linear intensity function and
+then estimate coefficients using the true model form. The second example
+introduces the my UXO site and provides a case study of fitting a polynomial
+trend surface to one realization of the the site.
 
-simulate from
+\subsection{True Log-Linear Intensity Surface}
+
+Previously this semester, we worked with an example of a Poisson process on
+the unit square \(\{(x,y):0<x<1,0<y<1\}\) with the log-linear intensity
+surface
 \begin{equation*}
-\log(\lambda(x,y)) = 5x + 2y; \quad (x, y) \in (0, 1)^{2}
+\log(\lambda(x,y)) = 5x + 2y.
 \end{equation*}
-but do not observe any events in \((0.5, 0.8)^{2}\). Then fit
+I will continue this example, fitting the model
 \begin{equation*}
@@
 \end{equation*}
-and predict on \((-0.5, 1.5)^{2}\).
+first using the process observed over the full region (the entire unit
+square), and then using the process observed in a subregion treating
+\(\{(x,y):0.6<x<0.9,0.6<y<0.9\}\) as unobserved. This unobserved section is a
+region of relatively high intensity, so if the estimation or prediction
+methods are sensitive to the observation window I expect omitting this
+section to have a large effect.
+
+I will use both estimated models to predict on
+\(\{(x,y):-0.5<x<1.5,-0.5<y<1.5\}\). Figure~\ref{fig:hw4_plot} shows the
+realization of this process that I will use for both model fits.
 
@@
 simulate_win <- owin(c(0, 1), c(0, 1))
-observe_win <- setminus.owin(simulate_win, owin(c(0.5, 0.8), c(0.5, 0.8)))
+omit_win <- owin(c(0.6, 0.9), c(0.6, 0.9))
+observe_win <- setminus.owin(simulate_win, omit_win)
 predict_win <- owin(c(-0.5, 1.5), c(-0.5, 1.5))
@@
 
-<<hw4_plot, fig.cap = 'One realization of a Poisson process with log-linear intensity on the unit square.'>>=
+<<hw4_plot, fig.pos = 'b!', fig.cap = 'One realization of a Poisson process with log-linear intensity on the unit square.
+The dashed square shows the outline of the ``unobserved" region; when fitting the model to the subregion, the points inside it will be discarded and it will not be considered in constructing the approximation to the likelihood.'>>=
 par(mar = c(0, 0, 1, 0))
 plot(hw4, main = 'Event Locations')
+plot(omit_win, lty = 2, add = TRUE)
 @
 
-<<hw4_fit_all>>=
-hw4_fit_a <- ppm(hw4 ~ x + y, correction = 'isotropic')
-hw4_pred_a <- predict(hw4_fit_a, window = predict_win, se = TRUE)
+For clarity, I denote the estimated models as
+\begin{equation*}
+\log\left(\widehat{\lambda}_{f}(x,y)\right) = \widehat{\beta}^{(f)}_{0}
++ \widehat{\beta}^{(f)}_{1}x + \widehat{\beta}^{(f)}_{2}y
+\end{equation*}
+for the model fit to the full region, and
+\begin{equation*}
+\log\left(\widehat{\lambda}_{s}(x,y)\right) = \widehat{\beta}^{(s)}_{0}
++ \widehat{\beta}^{(s)}_{1}x + \widehat{\beta}^{(s)}_{2}y
+\end{equation*}
+for the model fit to the subregion. I use \texttt{ppm} to fit the models
+with its default likelihood approximation and isotropic edge correction.
+The coefficitents and standard errors are similar for both estimated models,
+and the true values \(\beta_{0} = 0\), \(\beta_{1} = 5\), and
+\(\beta_{2} = 2\) all fall within one standard error of the
+estimates~(Tables~\ref{hw4estf} and \ref{hw4ests}).
+
+After estimating the models, I use \texttt{spatstat}'s \texttt{predict.ppm}
+method to predict the intensity surface on a \(128 \times 128\) lattice of
+locations in the region
+\(\{(x,y):-0.5<x<1.5,-0.5<y<1.5\}\)~(Figure~\ref{fig:hw4_fitplot}). As
+expected from the similarity of the coefficient estimates, the predicted
+log-intensity surfaces for both models show very similar linear trends~(top
+row of the figure). At a glance the prediction standard errors for both
+models appear similar to the predicted intensities, as seen in the middle
+row of the figure~(shown on a logarithmic scale for visibility). In a linear
+model setting we would expect the prediction standard error to increase with
+the distance from the observations, but no such trend is apparent here; even
+in the bottom left of the prediction region the standard errors decrease as
+the predicted intensity decreases. However, it is illuminating to plot the
+predictor's relative standard error (the ratio of the standard error to the
+predicted intensity, bottom row of the figure). This ratio is not constant,
+and in fact it is lowest where the highest intensity of events was observed.
+
+In summary, the estimation procedure does an adequate job of estimating the
+parameters of the true model for these data, and the standard error of the
+predicted intensity appears to be more strongly related to the distance from
+the observed events than to the distance from the observed region.
+
+<<hw4_fit_full>>=
+hw4_fit_f <- ppm(hw4 ~ x + y, correction = 'isotropic')
+hw4_pred_f <- predict(hw4_fit_f, window = predict_win, se = TRUE)
 # By default predictions are on a 128 by 128 grid.
 @
-<<hw4_fit_unobserved>>=
-hw4_fit_u <- ppm(hw4[observe_win] ~ x + y, correction = 'isotropic')
-hw4_pred_u <- predict(hw4_fit_u, window = predict_win, se = TRUE)
+
+<<hw4_fit_s>>=
+hw4_fit_s <- ppm(hw4[observe_win] ~ x + y, correction = 'isotropic')
+hw4_pred_s <- predict(hw4_fit_s, window = predict_win, se = TRUE)
 @
 
-<<hw4_fitcoefs_a, results = 'asis'>>=
-hw4_coef_a <- summary(hw4_fit_a)$coefs.SE.CI[,c('Estimate', 'S.E.')]
-rownames(hw4_coef_a) <- c('\\(\\widehat{\\beta_{0}}\\)',
-                          '\\(\\widehat{\\beta_{1}}\\)',
-                          '\\(\\widehat{\\beta_{2}}\\)')
-xtable(hw4_coef_a, label = 'hw4esta',
-       caption = 'Estimated coefficients for the log-linear trend model fit using the full region.')
+<<hw4_fitcoefs_f, results = 'asis'>>=
+hw4_coef_f <- summary(hw4_fit_f)$coefs.SE.CI[,c('Estimate', 'S.E.')]
+rownames(hw4_coef_f) <- c('\\(\\widehat{\\beta}^{(f)}_{0}\\)',
+                          '\\(\\widehat{\\beta}^{(f)}_{1}\\)',
+                          '\\(\\widehat{\\beta}^{(f)}_{2}\\)')
+print(xtable(hw4_coef_f, label = 'hw4estf',
+             caption = 'Estimated coefficients for the log-linear trend model fit using the full region.'),
+      table.placement = 't!')
 @
 
-<<hw4_fitcoefs_u, results = 'asis'>>=
-hw4_coef_u <- summary(hw4_fit_u)$coefs.SE.CI[,c('Estimate', 'S.E.')]
-rownames(hw4_coef_u) <- c('\\(\\widehat{\\beta_{0}}\\)',
-                          '\\(\\widehat{\\beta_{1}}\\)',
-                          '\\(\\widehat{\\beta_{2}}\\)')
-xtable(hw4_coef_u, label = 'hw4estu',
-       caption = 'Estimated coefficients for the log-linear trend model fit using a subset of the region.')
+<<hw4_fitcoefs_s, results = 'asis'>>=
+hw4_coef_s <- summary(hw4_fit_s)$coefs.SE.CI[,c('Estimate', 'S.E.')]
+rownames(hw4_coef_s) <- c('\\(\\widehat{\\beta}^{(s)}_{0}\\)',
+                          '\\(\\widehat{\\beta}^{(s)}_{1}\\)',
+                          '\\(\\widehat{\\beta}^{(s)}_{2}\\)')
+print(xtable(hw4_coef_s, label = 'hw4ests',
+             caption = 'Estimated coefficients for the log-linear trend model fit using the subregion.'),
+      table.placement = 't!')
 @
 
-<<hw4_fitplot, fig.pos = 'p', fig.height = 8, fig.cap = 'Estimated log-intensity surface, log-scale prediction standard errors, and prediction coeffiecient of variation from models fit using all events in the full simulation region (left) and events in a subset of the simulation region (right).
+<<hw4_fitplot, fig.pos = 'p', fig.height = 8, fig.cap = 'Estimated log-intensity surface, log-scale prediction standard errors, and relative standard error from models fit using all events in the full simulation region (left) and events in a subset of the simulation region (right).
 The regions and event locations are overlaid in white.'>>=
-par(mfrow = c(3, 2), mar = c(0, 0, 1.2, 2), las = 2, cex = 1)
+par(mfrow = c(3, 2), mar = c(1, 0, 1.2, 2), las = 2, cex = 1)
 
-plot(log(hw4_pred_a$estimate), zlim = c(-4.5, 10.5),
+plot(log(hw4_pred_f$estimate), zlim = c(-4.5, 10.5),
      main = expression(log(hat(lambda)[f](italic(list(x,y))))))
-plot(simulate_win, add = TRUE, border = '#ffffff40')
+plot(simulate_win, add = TRUE, border = '#ffffff80')
 points(hw4, pch = '.', col = 'white')
 
-plot(log(hw4_pred_u$estimate), zlim = c(-4.5, 10.5),
+plot(log(hw4_pred_s$estimate), zlim = c(-4.5, 10.5),
      main = expression(log(hat(lambda)[s](italic(list(x,y))))))
-plot(observe_win, add = TRUE, border = '#ffffff40')
+plot(observe_win, add = TRUE, border = '#ffffff80')
 points(hw4[observe_win], pch = '.', col = 'white')
 
-plot(log(hw4_pred_a$se), zlim = c(-4.5, 10.5),
+plot(log(hw4_pred_f$se), zlim = c(-4.5, 10.5),
      main = expression(log(SE(hat(lambda)[f](italic(list(x,y)))))))
-plot(simulate_win, add = TRUE, border = '#ffffff40')
+plot(simulate_win, add = TRUE, border = '#ffffff80')
 points(hw4, pch = '.', col = 'white')
 
-plot(log(hw4_pred_u$se), zlim = c(-4.5, 10.5),
+plot(log(hw4_pred_s$se), zlim = c(-4.5, 10.5),
      main = expression(log(SE(hat(lambda)[s](italic(list(x,y)))))))
-plot(observe_win, add = TRUE, border = '#ffffff40')
+plot(observe_win, add = TRUE, border = '#ffffff80')
 points(hw4[observe_win], pch = '.', col = 'white')
 
-plot(hw4_pred_a$se/hw4_pred_a$estimate, zlim = c(0, 1),
+plot(hw4_pred_f$se/hw4_pred_f$estimate, zlim = c(0, 1),
      main = expression(SE(hat(lambda)[f](italic(list(x,y)))) /
                          hat(lambda)[f](italic(list(x,y)))))
-plot(simulate_win, add = TRUE, border = '#ffffff40')
+plot(simulate_win, add = TRUE, border = '#ffffff80')
 points(hw4, pch = '.', col = 'white')
 
-plot(hw4_pred_u$se/hw4_pred_u$estimate, zlim = c(0, 1),
+plot(hw4_pred_s$se/hw4_pred_s$estimate, zlim = c(0, 1),
      main = expression(SE(hat(lambda)[s](italic(list(x,y)))) /
                          hat(lambda)[s](italic(list(x,y)))))
-plot(observe_win, add = TRUE, border = '#ffffff40')
+plot(observe_win, add = TRUE, border = '#ffffff80')
 points(hw4[observe_win], pch = '.', col = 'white')
 @
+
+\subsection{Simple UXO Site}
 
@@
 
+What geostatisticians and ordnance engineers will notice is that that SEs
+don't reflect added uncertainty about unobserved regions.
+Note kriging accounts for lack of model fit by autocorrelated errors --
+prediction SEs increase far from obs because of lack of info about local
+deviations from mean. For Poisson process, the events are independent given
+the intensity so local variation is not helpful for prediction. Thus we are
+heavily reliant on the correctness of the model form and the
+representativeness of the observed subregion, so model checking is crucial.
+If there is prior uncertainty about the form of the trend in a certain part of
+the site, that region must be observed!
+
 how much of the region do you need to observe to trust your model?

### 2017-03-18 project/paper.rnw
Described the UXO site.

@@
 <<packages, message = FALSE, cache = FALSE>>=
+library(parallel)
 library(spatstat)
 
-set.seed(83534)
+set.seed(c(83534, 251, -18691, 283, -2526, 782), kind = 'L\'Ecuyer-CMRG')
 @
@@
 
+<<hw4_sim>>=
+simulate_win <- owin(c(0, 1), c(0, 1))
+omit_win <- owin(c(0.6, 0.9), c(0.6, 0.9))
+observe_win <- setminus.owin(simulate_win, omit_win)
+predict_win <- owin(c(-0.5, 1.5), c(-0.5, 1.5))
+hw4 <- rpoispp(function(x, y){exp(5 * x + 2 * y)}, win = simulate_win)
+@
+
 Previously this semester, we worked with an example of a Poisson process on
-the unit square \(\{(x,y):0<x<1,0<y<1\}\) with the log-linear intensity
-surface
+the unit square \(\{(x,y):\Sexpr{min(simulate_win$xrange)} < x
+< \Sexpr{max(simulate_win$xrange)}, \Sexpr{min(simulate_win$yrange)} < y
+< \Sexpr{max(simulate_win$yrange)}\}\) with the log-linear intensity surface
 \begin{equation*}
@@
 square), and then using the process observed in a subregion treating
-\(\{(x,y):0.6<x<0.9,0.6<y<0.9\}\) as unobserved. This unobserved section is a
-region of relatively high intensity, so if the estimation or prediction
-methods are sensitive to the observation window I expect omitting this
-section to have a large effect.
+\(\{(x,y): \Sexpr{min(omit_win$xrange)} < x < \Sexpr{max(omit_win$xrange)},
+\Sexpr{min(omit_win$yrange)} < y < \Sexpr{max(omit_win$yrange)}\}\) as
+unobserved. This unobserved section is a region of relatively high intensity,
+so if the estimation or prediction methods are sensitive to the observation
+window I expect omitting this section to have a large effect.
 
 I will use both estimated models to predict on
-\(\{(x,y):-0.5<x<1.5,-0.5<y<1.5\}\). Figure~\ref{fig:hw4_plot} shows the
+\(\{(x,y): \Sexpr{min(predict_win$xrange)} < x
+< \Sexpr{max(predict_win$xrange)}, \Sexpr{min(predict_win$yrange)} < y
+< \Sexpr{max(predict_win$yrange)}\}\). Figure~\ref{fig:hw4_plot} shows the
 realization of this process that I will use for both model fits.
-
-<<hw4_sim>>=
-simulate_win <- owin(c(0, 1), c(0, 1))
-omit_win <- owin(c(0.6, 0.9), c(0.6, 0.9))
-observe_win <- setminus.owin(simulate_win, omit_win)
-predict_win <- owin(c(-0.5, 1.5), c(-0.5, 1.5))
-hw4 <- rpoispp(function(x, y){exp(5 * x + 2 * y)}, win = simulate_win)
-@
 
@@
 method to predict the intensity surface on a \(128 \times 128\) lattice of
-locations in the region
-\(\{(x,y):-0.5<x<1.5,-0.5<y<1.5\}\)~(Figure~\ref{fig:hw4_fitplot}). As
+locations in the region \(\{(x,y): \Sexpr{min(predict_win$xrange)} < x
+< \Sexpr{max(predict_win$xrange)}, \Sexpr{min(predict_win$yrange)} < y
+< \Sexpr{max(predict_win$yrange)}\}\)~(Figure~\ref{fig:hw4_fitplot}). As
 expected from the similarity of the coefficient estimates, the predicted
@@
 
-easy site, realization 2,000, 390 ft spacing
+<<uxoexampledata>>=
+# Path to my writing project root directory.
+vspuxo <- file.path('..', '..', 'vspuxo')
+
+# Load my functions for working with UXO data.
+source(file.path(vspuxo, 'rfns', 'data_functions.r'))
+source(file.path(vspuxo, 'rfns', 'sampling_functions.r'))
+source(file.path(vspuxo, 'rfns', 'spatial_functions.r'))
+
+# Define the window for the UXO site.
+sitewindow <- owin(poly = cbind(x = c(1564294, 1564495, 1556870, 1557126),
+                                y = c(535421, 541130, 541085, 535576)))
+@
+
+In my writing project, I evaluated ordinary kriging of moving average
+intensity values on a simulated tank training range. I will use the same site
+for this project. The site is a
+\Sexpr{sprintf('%.1f', area(sitewindow) / 43560)} acre quadrilateral region
+with homogeneous Poisson process producing background anomalies with an
+intensity of 100 anomalies per acre, and two regions of concentrated munitions
+use generated as Poisson processes with bivariate Gaussian intensity. The
+total intesnsity at the center of each concentrated munitions use region is
+300 anomalies per acre~(Figure~\ref{fig:truth}).
+
+<<truth, cache = TRUE, fig.width = 3.75, fig.pos = 't', fig.cap = 'True intensity at the simulated UXO site.'>>=
+x <- seq(1556870, 1564495, by = 20)
+y <- seq(535421, 541130, by = 20)
+intense.mat <- matrix(100, nrow = length(x), length(y))
+for(i in seq_along(x)){
+  for(j in seq_along(y)){
+    intense.mat[i, j] <- 100 + gauss.elliptic(x[i]+10, y[j]+10,
+        mu.x = 1558400, mu.y = 540000,
+        s.a = 800/(2*qnorm(0.995)), s.b = 1200/(2*qnorm(0.995)),
+        r = pi/6, maxrate = 200
+      ) + gauss.elliptic(x[i]+10, y[j]+10,
+        mu.x = 1562000, mu.y = 537000,
+        s.a = 2000/(2*qnorm(0.995)), s.b = 900/(2*qnorm(0.995)),
+        r = 0, maxrate = 200
+      )
+  }
+}
+intense.im <- im(t(intense.mat), x+10, y+10, unitname = c('foot', 'feet'))
+
+par(mar = c(0, 0, 1, 3))
+plot(intense.im, main = 'True Intensity Surface', zlim = c(100, 300), las = 2)
+plot(setminus.owin(owin(c(1556870, 1564495), c(535421, 541130)),
+                   sitewindow), col = 'white', add = TRUE)
+mtext('Anomalies per Acre', 4, line = 2)
+@
+
+easy site, realization 2,000, 390 ft spacing, 6 ft width
 

### 2017-03-19 project/paper.rnw
Almost ready to start fitting models.

@@
 
-easy site, realization 2,000, 390 ft spacing, 6 ft width
+I simulate data collection along parallel straight-line transects spaced 390
+feet apart using a detector with a footprint six feet wide. The location of
+the first transect was selected randomly, and all anomalies within three feet
+of a transect are observed~(Figure~\ref{fig:sim2000obs}).
+
+<<sim2000obs>>=
+# Load the sample data used in my writing project.
+ex_anom <- read.anomaly(file.path(vspuxo, 'datasets', 'easy', 'sample',
+                        'easy_sample_tTA2_p100_bg100_fg200_rep2000.anomaly'))
+ex_path <- read.anomaly(file.path(vspuxo, 'datasets', 'easy', 'sample',
+                        'easy_sample_tTA2_p100_bg100_fg200_rep2000.cog'))
+
+par(mar = c(0, 0, 1, 0))
+plot(sitewindow, main = 'Observed Geomagnetic Anomalies')
+points(y ~ x, data = ex_anom, pch = '.')
+@
 

### 2017-03-19 project/paper.rnw
The quadrature scheme is too complicated.

--- 2017-03-18 project/paper.rnw
@@
 <<packages, message = FALSE, cache = FALSE>>=
-library(parallel)
 library(spatstat)
 
-set.seed(c(83534, 251, -18691, 283, -2526, 782), kind = 'L\'Ecuyer-CMRG')
+set.seed(83534)
 @
@@
 # Define the window for the UXO site.
-sitewindow <- owin(poly = cbind(x = c(1564294, 1564495, 1556870, 1557126),
+site_window <- owin(poly = cbind(x = c(1564294, 1564495, 1556870, 1557126),
                                  y = c(535421, 541130, 541085, 535576)))
@@
 for this project. The site is a
-\Sexpr{sprintf('%.1f', area(sitewindow) / 43560)} acre quadrilateral region
+\Sexpr{sprintf('%.2f', area(site_window) / 43560)} acre quadrilateral region
 with homogeneous Poisson process producing background anomalies with an
@@
 
-<<truth, cache = TRUE, fig.width = 3.75, fig.pos = 't', fig.cap = 'True intensity at the simulated UXO site.'>>=
-x <- seq(1556870, 1564495, by = 20)
-y <- seq(535421, 541130, by = 20)
+<<truth, cache = TRUE, fig.width = 4, fig.pos = 't', fig.cap = 'True intensity at the simulated UXO site.'>>=
+# Create an image of the surface in 20 ft by 20 ft pixels.
+# The gauss.elliptic function from spatial_functions.r is a bivariate normal
+# intensity parameterized in terms of the center, major and minor axis scales,
+# and angle of the major axis.
+x <- seq(1556880, 1564495, by = 20)
+y <- seq(535431, 541130, by = 20)
 intense.mat <- matrix(100, nrow = length(x), length(y))
@@
   for(j in seq_along(y)){
-    intense.mat[i, j] <- 100 + gauss.elliptic(x[i]+10, y[j]+10,
+    intense.mat[i, j] <- 100 + gauss.elliptic(x[i], y[j],
         mu.x = 1558400, mu.y = 540000,
-        s.a = 800/(2*qnorm(0.995)), s.b = 1200/(2*qnorm(0.995)),
+        s.a = 800 / (2 * qnorm(0.995)), s.b = 1200 / (2 * qnorm(0.995)),
         r = pi/6, maxrate = 200
-      ) + gauss.elliptic(x[i]+10, y[j]+10,
+      ) + gauss.elliptic(x[i], y[j],
         mu.x = 1562000, mu.y = 537000,
-        s.a = 2000/(2*qnorm(0.995)), s.b = 900/(2*qnorm(0.995)),
+        s.a = 2000 / (2 * qnorm(0.995)), s.b = 900 / (2 * qnorm(0.995)),
         r = 0, maxrate = 200
@@
 }
-intense.im <- im(t(intense.mat), x+10, y+10, unitname = c('foot', 'feet'))
+intense.im <- im(t(intense.mat), x, y, unitname = c('foot', 'feet'))
 
 par(mar = c(0, 0, 1, 3))
-plot(intense.im, main = 'True Intensity Surface', zlim = c(100, 300), las = 2)
+plot(intense.im, main = 'True Intensity Surface',
+     border = NA, zlim = c(100, 300), las = 2)
 plot(setminus.owin(owin(c(1556870, 1564495), c(535421, 541130)),
-                   sitewindow), col = 'white', add = TRUE)
+     site_window), col = 'white', border = 'white', lwd = 2, add = TRUE)
 mtext('Anomalies per Acre', 4, line = 2)
@@
 
-easy site, realization 2,000, 390 ft spacing, 6 ft width
+<<sim2000obs, fig.pos = 't', fig.cap = 'Observed geomagnetic anomalies from one realization of the UXO site. The transects are not shown on this plot because they would obscure the points.'>>=
+# Load the sample data used in my writing project.
+# The anomaly file contains the locations of the observed anamolies and the
+# cog file contains the course-over-ground waypoints on the transects.
+ex_anom <- read.csv(file.path(vspuxo, 'datasets', 'easy', 'sample',
+                    'easy_sample_tTA2_p200_bg100_fg200_rep2000.anomaly'))
+ex_path <- read.csv(file.path(vspuxo, 'datasets', 'easy', 'sample',
+                    'easy_sample_tTA2_p200_bg100_fg200_rep2000.cog'))
+
+# Create a spatstat window and ppp object.
+ex_win <- intersect.owin(site_window,
+  do.call(union.owin, lapply(unique(ex_path$x), function(x){
+    return(owin(c(x - 3, x + 3), c(535421, 541130)))
+  })))
+ex_ppp <- ppp(ex_anom$x, ex_anom$y, window = ex_win)
+
+par(mar = c(0, 0, 1, 0))
+plot(ex_ppp, main = 'Observed Geomagnetic Anomalies', pch = '.', border = NA)
+plot(site_window, add = TRUE)
+@
+
+I simulate data collection along parallel straight-line transects spaced 396
+feet apart using a detector with a footprint six feet wide (so there is an
+unobserved region 390 feet wide between each pair of adjacent transects). The
+location of the first transect is selected randomly, and the locations of all
+anomalies within three feet of a transect are
+recorded~(Figure~\ref{fig:sim2000obs}). In total,
+\Sexpr{sprintf('%.1f', area(ex_win) / 43560)} acres are observed, covering
+\Sexpr{sprintf('%.1f\\%%', 100 * area(ex_win) / area(site_window))} of the
+qegion of interest.
 
 use \texttt{polynom()} not \texttt{poly()}!
+
+<<polyfit1, eval = FALSE>>=
+ex_fit4 <- ppm(ex_ppp ~ polynom(x, y, 4))
+@
 

### 2017-03-22 project/paper.rnw
Just needed to rescale!

@@
 
-\section{Examples}
+%\section{Examples}
 
@@
 
-\subsection{True Log-Linear Intensity Surface}
+\section{True Log-Linear Intensity Surface}
 
@@
 
-\subsection{Simple UXO Site}
+\section{Simple UXO Site}
 
@@
      border = NA, zlim = c(100, 300), las = 2)
-plot(setminus.owin(owin(c(1556870, 1564495), c(535421, 541130)),
+plot(setminus.owin(owin(c(1556870, 1564495),
+                        c(535421, 541130)),
      site_window), col = 'white', border = 'white', lwd = 2, add = TRUE)
@@
 
-<<polyfit1, eval = FALSE>>=
-ex_fit4 <- ppm(ex_ppp ~ polynom(x, y, 4))
+<<polyfit1, warning = FALSE, fig.height = 8, fig.pos = 'p', fig.cap = ''>>=
+# Create a quadrature scheme manually to ensure there are dummy points on
+# each transect.
+# TODO: Figure out weights.
+ex_Q <- quad(data = ex_ppp,
+             dummy = as.ppp(rbind(
+               data.frame(vertices(Window(ex_ppp))),
+               do.call(rbind, lapply(unique(ex_path$x), function(x){
+                 return(data.frame(
+                   x = x, y = seq(min(ex_path$y), max(ex_path$y),
+                                  length.out = 128)
+                 ))}))), W = Window(ex_ppp)))
+
+ex_fit2 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 2), correction = 'iso')
+ex_fit3 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 3), correction = 'iso')
+ex_fit4 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 4), correction = 'iso')
+ex_fit5 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 5), correction = 'iso')
+ex_fit6 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 6), correction = 'iso')
+ex_fit7 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 7), correction = 'iso')
+ex_fit8 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 8), correction = 'iso')
+ex_fit9 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 9), correction = 'iso')
+ex_fit10 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 10), correction = 'iso')
+ex_fit11 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 11), correction = 'iso')
+ex_fit12 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 12), correction = 'iso')
+
+par(mfrow = c(4, 3), mar = c(1, 0, 1.2, 2), las = 2, cex = 1)
+plot(predict(ex_fit2, window = site_window), main = '2nd Degree')
+plot(predict(ex_fit3, window = site_window), main = '3rd Degree')
+plot(predict(ex_fit4, window = site_window), main = '4th Degree')
+plot(predict(ex_fit5, window = site_window), main = '5th Degree')
+plot(predict(ex_fit6, window = site_window), main = '6th Degree')
+plot(predict(ex_fit7, window = site_window), main = '7th Degree')
+plot(predict(ex_fit8, window = site_window), main = '8th Degree')
+plot(predict(ex_fit9, window = site_window), main = '9th Degree')
+plot(predict(ex_fit10, window = site_window), main = '10th Degree')
+plot(predict(ex_fit11, window = site_window), main = '11th Degree')
+plot(predict(ex_fit12, window = site_window), main = '12th Degree')
+# TODO: Kernel density plot.
 @
-
-\section{Simulation Study}
-
-What to compare? MSPE for this method vs the moving average kriging? I'd
-like to cross-validate (could leave one transect out) but I don't know how to
-contruct a criterion. Maybe compare modeled intensity within the rectangle to
-a kernel intensity estimate?
 

### 2017-03-22 project/paper.rnw
Made progress on the presentation.

--- 2017-03-19 project/paper.rnw
@@
 
-\subsection{Simple UXO Site}
+\section{Simulated UXO Site}
 
 <<uxoexampledata>>=
-# Path to my writing project root directory.
-vspuxo <- file.path('..', '..', 'vspuxo')
-
-# Load my functions for working with UXO data.
-source(file.path(vspuxo, 'rfns', 'data_functions.r'))
-source(file.path(vspuxo, 'rfns', 'sampling_functions.r'))
-source(file.path(vspuxo, 'rfns', 'spatial_functions.r'))
-
 # Define the window for the UXO site.
 site_window <- owin(poly = cbind(x = c(1564294, 1564495, 1556870, 1557126),
-                                 y = c(535421, 541130, 541085, 535576)))
+                                 y = c(535421, 541130, 541085, 535576)),
+                    unitname = c('foot', 'feet'))
 @
@@
 <<truth, cache = TRUE, fig.width = 4, fig.pos = 't', fig.cap = 'True intensity at the simulated UXO site.'>>=
+# Bivariate normal intensity function that I originalally wrote for my
+# writing project, parameterized in terms of the major and minor axes scales
+# and the angle of the major axis from horizontal.
+#   a is horizontal axis, b is vertical axis, r is rotation angle
+gauss.elliptic <- function(x, y, mu.x = 0, mu.y = 0, s.a = 1, s.b = 1,
+                           r = 0, maxrate = 1){
+  rot <- zapsmall(matrix(c(cos(r), sin(r), -sin(r), cos(r)), nrow = 2))
+  ab <- diag(c(s.a^2, s.b^2))
+  sigma <- rot %*% ab %*% t(rot)
+  siginv <- solve(sigma)
+  mu <- matrix(c(mu.x, mu.y), nrow = 2)
+  mat <- matrix(rbind(x, y), nrow = 2)
+  return(maxrate * apply(mat, 2, function(vec){
+      exp(-t(vec - mu) %*% siginv %*% (vec - mu) / 2)
+    }))
+}
+
 # Create an image of the surface in 20 ft by 20 ft pixels.
-# The gauss.elliptic function from spatial_functions.r is a bivariate normal
-# intensity parameterized in terms of the center, major and minor axis scales,
-# and angle of the major axis.
 x <- seq(1556880, 1564495, by = 20)
@@
      border = NA, zlim = c(100, 300), las = 2)
-plot(setminus.owin(owin(c(1556870, 1564495), c(535421, 541130)),
+plot(setminus.owin(owin(c(1556870, 1564495), c(535421, 541130),
+                        unitname = c('foot', 'feet')),
      site_window), col = 'white', border = 'white', lwd = 2, add = TRUE)
@@
 # cog file contains the course-over-ground waypoints on the transects.
-ex_anom <- read.csv(file.path(vspuxo, 'datasets', 'easy', 'sample',
-                    'easy_sample_tTA2_p200_bg100_fg200_rep2000.anomaly'))
-ex_path <- read.csv(file.path(vspuxo, 'datasets', 'easy', 'sample',
-                    'easy_sample_tTA2_p200_bg100_fg200_rep2000.cog'))
+ex_anom <- read.csv('easy_sample_tTA2_p200_bg100_fg200_rep2000.anomaly')
+ex_path <- read.csv('easy_sample_tTA2_p200_bg100_fg200_rep2000.cog')
 
@@
   do.call(union.owin, lapply(unique(ex_path$x), function(x){
-    return(owin(c(x - 3, x + 3), c(535421, 541130)))
+    return(owin(c(x - 3, x + 3), c(535421, 541130),
+           unitname = c('foot', 'feet')))
   })))
 ex_ppp <- ppp(ex_anom$x, ex_anom$y, window = ex_win)
+
+# Create a ppp object the uses the full region as the window.
+ex_ppp_full <- ppp(ex_anom$x, ex_anom$y, window = site_window)
 
@@
 
-<<polyfit1, eval = FALSE>>=
-ex_fit4 <- ppm(ex_ppp ~ polynom(x, y, 4))
+<<polyfit1, warning = FALSE, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = ''>>=
+# Create a quadrature scheme manually to ensure there are dummy points on
+# each transect.
+# TODO: Figure out weights.
+ex_Q <- quad(data = ex_ppp,
+             dummy = as.ppp(rbind(
+               data.frame(vertices(Window(ex_ppp))),
+               do.call(rbind, lapply(unique(ex_path$x), function(x){
+                 return(data.frame(
+                   x = x, y = seq(min(ex_path$y), max(ex_path$y),
+                                  length.out = 128)
+                 ))}))), W = Window(ex_ppp)))
+
+ex_fit2 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
+                               scale(y, center = 535421, scale = 5709),
+                                     2), correction = 'iso')
+ex_fit3 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
+                               scale(y, center = 535421, scale = 5709),
+                                     3), correction = 'iso')
+ex_fit4 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
+                               scale(y, center = 535421, scale = 5709),
+                                     4), correction = 'iso')
+ex_fit5 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
+                               scale(y, center = 535421, scale = 5709),
+                                     5), correction = 'iso')
+ex_fit6 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
+                               scale(y, center = 535421, scale = 5709),
+                                     6), correction = 'iso')
+ex_fit7 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
+                               scale(y, center = 535421, scale = 5709),
+                                     7), correction = 'iso')
+ex_fit8 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
+                               scale(y, center = 535421, scale = 5709),
+                                     8), correction = 'iso')
+ex_fit9 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
+                               scale(y, center = 535421, scale = 5709),
+                                     9), correction = 'iso')
+ex_fit10 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
+                                scale(y, center = 535421, scale = 5709),
+                                      10), correction = 'iso')
+ex_fit11 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
+                                scale(y, center = 535421, scale = 5709),
+                                      11), correction = 'iso')
+ex_fit12 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
+                                scale(y, center = 535421, scale = 5709),
+                                      12), correction = 'iso')
+
+par(mfrow = c(4, 3), mar = c(1, 0, 1.2, 3), las = 2, cex = 1)
+
+# Multiply predictions by 43560 to convert from anomalies per square foot
+# to anomalies per acre.
+plot(43560 * predict(ex_fit2, window = site_window), main = '2nd Degree')
+plot(43560 * predict(ex_fit3, window = site_window), main = '3rd Degree')
+plot(43560 * predict(ex_fit4, window = site_window), main = '4th Degree')
+plot(43560 * predict(ex_fit5, window = site_window), main = '5th Degree')
+plot(43560 * predict(ex_fit6, window = site_window), main = '6th Degree')
+plot(43560 * predict(ex_fit7, window = site_window), main = '7th Degree')
+plot(43560 * predict(ex_fit8, window = site_window), main = '8th Degree')
+plot(43560 * predict(ex_fit9, window = site_window), main = '9th Degree')
+plot(43560 * predict(ex_fit10, window = site_window), main = '10th Degree')
+plot(43560 * predict(ex_fit11, window = site_window), main = '11th Degree')
+plot(43560 * predict(ex_fit12, window = site_window), main = '12th Degree')
+plot(43560 * density(ex_ppp_full) * area(ex_ppp_full) / ex_ppp_full$n,
+     main = 'Kernel Density')
 @
 
-\section{Simulation Study}
-
-What to compare? MSPE for this method vs the moving average kriging? I'd
-like to cross-validate (could leave one transect out) but I don't know how to
-contruct a criterion. Maybe compare modeled intensity within the rectangle to
-a kernel intensity estimate?
+\texttt{plot} doesn't work, and \texttt{predict} doesn't work on the original
+window. \texttt{predict} can't compute SEs.
 

### 2017-03-22 project/pres.rnw
Made progress on the presentation.

\documentclass[unknownkeysallowed]{beamer}
\usepackage{graphicx,epsfig,hyperref,booktabs,lmodern}

\usepackage[backend=bibtex,style=authoryear,citestyle=authoryear-comp]{biblatex}
\addbibresource{references.bib}

\beamertemplatenavigationsymbolsempty

\mode<presentation>
{
  \usetheme{bunsen}
  \setbeamercovered{transparent}
  \setbeamertemplate{items}[circle]
}

% set fonts
%\usepackage{fontspec}
%\setsansfont{Fontin Sans}
%\setbeamerfont{frametitle}{size=\LARGE,series=\bfseries}

% color definitions
\usepackage{color}
\definecolor{uipoppy}{RGB}{225, 64, 5}
\definecolor{uipaleblue}{RGB}{96,123,139}
\definecolor{uiblack}{RGB}{0, 0, 0}

% title slide definition
\title{Stat 534 Project: \\
Extrapolation from Poisson Process Intensity Surface Models}
\author{Kenny Flagg}
\institute[MSU]{Department of Mathematical Sciences \\
Montana State University}
\date{March 24, 2017}

\begin{document}

<<setup, echo = FALSE, message = FALSE, cache = FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, comment = NA, message = FALSE,
               fig.path = 'pres_figure/', cache.path = 'pres_cache/',
               show.signif.stars = FALSE,
               fig.align = 'center', fig.width = 6.5, fig.height = 2.5,
               fig.pos = 'H', size = 'footnotesize', dev = 'pdf',
               dev.args = list(pointsize = 10))
knit_theme$set('print')

library(xtable)
options(xtable.table.placement = 'H', width = 80, scipen = 2,
        xtable.sanitize.rownames.function = function(x){return(x)},
        show.signif.stars = FALSE)
@

<<packages, message = FALSE, cache = FALSE>>=
library(spatstat)

set.seed(83534)
@

\setbeamertemplate{background}
{\includegraphics[width=\paperwidth,height=\paperheight]{beamer_title_page1}}
\setbeamertemplate{footline}[default]

\begin{frame}
	\vspace*{.5cm}
	\begin{center}
		\textcolor{white}{\huge Stat 534 Project: \\
Extrapolation from Poisson Process Intensity Surface Models \\}
		\vspace{.5cm}
		\textcolor{white}{\Large Kenny Flagg}\\
	\end{center}
\end{frame}

% Set the background for the rest of the slides.
% Insert infoline
\setbeamertemplate{background}
 {\includegraphics[width=\paperwidth,height=\paperheight]{msubg2}}
%\setbeamertemplate{footline}[bunsentheme]

\begin{frame}
\frametitle{Introduction}

\begin{itemize}

\item Goals

\begin{itemize}

\item Estimate inhomogeneous intensity surface from events in a subregion

\item Infer intensity across entire region

\end{itemize}

\item Applications

\begin{itemize}

\item Mapping where endangered species are located

\item Mapping geomegnetic anomalies prior to an unexploded ordnace (UXO)
remediation

\end{itemize}

\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Maximum Likelihood Intensity Surface Fitting}

\begin{itemize}

\item General point processes

\begin{itemize}

\item The theory is not too complicated but the computation is very difficult

\end{itemize}

\item Poisson processes

\begin{itemize}

\item Doable with numerical methods

\item Log-likelihood of Poisson with intensity \(\lambda(\mathbf{s})\) on
region \(D\) (note typos in \textcite{digglebook})
\begin{align*}
\ell(\lambda) &= \left\{-\mu + n\log(\mu) - \log(n!)\right\}
+ \sum_{i=1}^{n}\left\{\log\left(\lambda(\mathbf{s}_{i})\right)
- \log(\mu)\right\} \\
&= \sum_{i=1}^{n}\log\left(\lambda(\mathbf{s}_{i})\right)
-\int_{D}\lambda(\mathbf{s})d\mathbf{s} - \log(n!).
\end{align*}
where \(\mu=\int_{D}\lambda(\mathbf{s})d\mathbf{s}\)
\end{itemize}

\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Poisson Process Log-Linear Model}

\begin{itemize}

\item Assuming events are independent (conditional on the intensity function),
\begin{equation*}
\log(\lambda(\mathbf{s})) = \mathbf{x}(\mathbf{s})^{T}\boldsymbol{\beta}
\end{equation*}
where \(\mathbf{x}(\mathbf{s})^{T}\) is a row of predictors at location
\(\mathbf{s}\)

\item Predictors can include covariates, but they must be known across the
whole region

\item \textcite{bermanturner} use dummy points and quadrature to set up an
approximation as a weighted Poisson regression

\item Their method is implemented in \texttt{spatstat}'s \texttt{ppm}, with
\texttt{glm} from base R or \texttt{gam} from \texttt{mgcv} as the back-end

\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Simple Example}

<<hw4_sim>>=
simulate_win <- owin(c(0, 1), c(0, 1))
omit_win <- owin(c(0.6, 0.9), c(0.6, 0.9))
observe_win <- setminus.owin(simulate_win, omit_win)
predict_win <- owin(c(-0.5, 1.5), c(-0.5, 1.5))
hw4 <- rpoispp(function(x, y){exp(5 * x + 2 * y)}, win = simulate_win)
hw4_fit_s <- ppm(hw4[observe_win] ~ x + y, correction = 'isotropic')
hw4_pred_s <- predict(hw4_fit_s, window = predict_win, se = TRUE)
@

\begin{itemize}

\item True model

\begin{itemize}

\item Poisson process on the unit square

\item \(\displaystyle \log(\lambda(x,y)) = 5x + 2y\)

\end{itemize}

\item But we don't observe
\(\Sexpr{min(omit_win$xrange)} < x < \Sexpr{max(omit_win$xrange)},
\Sexpr{min(omit_win$yrange)} < y < \Sexpr{max(omit_win$yrange)}\)

\end{itemize}

<<hw4_plot>>=
par(mar = c(0, 0, 1, 0))
plot(hw4, main = 'Event Locations')
plot(omit_win, lty = 2, add = TRUE)
@
\end{frame}

\begin{frame}
\frametitle{Estimated Model}

\begin{itemize}

\item Fit the model
\(\displaystyle \log(\lambda(x,y)) = \beta_{0} + \beta_{1}x + \beta_{2}y\)

\end{itemize}

\begin{minipage}{0.5\textwidth}
<<hw4_fitcoefs_s, results = 'asis'>>=
hw4_coef_s <- summary(hw4_fit_s)$coefs.SE.CI[,c('Estimate', 'S.E.')]
rownames(hw4_coef_s) <- c('\\(\\widehat{\\beta}_{0}\\)',
                          '\\(\\widehat{\\beta}_{1}\\)',
                          '\\(\\widehat{\\beta}_{2}\\)')
xtable(hw4_coef_s)
@
\end{minipage}\begin{minipage}{0.5\textwidth}
<<hw4_dummy_s, fig.width = 3.25>>=
par(mar = c(0, 0, 1, 0))
plot(hw4_fit_s$Q, main = 'Data and Dummy Points')
@
\end{minipage}

\end{frame}

\begin{frame}
\frametitle{Extrapolate the Surface}

\begin{itemize}

\item Use the \texttt{predict} method

\item Specify a new window
\(\Sexpr{min(predict_win$xrange)} < x < \Sexpr{max(predict_win$xrange)},
\Sexpr{min(predict_win$yrange)} < y < \Sexpr{max(predict_win$yrange)}\)

\end{itemize}

<<hw4_resultplot>>=
par(mfrow = c(1, 2), mar = c(0, 0, 3, 0), cex = 1, las = 2)
plot(log(hw4_pred_s$estimate), zlim = c(-4.5, 10.5),
     main = expression(log(hat(lambda)(italic(list(x,y))))))
plot(observe_win, add = TRUE, border = '#ffffff80')
points(hw4[observe_win], pch = '.', col = 'white')
plot(log(hw4_pred_s$se), zlim = c(-4.5, 10.5),
     main = expression(log(SE(hat(lambda)(italic(list(x,y)))))))
plot(observe_win, add = TRUE, border = '#ffffff80')
points(hw4[observe_win], pch = '.', col = 'white')
@
\end{frame}

\begin{frame}
\frametitle{Where is the Uncertainty?}

\begin{itemize}

\item Relative standard error is lowest where the highest intensity was
observed

\end{itemize}

<<hw4_rse>>=
par(mar = c(0, 0, 1.5, 0), las = 2)
plot(hw4_pred_s$se/hw4_pred_s$estimate, zlim = c(0, 1),
     main = expression(SE(hat(lambda)(italic(list(x,y)))) /
                         hat(lambda)(italic(list(x,y)))))
plot(observe_win, add = TRUE, border = '#ffffff80')
points(hw4[observe_win], pch = '.', col = 'white')
@

\end{frame}

\begin{frame}
\frametitle{Simple UXO Site}

<<uxoexampledata>>=
# Define the window for the UXO site.
site_window <- owin(poly = cbind(x = c(1564294, 1564495, 1556870, 1557126),
                                 y = c(535421, 541130, 541085, 535576)),
                    unitname = c('foot', 'feet'))
@

\begin{itemize}

\item \Sexpr{sprintf('%.2f', area(site_window) / 43560)} acre region (roughly
\Sexpr{prettyNum(c(-1, 1) %*% site_window$xrange, big.mark = ',')} ft by
\Sexpr{prettyNum(c(-1, 1) %*% site_window$yrange, big.mark = ',')} ft)

\item High density of geomagnetic anomalies around targets

\item Low density of background anomalies

\end{itemize}

<<truth, cache = TRUE>>=
gauss.elliptic <- function(x, y, mu.x = 0, mu.y = 0, s.a = 1, s.b = 1,
                           r = 0, maxrate = 1){
  rot <- zapsmall(matrix(c(cos(r), sin(r), -sin(r), cos(r)), nrow = 2))
  ab <- diag(c(s.a^2, s.b^2))
  sigma <- rot %*% ab %*% t(rot)
  siginv <- solve(sigma)
  mu <- matrix(c(mu.x, mu.y), nrow = 2)
  mat <- matrix(rbind(x, y), nrow = 2)
  return(maxrate * apply(mat, 2, function(vec){
      exp(-t(vec - mu) %*% siginv %*% (vec - mu) / 2)
    }))
}
x <- seq(1556880, 1564495, by = 20)
y <- seq(535431, 541130, by = 20)
intense.mat <- matrix(100, nrow = length(x), length(y))
for(i in seq_along(x)){
  for(j in seq_along(y)){
    intense.mat[i, j] <- 100 + gauss.elliptic(x[i], y[j],
        mu.x = 1558400, mu.y = 540000,
        s.a = 800 / (2 * qnorm(0.995)), s.b = 1200 / (2 * qnorm(0.995)),
        r = pi/6, maxrate = 200
      ) + gauss.elliptic(x[i], y[j],
        mu.x = 1562000, mu.y = 537000,
        s.a = 2000 / (2 * qnorm(0.995)), s.b = 900 / (2 * qnorm(0.995)),
        r = 0, maxrate = 200
      )
  }
}
intense.im <- im(t(intense.mat), x, y, unitname = c('foot', 'feet'))

par(mar = c(0, 0, 1, 3))
plot(intense.im, main = 'True Intensity Surface',
     border = NA, zlim = c(100, 300), las = 2)
plot(setminus.owin(owin(c(1556870, 1564495), c(535421, 541130),
                        unitname = c('foot', 'feet')),
     site_window), col = 'white', border = 'white', lwd = 2, add = TRUE)
mtext('Anomalies per Acre', 4, line = -1)
@

\end{frame}

\begin{frame}[allowframebreaks]
\frametitle{References}

\vspace{-0.5cm}{\tiny
\printbibliography
}
\end{frame}

\end{document}

### 2017-03-22 project/paper.rnw
Just need to make plots and wrap up for the presentation.

@@
 
-<<polyfit1, warning = FALSE, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = ''>>=
+<<polyfit1, cache = TRUE, warning = FALSE, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = ''>>=
 # Create a quadrature scheme manually to ensure there are dummy points on
@@
 
+# Get predictions and SEs.
+ex_p2 <- predict(ex_fit2, window = site_window, se = TRUE)
+ex_p3 <- predict(ex_fit3, window = site_window, se = TRUE)
+ex_p4 <- predict(ex_fit4, window = site_window, se = TRUE)
+ex_p5 <- predict(ex_fit5, window = site_window, se = TRUE)
+ex_p6 <- predict(ex_fit6, window = site_window, se = TRUE)
+ex_p7 <- predict(ex_fit7, window = site_window, se = TRUE)
+ex_p8 <- predict(ex_fit8, window = site_window, se = TRUE)
+ex_p9 <- predict(ex_fit9, window = site_window, se = TRUE)
+ex_p10 <- predict(ex_fit10, window = site_window)
+ex_p11 <- predict(ex_fit11, window = site_window)
+ex_p12 <- predict(ex_fit12, window = site_window)
+
 # Multiply predictions by 43560 to convert from anomalies per square foot
 # to anomalies per acre.
-plot(43560 * predict(ex_fit2, window = site_window), main = '2nd Degree')
-plot(43560 * predict(ex_fit3, window = site_window), main = '3rd Degree')
-plot(43560 * predict(ex_fit4, window = site_window), main = '4th Degree')
-plot(43560 * predict(ex_fit5, window = site_window), main = '5th Degree')
-plot(43560 * predict(ex_fit6, window = site_window), main = '6th Degree')
-plot(43560 * predict(ex_fit7, window = site_window), main = '7th Degree')
-plot(43560 * predict(ex_fit8, window = site_window), main = '8th Degree')
-plot(43560 * predict(ex_fit9, window = site_window), main = '9th Degree')
-plot(43560 * predict(ex_fit10, window = site_window), main = '10th Degree')
-plot(43560 * predict(ex_fit11, window = site_window), main = '11th Degree')
-plot(43560 * predict(ex_fit12, window = site_window), main = '12th Degree')
+plot(43560 * ex_p2$estimate, main = '2nd Degree')
+plot(43560 * ex_p3$estimate, main = '3rd Degree')
+plot(43560 * ex_p4$estimate, main = '4th Degree')
+plot(43560 * ex_p5$estimate, main = '5th Degree')
+plot(43560 * ex_p6$estimate, main = '6th Degree')
+plot(43560 * ex_p7$estimate, main = '7th Degree')
+plot(43560 * ex_p8$estimate, main = '8th Degree')
+plot(43560 * ex_p9$estimate, main = '9th Degree')
+plot(43560 * ex_p10, main = '10th Degree')
+plot(43560 * ex_p11, main = '11th Degree')
+plot(43560 * ex_p12, main = '12th Degree')
 plot(43560 * density(ex_ppp_full) * area(ex_ppp_full) / ex_ppp_full$n,
      main = 'Kernel Density')
+@
+
+<<polyse1, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = ''>>=
+par(mfrow = c(4, 3), mar = c(1, 0, 2, 3), las = 2, cex = 1)
+plot(log(43560 * ex_p2$se), main = '2nd Degree\nlog(SE)', zlim = c(6.5, 10.5))
+plot(log(43560 * ex_p3$se), main = '3rd Degree\nlog(SE)', zlim = c(6.5, 10.5))
+plot(log(43560 * ex_p4$se), main = '4th Degree\nlog(SE)', zlim = c(6.5, 10.5))
+plot(ex_p2$se / ex_p2$estimate, main = '2nd Degree\nRSE', zlim = c(0, 2))
+plot(ex_p3$se / ex_p3$estimate, main = '3rd Degree\nRSE', zlim = c(0, 2))
+plot(ex_p4$se / ex_p4$estimate, main = '4th Degree\nRSE', zlim = c(0, 2))
+
+plot(log(43560 * ex_p6$se), main = '6th Degree\nlog(SE)', zlim = c(6.5, 10.5))
+plot(log(43560 * ex_p8$se), main = '8th Degree\nlog(SE)', zlim = c(6.5, 10.5))
+plot(log(43560 * ex_p9$se), main = '9th Degree\nlog(SE)', zlim = c(6.5, 10.5))
+plot(ex_p6$se / ex_p6$estimate, main = '6th Degree\nRSE', zlim = c(0, 2))
+plot(ex_p8$se / ex_p8$estimate, main = '8th Degree\nRSE', zlim = c(0, 2))
+plot(ex_p9$se / ex_p9$estimate, main = '9th Degree\nRSE', zlim = c(0, 2))
 @

### 2017-03-22 project/pres.rnw
Just need to make plots and wrap up for the presentation.

@@
                     unitname = c('foot', 'feet'))
+ex_anom <- read.csv('easy_sample_tTA2_p200_bg100_fg200_rep2000.anomaly')
+ex_path <- read.csv('easy_sample_tTA2_p200_bg100_fg200_rep2000.cog')
+ex_win <- intersect.owin(site_window,
+  do.call(union.owin, lapply(unique(ex_path$x), function(x){
+    return(owin(c(x - 3, x + 3), c(535421, 541130),
+           unitname = c('foot', 'feet')))
+  })))
+ex_ppp <- ppp(ex_anom$x, ex_anom$y, window = ex_win)
+ex_ppp_full <- ppp(ex_anom$x, ex_anom$y, window = site_window)
 @
@@
 
+\begin{frame}
+\frametitle{Observed Events}
+
+\begin{itemize}
+
+\item Metal detectors record anomalies in six foot wide strips along parallel
+transects with 396 feet between centerlines
+
+\item Observed \Sexpr{sprintf('%.1f', area(ex_win) / 43560)} acres,
+\Sexpr{sprintf('%.1f\\%%', 100 * area(ex_win) / area(site_window))}
+of the site
+
+\end{itemize}
+
+<<sim2000obs>>=
+par(mar = c(0, 0, 1, 0))
+plot(ex_ppp, main = 'Observed Geomagnetic Anomalies', pch = '.', border = NA)
+plot(site_window, add = TRUE)
+@
+\end{frame}
+
+\begin{frame}
+\frametitle{Models}
+
+\begin{itemize}
+
+\item Polynomial models
+\(\displaystyle\log(\lambda(x,y))
+= \sum_{i=0}^{p}\sum_{j=0}^{p-i}\beta_{ij}x^{i}y^{j}\)
+for \(p = 2, 3, \dots, 12\)
+
+\end{itemize}
+
+<<dummy, warning = FALSE>>=
+ex_fit1 <- ppm(ex_ppp_full ~ x + y)
+ex_Q <- quad(data = ex_ppp,
+             dummy = as.ppp(rbind(
+               data.frame(vertices(Window(ex_ppp))),
+               do.call(rbind, lapply(unique(ex_path$x), function(x){
+                 return(data.frame(
+                   x = x, y = seq(min(ex_path$y), max(ex_path$y),
+                                  length.out = 128)
+                 ))}))), W = Window(ex_ppp)))
+
+par(mfrow = c(1, 2), mar = c(0, 0, 1, 0), cex = 1)
+plot(ex_fit1$Q$dummy, pch = '.', border = NA,
+     main = 'Default Dummy Points')
+plot(site_window, add = TRUE)
+plot(ex_Q$dummy, pch = '.', border = NA,
+     main = 'Manual Dummy Points')
+plot(site_window, add = TRUE)
+@
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Problems with Implementation}
+
+\begin{itemize}
+
+\item Needed to rescale coordinates
+
+\item For \(p \geq 10\), \texttt{pmm} cannot compute SEs because the
+``Fisher information matrix is singular''
+
+\item The \texttt{plot} method gives an error about infinite values
+
+\item When the window isn't specified, the \texttt{predict} method's default
+grid misses all but two transects
+
+\item The predict method does not work with spline smoothers
+
+\item The magnitudes of the predictions are much to large
+
+\end{itemize}
+
+\end{frame}
+
 \begin{frame}[allowframebreaks]

### 2017-03-23 project/paper.rnw
Done with the presentation.

--- 2017-03-22 project/paper.rnw
@@
 
-ex_fit2 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
-                               scale(y, center = 535421, scale = 5709),
-                                     2), correction = 'iso')
-ex_fit3 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
-                               scale(y, center = 535421, scale = 5709),
-                                     3), correction = 'iso')
-ex_fit4 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
-                               scale(y, center = 535421, scale = 5709),
-                                     4), correction = 'iso')
-ex_fit5 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
-                               scale(y, center = 535421, scale = 5709),
-                                     5), correction = 'iso')
-ex_fit6 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
-                               scale(y, center = 535421, scale = 5709),
-                                     6), correction = 'iso')
-ex_fit7 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
-                               scale(y, center = 535421, scale = 5709),
-                                     7), correction = 'iso')
-ex_fit8 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
-                               scale(y, center = 535421, scale = 5709),
-                                     8), correction = 'iso')
-ex_fit9 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
-                               scale(y, center = 535421, scale = 5709),
-                                     9), correction = 'iso')
-ex_fit10 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
-                                scale(y, center = 535421, scale = 5709),
-                                      10), correction = 'iso')
-ex_fit11 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
-                                scale(y, center = 535421, scale = 5709),
-                                      11), correction = 'iso')
-ex_fit12 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
-                                scale(y, center = 535421, scale = 5709),
-                                      12), correction = 'iso')
+ex_fit2 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 2), correction = 'iso')
+ex_fit3 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 3), correction = 'iso')
+ex_fit4 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 4), correction = 'iso')
+ex_fit5 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 5), correction = 'iso')
+ex_fit6 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 6), correction = 'iso')
+ex_fit7 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 7), correction = 'iso')
+ex_fit8 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 8), correction = 'iso')
+ex_fit9 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 9), correction = 'iso')
+ex_fit10 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 10), correction = 'iso')
+ex_fit11 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 11), correction = 'iso')
+ex_fit12 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 12), correction = 'iso')
+
+# Get predictions and SEs.
+ex_p2 <- predict(ex_fit2, window = site_window, se = TRUE)
+ex_p3 <- predict(ex_fit3, window = site_window, se = TRUE)
+ex_p4 <- predict(ex_fit4, window = site_window, se = TRUE)
+ex_p5 <- predict(ex_fit5, window = site_window, se = TRUE)
+ex_p6 <- predict(ex_fit6, window = site_window, se = TRUE)
+ex_p7 <- predict(ex_fit7, window = site_window, se = TRUE)
+ex_p8 <- predict(ex_fit8, window = site_window, se = TRUE)
+ex_p9 <- predict(ex_fit9, window = site_window, se = TRUE)
+ex_p10 <- predict(ex_fit10, window = site_window, se = TRUE)
+ex_p11 <- predict(ex_fit11, window = site_window, se = TRUE)
+ex_p12 <- predict(ex_fit12, window = site_window, se = TRUE)
 
@@
 # to anomalies per acre.
-plot(43560 * predict(ex_fit2, window = site_window), main = '2nd Degree')
-plot(43560 * predict(ex_fit3, window = site_window), main = '3rd Degree')
-plot(43560 * predict(ex_fit4, window = site_window), main = '4th Degree')
-plot(43560 * predict(ex_fit5, window = site_window), main = '5th Degree')
-plot(43560 * predict(ex_fit6, window = site_window), main = '6th Degree')
-plot(43560 * predict(ex_fit7, window = site_window), main = '7th Degree')
-plot(43560 * predict(ex_fit8, window = site_window), main = '8th Degree')
-plot(43560 * predict(ex_fit9, window = site_window), main = '9th Degree')
-plot(43560 * predict(ex_fit10, window = site_window), main = '10th Degree')
-plot(43560 * predict(ex_fit11, window = site_window), main = '11th Degree')
-plot(43560 * predict(ex_fit12, window = site_window), main = '12th Degree')
+plot(43560 * ex_p2$estimate, main = '2nd Degree')
+plot(43560 * ex_p3$estimate, main = '3rd Degree')
+plot(43560 * ex_p4$estimate, main = '4th Degree')
+plot(43560 * ex_p5$estimate, main = '5th Degree')
+plot(43560 * ex_p6$estimate, main = '6th Degree')
+plot(43560 * ex_p7$estimate, main = '7th Degree')
+plot(43560 * ex_p8$estimate, main = '8th Degree')
+plot(43560 * ex_p9$estimate, main = '9th Degree')
+plot(43560 * ex_p10$estimate, main = '10th Degree')
+plot(43560 * ex_p11$estimate, main = '11th Degree')
+plot(43560 * ex_p12$estimate, main = '12th Degree')
 plot(43560 * density(ex_ppp_full) * area(ex_ppp_full) / ex_ppp_full$n,
@@
 
+<<polyse1, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = ''>>=
+par(mfrow = c(4, 3), mar = c(1, 0, 2, 3), las = 2, cex = 1)
+plot(log(43560 * ex_p2$se), main = '2nd Degree\nlog(SE)')
+plot(log(43560 * ex_p4$se), main = '4th Degree\nlog(SE)')
+plot(log(43560 * ex_p6$se), main = '6th Degree\nlog(SE)')
+plot(ex_p2$se / ex_p2$estimate, main = '2nd Degree\nRSE')
+plot(ex_p4$se / ex_p4$estimate, main = '4th Degree\nRSE')
+plot(ex_p6$se / ex_p6$estimate, main = '6th Degree\nRSE')
+
+plot(log(43560 * ex_p8$se), main = '8th Degree\nlog(SE)')
+plot(log(43560 * ex_p10$se), main = '10th Degree\nlog(SE)')
+plot(log(43560 * ex_p12$se), main = '12th Degree\nlog(SE)')
+plot(ex_p8$se / ex_p8$estimate, main = '8th Degree\nRSE')
+plot(ex_p10$se / ex_p10$estimate, main = '10th Degree\nRSE')
+plot(ex_p12$se / ex_p12$estimate, main = '12th Degree\nRSE')
+@
+
 \texttt{plot} doesn't work, and \texttt{predict} doesn't work on the original
-window. \texttt{predict} can't compute SEs.
+window
 

### 2017-03-23 project/pres.rnw
Done with the presentation.

--- 2017-03-22 project/pres.rnw
@@
 
-\item Mapping geomegnetic anomalies prior to an unexploded ordnace (UXO)
+\item Mapping geomagnetic anomalies prior to an unexploded ordnance (UXO)
 remediation
@@
 \item Their method is implemented in \texttt{spatstat}'s \texttt{ppm}, with
-\texttt{glm} from base R or \texttt{gam} from \texttt{mgcv} as the back-end
+\texttt{glm} from base R or \texttt{gam} from \texttt{mgcv} as the back-end,
+using the \texttt{quasi} family
 
@@
 <<uxoexampledata>>=
-# Define the window for the UXO site.
 site_window <- owin(poly = cbind(x = c(1564294, 1564495, 1556870, 1557126),
@@
 
+\begin{frame}
+\frametitle{Observed Events}
+
+\begin{itemize}
+
+\item Metal detectors record anomalies in six foot wide strips along parallel
+transects with 396 feet between centerlines
+
+\item Observed \Sexpr{sprintf('%.1f', area(ex_win) / 43560)} acres,
+\Sexpr{sprintf('%.1f\\%%', 100 * area(ex_win) / area(site_window))}
+of the site
+
+\end{itemize}
+
+<<sim2000obs>>=
+par(mar = c(0, 0, 1, 0))
+plot(ex_ppp, main = 'Observed Geomagnetic Anomalies', pch = '.', border = NA)
+plot(site_window, add = TRUE)
+@
+\end{frame}
+
+\begin{frame}
+\frametitle{Trend Surface Models}
+
+\begin{itemize}
+
+\item Polynomial models
+\begin{equation*}
+\log(\lambda(x,y)) = \sum_{i=0}^{p}\sum_{j=0}^{p-i}\beta_{ij}x^{i}y^{j};
+\qquad p = 2, 3, \dots, 12
+\end{equation*}
+
+\item Can approximate complicated surfaces
+
+\item Expect two peaks, so even \(p \geq 4\) could work well
+
+\item Rescaling \(x\) and \(y\) to mean 0 and variance 1 reduces numerical
+instability for large \(p\)
+
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Extrapolated Surfaces}
+
+<<polyfit, warning = FALSE, cache = TRUE>>=
+ex_fit1 <- ppm(ex_ppp ~ x + y)
+ex_Q <- quad(data = ex_ppp,
+             dummy = as.ppp(rbind(
+               data.frame(vertices(Window(ex_ppp))),
+               do.call(rbind, lapply(unique(ex_path$x), function(x){
+                 return(data.frame(
+                   x = x, y = seq(min(ex_path$y), max(ex_path$y),
+                                  length.out = 128)
+                 ))}))), W = Window(ex_ppp)))
+
+ex_fit2 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 2), correction = 'iso')
+ex_fit3 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 3), correction = 'iso')
+ex_fit4 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 4), correction = 'iso')
+ex_fit5 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 5), correction = 'iso')
+ex_fit6 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 6), correction = 'iso')
+ex_fit7 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 7), correction = 'iso')
+ex_fit8 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 8), correction = 'iso')
+ex_fit9 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 9), correction = 'iso')
+ex_fit10 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 10), correction = 'iso')
+ex_fit11 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 11), correction = 'iso')
+ex_fit12 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 12), correction = 'iso')
+
+ex_p2 <- predict(ex_fit2, window = site_window, se = TRUE)
+ex_p3 <- predict(ex_fit3, window = site_window, se = TRUE)
+ex_p4 <- predict(ex_fit4, window = site_window, se = TRUE)
+ex_p5 <- predict(ex_fit5, window = site_window, se = TRUE)
+ex_p6 <- predict(ex_fit6, window = site_window, se = TRUE)
+ex_p7 <- predict(ex_fit7, window = site_window, se = TRUE)
+ex_p8 <- predict(ex_fit8, window = site_window, se = TRUE)
+ex_p9 <- predict(ex_fit9, window = site_window, se = TRUE)
+ex_p10 <- predict(ex_fit10, window = site_window, se = TRUE)
+ex_p11 <- predict(ex_fit11, window = site_window, se = TRUE)
+ex_p12 <- predict(ex_fit12, window = site_window, se = TRUE)
+@
+
+<<resplots1>>=
+par(mfrow = c(2, 3), mar = c(1, 0, 1.2, 3), las = 2, cex = 1)
+plot(43560 * ex_p2$estimate, main = '2nd Degree Polynom')
+plot(43560 * ex_p3$estimate, main = '3rd Degree Polynom')
+plot(43560 * ex_p4$estimate, main = '4th Degree Polynom')
+plot(43560 * ex_p5$estimate, main = '5th Degree Polynom')
+plot(43560 * ex_p6$estimate, main = '6th Degree Polynom')
+plot(43560 * ex_p7$estimate, main = '7th Degree Polynom')
+@
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Extrapolated Surfaces}
+
+<<resplots2>>=
+par(mfrow = c(2, 3), mar = c(1, 0, 1.2, 3), las = 2, cex = 1)
+plot(43560 * ex_p8$estimate, main = '8th Degree Polynom')
+plot(43560 * ex_p9$estimate, main = '9th Degree Polynom')
+plot(43560 * ex_p10$estimate, main = '10th Degree Polynom')
+plot(43560 * ex_p11$estimate, main = '11th Degree Polynom')
+plot(43560 * ex_p12$estimate, main = '12th Degree Polynom')
+plot(43560 * density(ex_ppp_full) * area(ex_ppp_full) / area(ex_ppp),
+     main = 'Kernel Density')
+@
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Standard Errors}
+
+<<seplots>>=
+par(mfrow = c(2, 3), mar = c(1, 0, 1.2, 3), las = 2, cex = 1)
+plot(log(43560 * ex_p2$se), main = '2nd Degree log(SE)')
+plot(log(43560 * ex_p4$se), main = '4th Degree log(SE)')
+plot(log(43560 * ex_p6$se), main = '6th Degree log(SE)')
+plot(log(43560 * ex_p8$se), main = '8th Degree log(SE)')
+plot(log(43560 * ex_p10$se), main = '10th Degree log(SE)')
+plot(log(43560 * ex_p12$se), main = '12th Degree log(SE)')
+@
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Problems with Implementation}
+
+\begin{itemize}
+
+\item By default, \texttt{ppm} places dummy points on a grid across a bounding
+box
+
+\begin{itemize}
+
+\item Dummy points outside the actual window are discarded
+
+\item Only \Sexpr{prettyNum(ex_fit1$Q$dummy$n, big.mark = ',')} points on two
+transects are kept
+
+\end{itemize}
+
+\item I place 128 evenly-spaced dummy points along each transect
+
+\begin{itemize}
+
+\item \Sexpr{prettyNum(ex_Q$dummy$n, big.mark = ',')} are used
+
+\end{itemize}
+
+\end{itemize}
+
+<<dummy>>=
+par(mfrow = c(1, 2), mar = c(1, 0, 1, 0), cex = 1)
+plot(ex_fit1$Q$dummy, pch = '.', border = NA,
+     main = 'Default Dummy Points')
+plot(site_window, add = TRUE)
+plot(ex_Q$dummy, pch = '.', border = NA,
+     main = 'Manual Dummy Points')
+plot(site_window, add = TRUE)
+@
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Problems with Implementation}
+
+\begin{itemize}
+
+\item For \(p \geq 18\), \texttt{ppm} cannot compute SEs because the
+``Fisher information matrix is singular'' --- 190 coefficients
+
+\item The \texttt{plot} method gives an error about infinite values
+
+\item When the window isn't specified, the \texttt{predict} method's default
+grid misses all but two transects
+
+\item The predict method does not work with spline smoothers
+
+\item The magnitudes of the predictions are much too large
+
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Conclusion}
+
+\begin{itemize}
+
+\item Polynomial surfaces are flexible but require much faith in the model
+
+\item How to check these models?
+
+\item How much of the region must be observed?
+
+\item Can implement with existing R packages but not easily
+
+\item What scale are the prediction SEs on?
+
+\end{itemize}
+
+\end{frame}
+
 \begin{frame}[allowframebreaks]

### 2017-03-23 project/pres.rnw
Getting into heavy-duty computation.

@@
 ex_fit1 <- ppm(ex_ppp ~ x + y)
-ex_Q <- quad(data = ex_ppp,
-             dummy = as.ppp(rbind(
-               data.frame(vertices(Window(ex_ppp))),
-               do.call(rbind, lapply(unique(ex_path$x), function(x){
-                 return(data.frame(
-                   x = x, y = seq(min(ex_path$y), max(ex_path$y),
-                                  length.out = 128)
-                 ))}))), W = Window(ex_ppp)))
+ex_Q <- quad(data = ex_ppp, dummy = runifpoint(500000, Window(ex_ppp)))
 
@@
 
+\vspace{-11pt}
 \begin{itemize}
@@
 
-\item Dummy points outside the actual window are discarded
-
 \item Only \Sexpr{prettyNum(ex_fit1$Q$dummy$n, big.mark = ',')} points on two
@@
 
-\item I place 128 evenly-spaced dummy points along each transect
+\item I draw \Sexpr{prettyNum(ex_Q$dummy$n, big.mark = ',')} points from a
+uniform distribution on the observed region
 
-\begin{itemize}
-
-\item \Sexpr{prettyNum(ex_Q$dummy$n, big.mark = ',')} are used
-
-\end{itemize}
+\item \textcite{wartonshepherd} recommend using enough dummy points that the
+maximized log-likelihood converges
 

### 2017-03-24 project/paper.rnw
Now the presentation is really done.

@@
 
-<<uxoexampledata>>=
+<<uxoexampledata, cache = TRUE>>=
 # Define the window for the UXO site.
@@
 
-<<truth, cache = TRUE, fig.width = 4, fig.pos = 't', fig.cap = 'True intensity at the simulated UXO site.'>>=
+<<truth, cache = TRUE, dependson = 'uxoexampledata', fig.width = 4, fig.pos = 't', fig.cap = 'True intensity at the simulated UXO site.'>>=
 # Bivariate normal intensity function that I originalally wrote for my
@@
 
-<<sim2000obs, fig.pos = 't', fig.cap = 'Observed geomagnetic anomalies from one realization of the UXO site. The transects are not shown on this plot because they would obscure the points.'>>=
+<<sim2000obs, cache = TRUE, dependson = 'uxoexampledata', fig.pos = 't', fig.cap = 'Observed geomagnetic anomalies from one realization of the UXO site.
+The transects are not shown on this plot because they would obscure the points.'>>=
 # Load the sample data used in my writing project.
@@
 
-<<polyfit1, cache = TRUE, warning = FALSE, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = ''>>=
+<<polyfit1, cache = TRUE, dependson = 'sim2000obs', warning = FALSE, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = ''>>=
+# Do a simple fit where spatstat sets up the quadrature so we can see where it
+# puts the dummy points.
+ex_fit1 <- ppm(ex_ppp ~ x + y)
+
 # Create a quadrature scheme manually to ensure there are dummy points on
@@
 # TODO: Figure out weights.
-ex_Q <- quad(data = ex_ppp,
-             dummy = as.ppp(rbind(
-               data.frame(vertices(Window(ex_ppp))),
-               do.call(rbind, lapply(unique(ex_path$x), function(x){
-                 return(data.frame(
-                   x = x, y = seq(min(ex_path$y), max(ex_path$y),
-                                  length.out = 128)
-                 ))}))), W = Window(ex_ppp)))
+ex_Q <- quad(data = ex_ppp, dummy = runifpoint(500000, Window(ex_ppp)))
 
@@
 plot(43560 * ex_p12$estimate, main = '12th Degree')
-plot(43560 * density(ex_ppp_full) * area(ex_ppp_full) / ex_ppp_full$n,
+plot(43560 * density(ex_ppp_full) * area(ex_ppp_full) / area(ex_ppp),
      main = 'Kernel Density')
@@
 
-<<polyse1, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = ''>>=
+<<polyse1, cache = TRUE, dependson = 'polyfit1', fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = ''>>=
 par(mfrow = c(4, 3), mar = c(1, 0, 2, 3), las = 2, cex = 1)

### 2017-03-24 project/pres.rnw
Now the presentation is really done.

@@
 
-<<uxoexampledata>>=
+<<uxoexampledata, cache = TRUE>>=
 site_window <- owin(poly = cbind(x = c(1564294, 1564495, 1556870, 1557126),
@@
 
-<<truth, cache = TRUE>>=
+<<truth, cache = TRUE, dependson = 'uxoexampledata'>>=
 gauss.elliptic <- function(x, y, mu.x = 0, mu.y = 0, s.a = 1, s.b = 1,
@@
 
-<<polyfit, warning = FALSE, cache = TRUE>>=
+<<polyfit, warning = FALSE, cache = TRUE, dependson = 'uxoexampledata'>>=
 ex_fit1 <- ppm(ex_ppp ~ x + y)
@@
 
-<<resplots1>>=
+<<resplots1, cache = TRUE, dependson = 'polyfit'>>=
 par(mfrow = c(2, 3), mar = c(1, 0, 1.2, 3), las = 2, cex = 1)
@@
 
-<<resplots2>>=
+<<resplots2, cache = TRUE, dependson = 'polyfit'>>=
 par(mfrow = c(2, 3), mar = c(1, 0, 1.2, 3), las = 2, cex = 1)
@@
 
-<<seplots>>=
+<<seplots, cache = TRUE, dependson = 'polyfit'>>=
 par(mfrow = c(2, 3), mar = c(1, 0, 1.2, 3), las = 2, cex = 1)
@@
 
-<<dummy>>=
+<<dummy, cache = TRUE, dependson = 'polyfit'>>=
 par(mfrow = c(1, 2), mar = c(1, 0, 1, 0), cex = 1)
@@
 
-\item The magnitudes of the predictions are much too large
-
 \end{itemize}
@@
 
-\item What scale are the prediction SEs on?
-
 \end{itemize}

### 2017-04-01 hw08/hw08.rnw
Homework 8

\documentclass[11pt]{article}
\usepackage{fullpage,graphicx,float,amsmath,enumitem,fancyhdr}

\setlist{parsep=5.5pt}
\setlength{\parindent}{0pt}

\newcommand\assignment{Stat 534 Homework 8}
\newcommand\myname{Kenny Flagg}
\newcommand\duedate{March 31, 2017}

\pagestyle{fancy}
\lhead{\assignment}
\chead{\duedate}
\rhead{\myname}
\setlength{\headheight}{18pt}
\setlength{\headsep}{2pt}

\title{\assignment}
\author{\myname}
\date{\duedate}

\begin{document}
\maketitle

<<setup, echo = FALSE, message = FALSE, cache = FALSE>>=
library(knitr)
library(extrafont)
opts_chunk$set(echo = TRUE, comment = NA, message = FALSE,
               show.signif.stars = FALSE,
               fig.align = 'center', fig.width = 6.5, fig.height = 4,
               fig.pos = 'H', size = 'footnotesize', dev = 'pdf',
               dev.args = list(family = 'CM Roman', pointsize = 11))
knit_theme$set('print')

library(gridExtra)
library(xtable)
options(xtable.floating = FALSE,
        xtable.sanitize.rownames.function = function(x) x,
        xtable.sanitize.colnames.function = function(x) x,
        width = 80, scipen = 2, show.signif.stars = FALSE)

set.seed(6262)
@

\begin{enumerate}

\item%1
{\it The New York Leukemia data are attached along with homework as a comma
delimited text file. You are to explore how excluding 3 potential outliers
(observations 110, 120, and 121) affects the residual spatial autocorrelation,
regression results, and the conclusions obtained from them. Is weighting still
necessary? Fit \(OLS\), \(WLS\), and appropriate \(GLS\) models following my
steps in the example in class.}

I begin by fitting the OLS model, omitting the outliers, and examining the
semivariogram of the residuals. This semivariogram does not look very
different from the semivariogram of from the model fit the the entire dataset,
so I again use an initial range of 18 and an initial nugget of 0.9. I compare
all six models on the next page.

<<prob1_1>>=
library(nlme)
leukemia <- read.csv('leukemia.csv')

leuk_ols <- gls(z ~ pexp + age65 + home, data = leukemia, subset = -c(110, 120, 121))
plot(Variogram(leuk_ols, maxDist = 120))
@

<<prob1_2>>=
# Fit the weighted and spatial models.
leuk_wls <- gls(z ~ pexp + age65 + home, data = leukemia, subset = -c(110, 120, 121),
                weights = varFixed(~1/pop))
leuk_exp <- gls(z ~ pexp + age65 + home, data = leukemia, subset = -c(110, 120, 121),
                correlation = corExp(c(6, 0.9), form = ~x+y, nugget = TRUE))
leuk_sph <- gls(z ~ pexp + age65 + home, data = leukemia, subset = -c(110, 120, 121),
                correlation = corSpher(c(6, 0.9), form = ~x+y, nugget = TRUE))
leuk_exp_w <- gls(z ~ pexp + age65 + home, data = leukemia, subset = -c(110, 120, 121),
                  correlation = corExp(c(6, 0.9), form = ~x+y, nugget = TRUE),
                  weights = varFixed(~1/pop))
leuk_sph_w <- gls(z ~ pexp + age65 + home, data = leukemia, subset = -c(110, 120, 121),
                  correlation = corSpher(c(6, 0.9), form = ~x+y, nugget = TRUE),
                  weights = varFixed(~1/pop))
@

The table below shows the \texttt{pexp} coefficient estimates and AICs for all
six models. All the slope estimates are similar, but the AICs indicate that
the OLS model has the best fit. With the outliers removed, there is no need
for weighting.

\begin{center}
<<prob1_3, echo = FALSE, results = 'asis'>>=
aic <- AIC(leuk_ols, leuk_wls, leuk_exp, leuk_sph, leuk_exp_w, leuk_sph_w)$AIC

# Recreate the table from the notes.
out_table <- data.frame(rbind(
    summary(leuk_ols)$tTable['pexp', c('Value', 'Std.Error')],
    summary(leuk_wls)$tTable['pexp', c('Value', 'Std.Error')],
    summary(leuk_exp)$tTable['pexp', c('Value', 'Std.Error')],
    summary(leuk_sph)$tTable['pexp', c('Value', 'Std.Error')],
    summary(leuk_exp_w)$tTable['pexp', c('Value', 'Std.Error')],
    summary(leuk_sph_w)$tTable['pexp', c('Value', 'Std.Error')]),
    aic, aic - min(aic),
    row.names = c('OLS', 'WLS', 'Exponential', 'Spherical',
                 'Exponential, Weighted', 'Spherical, Weighted')
  )
colnames(out_table) <- c('\\(\\widehat{\\beta}_{1}\\)',
                         '\\(SE\\left(\\widehat{\\beta}_{1}\\right)\\)',
                         '\\(AIC\\)', '\\(\\Delta AIC\\)')

xtable(out_table, digits = c(0, 4, 4, 3, 3))
@
\end{center}

\item%2
{\it Imagine a lattice process on a \(2 \times 3\) rectangle. The sites
\(\mathbf{s}_{1}\), \(\mathbf{s}_{2}\), and \(\mathbf{s}_{3}\) make up the
first row, the remaining sites make up the second row. Assume that the spatial
connectivity matrix is given by}
\begin{equation*}
\mathbf{W} = \begin{bmatrix}
0 & 1 & 0 & 1 & 0 & 0 \\
1 & 0 & 1 & 0 & 1 & 0 \\
0 & 1 & 0 & 0 & 0 & 1 \\
1 & 0 & 0 & 0 & 1 & 0 \\
0 & 1 & 0 & 1 & 0 & 1 \\
0 & 0 & 1 & 0 & 1 & 0
\end{bmatrix}
\end{equation*}
{\it For a simultaneous and conditional autoregressive scheme with}
\begin{equation*}
\mathrm{Var}\left[\mathbf{Z}(\mathbf{s})\right]
= \sigma^{2} \left(\mathbf{I} - \rho \mathbf{W}\right)^{-1}
\left(\mathbf{I} - \rho \mathbf{W}'\right)^{-1}
\end{equation*}
{\it and}
\begin{equation*}
\mathrm{Var}\left[\mathbf{Z}(\mathbf{s})\right]
= \sigma^{2} \left(\mathbf{I} - \rho \mathbf{W}\right)^{-1}
\end{equation*}
{\it respectively, and with \(\rho = 0.25\) do the following:}

\begin{enumerate}

\item%a
{\it Identify the neighbors of lattice cell \(\mathbf{s}_2\)}.

The neighbors of \(\mathbf{s}_{2}\), corresponding to the columns of
\(\mathbf{W}\) that are nonzero in the second row, are \(\mathbf{s}_{1}\),
\(\mathbf{s}_{3}\), and \(\mathbf{s}_{5}\).

\item%b
{\it Compute the variance-covariance matrices for the SAR and CAR schemes.}

For SAR,

<<prob2b1>>=
W <- matrix(c(0, 1, 0, 1, 0, 0,
              1, 0, 1, 0, 1, 0,
              0, 1, 0, 0, 0, 1,
              1, 0, 0, 0, 1, 0,
              0, 1, 0, 1, 0, 1,
              0, 0, 1, 0, 1, 0),
            nrow = 6)
rho <- 0.25

Sigma_SAR <- solve(diag(6) - rho * W) %*% solve(diag(6) - rho * t(W))
print(Sigma_SAR)
@
so
\begin{equation*}
\mathrm{Var}\left[\mathbf{Z}(\mathbf{s})\right]
= \sigma^{2} \begin{bmatrix}
<<prob2b2, echo = FALSE, results = 'asis'>>=
cat(apply(format(Sigma_SAR, digits = 3), 1, paste, collapse = ' & '),
    sep = ' \\\\\n')
@
\end{bmatrix}.
\end{equation*}

For CAR,

<<prob2b3>>=
Sigma_CAR <- solve(diag(6) - rho * W) 
print(Sigma_CAR)
@
so
\begin{equation*}
\mathrm{Var}\left[\mathbf{Z}(\mathbf{s})\right]
= \sigma^{2} \begin{bmatrix}
<<prob2b4, echo = FALSE, results = 'asis'>>=
cat(apply(format(Sigma_CAR, digits = 3), 1, paste, collapse = ' & '),
    sep = ' \\\\\n')
@
\end{bmatrix}.
\end{equation*}

\pagebreak
\item%c
{\it Determine which of the processes is second-order stationary. Justify
your answer.}

Neither process is second-order stationary because the variance is not
spatially constant. For both processes, the variance of
\(\mathrm{Z}(\mathbf{s}_{2})\) and \(\mathrm{Z}(\mathbf{s}_{5})\) is larger
than the variance in other cells.

\item%d
{\it Describe the correlation patterns that result. Are observations
equicorrelated that are the same distance apart? Do correlations decrease
with increasing lag distance?}

The SAR correlation matrix is
<<prob2d1>>=
D_SAR_invsqrt <- diag(1 / sqrt(diag(Sigma_SAR)))
D_SAR_invsqrt %*% Sigma_SAR %*% D_SAR_invsqrt
@
and the CAR correlation matrix is
<<prob2d2>>=
D_CAR_invsqrt <- diag(1 / sqrt(diag(Sigma_CAR)))
D_CAR_invsqrt %*% Sigma_CAR %*% D_CAR_invsqrt
@
so we see that the correlations generally decrease as the lag increases, but
not all pairs at a given lag are equally correlated. Neighbors within a row
are more highly correlated than neighbors across rows, but pairs of cells in
the same row with lag 2 are less correlated than pairs in different rows with
lag 2.

Assuming square grid cells, the lag distance is equivalent to the Manhattan
distance. The correlation structure appears to reflect the Euclidean distance
because diagonally adjacent cells (lag 2; cells \(\mathbf{s}_{1}\) and
\(\mathbf{s}_{5}\), for example) are more correlated than pairs at lag 2 but
in the same row (e.g., cells \(\mathbf{s}_{1}\) and \(\mathbf{s}_{3}\)), and
the latter have a larger Euclidean distance.

\end{enumerate}

\pagebreak
\item%3
{\it We have 5 binomial count responses at 5 locations, i.e. the number of
successes out of \(n(\mathbf{s}_{i})\) trials. Assume a single covariate
\(X_{1}\) and that there is no overdispersion. Find the diagonal elements of
\(\mathbf{V}_{\mu}^{1/2}\) being sure to express these in terms of
\(\beta_{0}\) and \(\beta_{1}\).}

The model is
\begin{align*}
Z(\mathbf{s}_{i}) &\sim
\mathrm{Binomial}(n(\mathbf{s}_{i}), \mu(\mathbf{s}_{i})); \\
\log\left(\frac{\mu(\mathbf{s}_{i})}{1-\mu(\mathbf{s}_{i})}\right)
&= \beta_{0} + \beta_{1} x_{1}(\mathbf{s}_{i})
\end{align*}
for \(i = 1, \dots, 5\). Then, assuming no overdispersion, the \(i\)th
diagonal element of \(\mathbf{V}_{\mu}^{1/2}\) is
\begin{align*}
\sqrt{v(\mu(\mathbf{s}_{i}))}
&= \sqrt{n(\mathbf{s}_{i})\mu(\mathbf{s}_{i})(1-\mu(\mathbf{s}_{i}))} \\
&= \sqrt{n(\mathbf{s}_{i})
\left(\frac{\exp\left[\beta_{0} + \beta_{1} x_{1}(\mathbf{s}_{i})\right]}
{1 + \exp\left[\beta_{0} + \beta_{1} x_{1}(\mathbf{s}_{i})\right]}\right)
\left(\frac{1}{1 + \exp\left[\beta_{0}
+ \beta_{1} x_{1}(\mathbf{s}_{i})\right]}\right)} \\
&= \frac{\sqrt{n(\mathbf{s}_{i})\exp\left[\beta_{0}
+ \beta_{1} x_{1}(\mathbf{s}_{i})\right]}}
{1 + \exp\left[\beta_{0} + \beta_{1} x_{1}(\mathbf{s}_{i})\right]}.
\end{align*}

\end{enumerate}

\end{document}

### 2017-04-01 project/paper.rnw
Some wording and organization changes.

@@
 data, and it is very sensitive to the window size~\parencite{vspguide,wp}. In
-this project, I evaluate the use of polynomial trend surface models for the
-log-intensity at a simulated UXO site.
+this project, I review the use of polynomial trend surface models for the
+log-intensity at a simulated UXO site, ultimately concluding that these
+methods are not yet developed well enough to be useful.
 
@@
 
-\section{Examples}
-
-To illustrate model fitting and prediction, I present two examples before
-proceeding to the simulation study. The first is a toy example where I
-simulate from a Poisson process with a true log-linear intensity function and
-then estimate coefficients using the true model form. The second example
-introduces the my UXO site and provides a case study of fitting a polynomial
-trend surface to one realization of the the site.
-
-\section{True Log-Linear Intensity Surface}
+\section{True Log-Linear Intensity Surface Example}
 
@@
 
-Previously this semester, we worked with an example of a Poisson process on
-the unit square \(\{(x,y):\Sexpr{min(simulate_win$xrange)} < x
+Before moving on to a realistic example UXO site, I present a simple toy
+example using a true log-linear intensity surface to illustrate model fitting
+and prediction outside the observed region. Previously this semester, we
+worked with an example of a Poisson process on the unit square
+\(\{(x,y):\Sexpr{min(simulate_win$xrange)} < x
 < \Sexpr{max(simulate_win$xrange)}, \Sexpr{min(simulate_win$yrange)} < y
@@
 first using the process observed over the full region (the entire unit
-square), and then using the process observed in a subregion treating
+square), and then using the process observed in a subregion, treating
 \(\{(x,y): \Sexpr{min(omit_win$xrange)} < x < \Sexpr{max(omit_win$xrange)},
@@
 
+<<hw4_fit_full>>=
+hw4_fit_f <- ppm(hw4 ~ x + y, correction = 'isotropic')
+hw4_pred_f <- predict(hw4_fit_f, window = predict_win, se = TRUE)
+# By default predictions are on a 128 by 128 grid.
+@
+
+<<hw4_fit_s>>=
+hw4_fit_s <- ppm(hw4[observe_win] ~ x + y, correction = 'isotropic')
+hw4_pred_s <- predict(hw4_fit_s, window = predict_win, se = TRUE)
+@
+
 <<hw4_plot, fig.pos = 'b!', fig.cap = 'One realization of a Poisson process with log-linear intensity on the unit square.
-The dashed square shows the outline of the ``unobserved" region; when fitting the model to the subregion, the points inside it will be discarded and it will not be considered in constructing the approximation to the likelihood.'>>=
-par(mar = c(0, 0, 1, 0))
+The left panel shows all of the event locations, with the dashed square outlining of the ``unobserved" region.
+The right panel shows the events and dummy points used when fitting the model to the subregion; the small square is not considered in constructing the approximation to the likelihood, and the points inside it are discarded.'>>=
+par(mfrow = c(1, 2), mar = c(0, 0, 1, 0), cex = 1)
 plot(hw4, main = 'Event Locations')
 plot(omit_win, lty = 2, add = TRUE)
+plot(hw4_fit_s$Q, main = 'Data and Dummy Points')
 @
@@
 with its default likelihood approximation and isotropic edge correction.
-The coefficitents and standard errors are similar for both estimated models,
+The coefficients and standard errors are similar for both estimated models,
 and the true values \(\beta_{0} = 0\), \(\beta_{1} = 5\), and
@@
 models appear similar to the predicted intensities, as seen in the middle
-row of the figure~(shown on a logarithmic scale for visibility). In a linear
-model setting we would expect the prediction standard error to increase with
-the distance from the observations, but no such trend is apparent here; even
-in the bottom left of the prediction region the standard errors decrease as
-the predicted intensity decreases. However, it is illuminating to plot the
-predictor's relative standard error (the ratio of the standard error to the
-predicted intensity, bottom row of the figure). This ratio is not constant,
-and in fact it is lowest where the highest intensity of events was observed.
+row of the figure~(shown on a logarithmic scale for visibility). This makes
+intuitive sense because higher true point process intensity implies greater
+variance in the number of points, which should lead to greater uncertainty in
+estimation or prediction of the intensity. In a linear model setting we would
+expect the prediction standard error to increase with the distance from the
+observations, but no such trend is apparent here; even in the bottom left of
+the prediction region the standard errors decrease as the predicted intensity
+decreases. However, it is illuminating to plot the predictor's relative
+standard error (the ratio of the standard error to the predicted intensity,
+bottom row of the figure). This ratio is not constant, and in fact it is
+lowest where the highest intensity of events was observed.
 
@@
 the observed events than to the distance from the observed region.
-
-<<hw4_fit_full>>=
-hw4_fit_f <- ppm(hw4 ~ x + y, correction = 'isotropic')
-hw4_pred_f <- predict(hw4_fit_f, window = predict_win, se = TRUE)
-# By default predictions are on a 128 by 128 grid.
-@
-
-<<hw4_fit_s>>=
-hw4_fit_s <- ppm(hw4[observe_win] ~ x + y, correction = 'isotropic')
-hw4_pred_s <- predict(hw4_fit_s, window = predict_win, se = TRUE)
-@
 
@@
 \Sexpr{sprintf('%.1f\\%%', 100 * area(ex_win) / area(site_window))} of the
-qegion of interest.
+region of interest.
+
+If the general structure of the site (such as the number and shape of the
+high-intensity regions and the type of background process) is known then the
+most appropriate model for the anomaly intensity could be a mixture model,
+but in practice these details are rarely available. Therefore, practitioners
+prefer flexible models without defined structure. Polynomial models are
+conceptually simple, and capable of appriximating complicated surfaces. I
+fit \(p\)-degree polynomial models of the form
+\begin{equation*}
+\log\left(\lambda\left(x, y\right)\right)
+= \sum_{i = 0}^{p} \sum_{j=0}^{p-i} \beta_{ij} x^{i} y^{j}
+\end{equation*}
+for \(p = 2, 3, \dots, 12\). Here, the coordinates \(x\) and \(y\) are
+standardized to have mean 0 and standard variance 1 in order to reduce
+numerical instability for the larger exponents.
 

### 2017-04-02 project/paper.rnw
Taking a break due to lazy load errors.

@@
 \newcommand\myname{Kenny Flagg}
-\newcommand\duedate{\today}
+\newcommand\duedate{April 2, 2017}
 
@@
 
-<<hw4_fitplot, fig.pos = 'p', fig.height = 8, fig.cap = 'Estimated log-intensity surface, log-scale prediction standard errors, and relative standard error from models fit using all events in the full simulation region (left) and events in a subset of the simulation region (right).
+<<hw4_fitplot, fig.pos = 'p', fig.height = 8, fig.cap = 'Estimated log-intensity surface, log-scale prediction standard errors, and relative standard errors from models fit using all events in the full simulation region (left) and events in a subset of the simulation region (right).
 The regions and event locations are overlaid in white.'>>=
@@
 high-intensity regions and the type of background process) is known then the
-most appropriate model for the anomaly intensity could be a mixture model,
-but in practice these details are rarely available. Therefore, practitioners
-prefer flexible models without defined structure. Polynomial models are
-conceptually simple, and capable of appriximating complicated surfaces. I
-fit \(p\)-degree polynomial models of the form
+most appropriate model for the anomaly intensity could be a mixture model or
+a trend surface with a particular form, but in practice these details are
+rarely available. Therefore, practitioners prefer flexible models without
+defined structure. I fit \(p\)-degree polynomial models of the form
 \begin{equation*}
@@
 \end{equation*}
-for \(p = 2, 3, \dots, 12\). Here, the coordinates \(x\) and \(y\) are
-standardized to have mean 0 and standard variance 1 in order to reduce
+for \(p = 2, 3, \dots, 12\) using 500,000 dummy points randomly selected from
+a uniform distribution over the observed region. The coordinates \(x\) and
+\(y\) are standardized to have mean 0 and variance 1 in order to reduce
 numerical instability for the larger exponents.
 
-use \texttt{polynom()} not \texttt{poly()}!
-
-<<polyfit1, cache = TRUE, dependson = 'sim2000obs', warning = FALSE, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = ''>>=
+<<polyfit1, cache = TRUE, dependson = 'sim2000obs', warning = FALSE>>=
 # Do a simple fit where spatstat sets up the quadrature so we can see where it
@@
 # each transect.
-# TODO: Figure out weights.
 ex_Q <- quad(data = ex_ppp, dummy = runifpoint(500000, Window(ex_ppp)))
@@
 
+ex_kde <- density(ex_ppp_full)
+@
+
+Figure~\ref{fig:polyplot1} shows the predicted intensity surfaces for the
+polynomial models, along with a kernel density estimate for comparison. The
+2nd-order polynomial surface is too simple to capture the bimodal surface,
+
+The kernel density estimate is computed using the \texttt{spatstat}'s default
+``simple rule of thumb'' to select the smoothing bandwidth. Because the
+anomalies located more than three feet from a transect were not observed, I
+would expect the kernel density estimate to show ridges of high intensity
+along the transects. As it turns out, the default bandwith is large enough to
+smooth across multiple transects so the ridges do not appear. However, the
+kernel density estimate is biased low because not all anomalies are observed;
+I divide the estimate by the proportion of the area observed as an approximate
+bias correction.
+
+<<polyplot1, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = 'Predicted intensity surfaces for eleven polynomial models, and a kernel density estimate for comparision.'>>=
 par(mfrow = c(4, 3), mar = c(1, 0, 1.2, 3), las = 2, cex = 1)
@@
 # to anomalies per acre.
-plot(43560 * ex_p2$estimate, main = '2nd Degree')
-plot(43560 * ex_p3$estimate, main = '3rd Degree')
-plot(43560 * ex_p4$estimate, main = '4th Degree')
-plot(43560 * ex_p5$estimate, main = '5th Degree')
-plot(43560 * ex_p6$estimate, main = '6th Degree')
-plot(43560 * ex_p7$estimate, main = '7th Degree')
-plot(43560 * ex_p8$estimate, main = '8th Degree')
-plot(43560 * ex_p9$estimate, main = '9th Degree')
-plot(43560 * ex_p10$estimate, main = '10th Degree')
-plot(43560 * ex_p11$estimate, main = '11th Degree')
-plot(43560 * ex_p12$estimate, main = '12th Degree')
-plot(43560 * density(ex_ppp_full) * area(ex_ppp_full) / area(ex_ppp),
+plot(43560 * ex_p2$estimate, zlim = c(0, 500), main = '2nd Degree')
+plot(43560 * ex_p3$estimate, zlim = c(0, 500), main = '3rd Degree')
+plot(43560 * ex_p4$estimate, zlim = c(0, 500), main = '4th Degree')
+plot(43560 * ex_p5$estimate, zlim = c(0, 500), main = '5th Degree')
+plot(43560 * ex_p6$estimate, zlim = c(0, 500), main = '6th Degree')
+plot(43560 * ex_p7$estimate, zlim = c(0, 500), main = '7th Degree')
+plot(43560 * ex_p8$estimate, zlim = c(0, 500), main = '8th Degree')
+plot(43560 * ex_p9$estimate, zlim = c(0, 500), main = '9th Degree')
+plot(43560 * ex_p10$estimate, zlim = c(0, 500), main = '10th Degree')
+plot(43560 * ex_p11$estimate, zlim = c(0, 500), main = '11th Degree')
+plot(43560 * ex_p12$estimate, zlim = c(0, 500), main = '12th Degree')
+plot(43560 * ex_kde * area(ex_ppp_full) / area(ex_ppp), zlim = c(0, 500),
      main = 'Kernel Density')
@@
 
-<<polyse1, cache = TRUE, dependson = 'polyfit1', fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = ''>>=
+<<polyse1, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = 'Log standard errors and relative standard errors for six of the polynomial models.'>>=
 par(mfrow = c(4, 3), mar = c(1, 0, 2, 3), las = 2, cex = 1)
@@
 @
+
+\section{Difficulties with Implementation}
 

### 2017-04-03 project/paper.rnw
Fix cache for presentation.

@@
 
-<<polyfit1, cache = TRUE, dependson = 'sim2000obs', warning = FALSE>>=
+<<quad1, cache = TRUE, dependson = 'sim2000obs', warning = FALSE>>=
 # Do a simple fit where spatstat sets up the quadrature so we can see where it
@@
 ex_Q <- quad(data = ex_ppp, dummy = runifpoint(500000, Window(ex_ppp)))
+@
 
+<<polyfit2, cache = TRUE, dependson = 'quad1'>>=
 ex_fit2 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 2), correction = 'iso')
+@
+<<polyfit3, cache = TRUE, dependson = 'quad1'>>=
 ex_fit3 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 3), correction = 'iso')
+@
+<<polyfit4, cache = TRUE, dependson = 'quad1'>>=
 ex_fit4 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 4), correction = 'iso')
+@
+<<polyfit5, cache = TRUE, dependson = 'quad1'>>=
 ex_fit5 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 5), correction = 'iso')
+@
+<<polyfit6, cache = TRUE, dependson = 'quad1'>>=
 ex_fit6 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 6), correction = 'iso')
+@
+<<polyfit7, cache = TRUE, dependson = 'quad1'>>=
 ex_fit7 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 7), correction = 'iso')
+@
+<<polyfit8, cache = TRUE, dependson = 'quad1'>>=
 ex_fit8 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 8), correction = 'iso')
+@
+<<polyfit9, cache = TRUE, dependson = 'quad1'>>=
 ex_fit9 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 9), correction = 'iso')
+@
+<<polyfit10, cache = TRUE, dependson = 'quad1'>>=
 ex_fit10 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 10), correction = 'iso')
+@
+<<polyfit11, cache = TRUE, dependson = 'quad1'>>=
 ex_fit11 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 11), correction = 'iso')
+@
+<<polyfit12, cache = TRUE, dependson = 'quad1'>>=
 ex_fit12 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 12), correction = 'iso')
+@
 
+<<polypred1>>=
 # Get predictions and SEs.
@@
 
-ex_kde <- density(ex_ppp_full)
+ex_kde <- density(ex_ppp_full) * area(ex_ppp_full) / area(ex_ppp)
 @
@@
 Figure~\ref{fig:polyplot1} shows the predicted intensity surfaces for the
-polynomial models, along with a kernel density estimate for comparison. The
-2nd-order polynomial surface is too simple to capture the bimodal surface,
+polynomial models, along with a kernel density estimate for comparison.
+500,000 dummy points do not seem to be enough for the likelihood
+approximation to converge to the Poisson likelihood, but the predicted
+intensity values are reasonable (discussed further in the next section).
+Because of the lack of convergence, I do not do any formal model comparisons.
+
+The second- and third-order polynomial models are too simple to capture the
+bimodality of the true surface, but the fourth- and higher-order models have
+large, high-intensity regions that coincide with both concentrated munitions
+use regions of the true surface. Up until degree nine, the polynomial models
+are smoother than the kernel intensity estimate. The ninth-, tenth-, and
+eleventh-degree models all look similar to the kernel intensity estimate,
+while the twelfth-degree model is noticeably noisier. Edge effects are
+noticeable for \(p \geq 4\) with some predictions along the boundary abruptly
+dropping to zero or increasing to hundreds of anomalies per acre higher than
+any of the predictions in the interior of the region.
 
@@
 I divide the estimate by the proportion of the area observed as an approximate
-bias correction.
+bias correction and the result is a surprisingly reasonable intensity map.
 
@@
 # to anomalies per acre.
-plot(43560 * ex_p2$estimate, zlim = c(0, 500), main = '2nd Degree')
-plot(43560 * ex_p3$estimate, zlim = c(0, 500), main = '3rd Degree')
-plot(43560 * ex_p4$estimate, zlim = c(0, 500), main = '4th Degree')
-plot(43560 * ex_p5$estimate, zlim = c(0, 500), main = '5th Degree')
-plot(43560 * ex_p6$estimate, zlim = c(0, 500), main = '6th Degree')
-plot(43560 * ex_p7$estimate, zlim = c(0, 500), main = '7th Degree')
-plot(43560 * ex_p8$estimate, zlim = c(0, 500), main = '8th Degree')
-plot(43560 * ex_p9$estimate, zlim = c(0, 500), main = '9th Degree')
-plot(43560 * ex_p10$estimate, zlim = c(0, 500), main = '10th Degree')
-plot(43560 * ex_p11$estimate, zlim = c(0, 500), main = '11th Degree')
-plot(43560 * ex_p12$estimate, zlim = c(0, 500), main = '12th Degree')
-plot(43560 * ex_kde * area(ex_ppp_full) / area(ex_ppp), zlim = c(0, 500),
-     main = 'Kernel Density')
+plot(43560 * ex_p2$estimate, zlim = c(0, 300), main = '2nd Degree')
+plot(43560 * ex_p3$estimate, zlim = c(0, 300), main = '3rd Degree')
+plot(43560 * ex_p4$estimate, zlim = c(0, 300), main = '4th Degree')
+plot(43560 * ex_p5$estimate, zlim = c(0, 300), main = '5th Degree')
+plot(43560 * ex_p6$estimate, zlim = c(0, 300), main = '6th Degree')
+plot(43560 * ex_p7$estimate, zlim = c(0, 300), main = '7th Degree')
+plot(43560 * ex_p8$estimate, zlim = c(0, 300), main = '8th Degree')
+plot(43560 * ex_p9$estimate, zlim = c(0, 300), main = '9th Degree')
+plot(43560 * ex_p10$estimate, zlim = c(0, 300), main = '10th Degree')
+plot(43560 * ex_p11$estimate, zlim = c(0, 300), main = '11th Degree')
+plot(43560 * ex_p12$estimate, zlim = c(0, 300), main = '12th Degree')
+plot(43560 * ex_kde, zlim = c(0, 500), main = 'Kernel Density')
 @
 
-<<polyse1, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = 'Log standard errors and relative standard errors for six of the polynomial models.'>>=
-par(mfrow = c(4, 3), mar = c(1, 0, 2, 3), las = 2, cex = 1)
+Figure~\ref{fig:polyse1} shows the log-scale standard errors for six of the
+polynomial models. For all models except the twelfth-degree model, the
+standard error is low across most of the site but high along the boundary. The
+standard errors increase with the degree of the model. This is expected and
+resonable because locations in the interior of the region have the most nearby
+anomalies and dummy points, and predictions from more complex models depend on
+more estimated parameters.
+
+If the true number of high-intensity regions was known to be 2, any polynomial
+model with \(p \geq 4\) (or the kernel density estimate) would provide an
+accurate (if imprecise) indication of where to focus the UXO remediation
+effort. However, the plots alone provide little indication of which model has
+the ideal level of complexity, but there are several practical difficulties
+with fitting these models that should be sorted out before I would be
+comfortable making recommendations about how to select an optimal model for
+prediction outside the observed subregion.
+
+<<polyse1, fig.width = 6, fig.height = 4, fig.pos = 't', fig.cap = 'Log standard errors for six of the polynomial models.'>>=
+par(mfrow = c(2, 3), mar = c(1, 0, 2, 3), las = 2, cex = 1)
 plot(log(43560 * ex_p2$se), main = '2nd Degree\nlog(SE)')
@@
 plot(log(43560 * ex_p6$se), main = '6th Degree\nlog(SE)')
-plot(ex_p2$se / ex_p2$estimate, main = '2nd Degree\nRSE')
-plot(ex_p4$se / ex_p4$estimate, main = '4th Degree\nRSE')
-plot(ex_p6$se / ex_p6$estimate, main = '6th Degree\nRSE')
-
 plot(log(43560 * ex_p8$se), main = '8th Degree\nlog(SE)')
@@
 plot(log(43560 * ex_p12$se), main = '12th Degree\nlog(SE)')
-plot(ex_p8$se / ex_p8$estimate, main = '8th Degree\nRSE')
-plot(ex_p10$se / ex_p10$estimate, main = '10th Degree\nRSE')
-plot(ex_p12$se / ex_p12$estimate, main = '12th Degree\nRSE')
 @
@@
 window
+
+<<dummy>>=
+par(mfrow = c(1, 2), mar = c(1, 0, 1, 0), cex = 1)
+plot(ex_fit1$Q$dummy, pch = '.', border = NA,
+     main = 'Default Dummy Points')
+plot(site_window, add = TRUE)
+plot(ex_Q$dummy, pch = '.', border = NA,
+     main = 'Manual Dummy Points')
+plot(site_window, add = TRUE)
+@
 
@@
 
-\appendix
-\section{R Code Appendix}
-
 \printbibliography
 
+\pagebreak
+\appendix
+\section*{R Code Appendix}
+
+<<packages, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-20pt}
+<<hw4_sim, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-20pt}
+<<hw4_fit_full, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-20pt}
+<<hw4_fit_s, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-20pt}
+<<hw4_plot, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-20pt}
+<<hw4_fitplot, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-20pt}
+<<uxoexampledata, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-20pt}
+<<truth, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-20pt}
+<<sim2000obs, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-20pt}
+<<quad1, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-20pt}
+<<polyfit2, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-30pt}
+<<polyfit3, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-30pt}
+<<polyfit4, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-30pt}
+<<polyfit5, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-30pt}
+<<polyfit6, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-30pt}
+<<polyfit7, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-30pt}
+<<polyfit8, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-30pt}
+<<polyfit9, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-30pt}
+<<polyfit10, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-30pt}
+<<polyfit11, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-30pt}
+<<polyfit12, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-20pt}
+<<polypred1, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-20pt}
+<<polyplot1, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-20pt}
+<<polyse1, echo = TRUE, eval = FALSE>>=
+@
+\vspace{-20pt}
+<<dummy, echo = TRUE, eval = FALSE>>=
+@
+
 \end{document}

### 2017-04-03 project/paper.rnw
Fixed cache.

@@
 
-# Create a ppp object the uses the full region as the window.
+# Create a ppp object that uses the full region as the window.
 ex_ppp_full <- ppp(ex_anom$x, ex_anom$y, window = site_window)
@@
 @
-\vspace{-20pt}
+\vspace{-15pt}
 <<hw4_sim, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-20pt}
+\vspace{-15pt}
 <<hw4_fit_full, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-20pt}
+\vspace{-15pt}
 <<hw4_fit_s, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-20pt}
+\vspace{-15pt}
 <<hw4_plot, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-20pt}
+\vspace{-15pt}
 <<hw4_fitplot, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-20pt}
+\vspace{-15pt}
 <<uxoexampledata, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-20pt}
+\vspace{-15pt}
 <<truth, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-20pt}
+\vspace{-15pt}
 <<sim2000obs, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-20pt}
+\vspace{-15pt}
 <<quad1, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-20pt}
+\vspace{-15pt}
 <<polyfit2, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-30pt}
+\vspace{-25pt}
 <<polyfit3, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-30pt}
+\vspace{-25pt}
 <<polyfit4, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-30pt}
+\vspace{-25pt}
 <<polyfit5, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-30pt}
+\vspace{-25pt}
 <<polyfit6, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-30pt}
+\vspace{-25pt}
 <<polyfit7, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-30pt}
+\vspace{-25pt}
 <<polyfit8, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-30pt}
+\vspace{-25pt}
 <<polyfit9, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-30pt}
+\vspace{-25pt}
 <<polyfit10, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-30pt}
+\vspace{-25pt}
 <<polyfit11, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-30pt}
+\vspace{-25pt}
 <<polyfit12, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-20pt}
+\vspace{-15pt}
 <<polypred1, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-20pt}
+\vspace{-15pt}
 <<polyplot1, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-20pt}
+\vspace{-15pt}
 <<polyse1, echo = TRUE, eval = FALSE>>=
 @
-\vspace{-20pt}
+\vspace{-15pt}
 <<dummy, echo = TRUE, eval = FALSE>>=

### 2017-04-03 project/pres.rnw
Typos are a pain.
Fixed cache.

@@
 &= \sum_{i=1}^{n}\log\left(\lambda(\mathbf{s}_{i})\right)
--\int_{D}\lambda(\mathbf{s})d\mathbf{s} - \log(n!).
+-\int_{D}\lambda(\mathbf{s})d\mathbf{s} - \log(n!)
 \end{align*}
@@
 ex_Q <- quad(data = ex_ppp, dummy = runifpoint(500000, Window(ex_ppp)))
+@
+<<polyfit2, cache = TRUE, dependson = 'polyfit'>>=
+ex_fit2 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 2), correction = 'iso')
+@
+<<polyfit3, cache = TRUE, dependson = 'polyfit'>>=
+ex_fit3 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 3), correction = 'iso')
+@
+<<polyfit4, cache = TRUE, dependson = 'polyfit'>>=
+ex_fit4 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 4), correction = 'iso')
+@
+<<polyfit5, cache = TRUE, dependson = 'polyfit'>>=
+ex_fit5 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 5), correction = 'iso')
+@
+<<polyfit6, cache = TRUE, dependson = 'polyfit'>>=
+ex_fit6 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 6), correction = 'iso')
+@
+<<polyfit7, cache = TRUE, dependson = 'polyfit'>>=
+ex_fit7 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 7), correction = 'iso')
+@
+<<polyfit8, cache = TRUE, dependson = 'polyfit'>>=
+ex_fit8 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 8), correction = 'iso')
+@
+<<polyfit9, cache = TRUE, dependson = 'polyfit'>>=
+ex_fit9 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 9), correction = 'iso')
+@
+<<polyfit10, cache = TRUE, dependson = 'polyfit'>>=
+ex_fit10 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 10), correction = 'iso')
+@
+<<polyfit11, cache = TRUE, dependson = 'polyfit'>>=
+ex_fit11 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 11), correction = 'iso')
+@
+<<polyfit12, cache = TRUE, dependson = 'polyfit'>>=
+ex_fit12 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 12), correction = 'iso')
+@
 
-ex_fit2 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 2), correction = 'iso')
-ex_fit3 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 3), correction = 'iso')
-ex_fit4 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 4), correction = 'iso')
-ex_fit5 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 5), correction = 'iso')
-ex_fit6 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 6), correction = 'iso')
-ex_fit7 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 7), correction = 'iso')
-ex_fit8 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 8), correction = 'iso')
-ex_fit9 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 9), correction = 'iso')
-ex_fit10 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 10), correction = 'iso')
-ex_fit11 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 11), correction = 'iso')
-ex_fit12 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 12), correction = 'iso')
-
+<<polypred>>=
 ex_p2 <- predict(ex_fit2, window = site_window, se = TRUE)
@@
 
-<<resplots1, cache = TRUE, dependson = 'polyfit'>>=
+<<resplots1>>=
 par(mfrow = c(2, 3), mar = c(1, 0, 1.2, 3), las = 2, cex = 1)
@@
 
-<<resplots2, cache = TRUE, dependson = 'polyfit'>>=
+<<resplots2>>=
 par(mfrow = c(2, 3), mar = c(1, 0, 1.2, 3), las = 2, cex = 1)

### 2017-04-04 project/paper.rnw
About done with the project.

@@
 \newcommand\myname{Kenny Flagg}
-\newcommand\duedate{April 2, 2017}
+\newcommand\duedate{April 3, 2017}
 
@@
 plot(intense.im, main = 'True Intensity Surface',
-     border = NA, zlim = c(100, 300), las = 2)
+     border = NA, zlim = c(0, 300), las = 2)
 plot(setminus.owin(owin(c(1556870, 1564495), c(535421, 541130),
@@
 
-<<polypred1>>=
+<<polypred1, cache = TRUE, dependson = 'quad1'>>=
 # Get predictions and SEs.
@@
 
-The kernel density estimate is computed using the \texttt{spatstat}'s default
+The kernel density estimate is computed using \texttt{spatstat}'s default
 ``simple rule of thumb'' to select the smoothing bandwidth. Because the
@@
 
-<<polyplot1, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = 'Predicted intensity surfaces for eleven polynomial models, and a kernel density estimate for comparision.'>>=
+<<polyplot1, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = 'Predicted intensity surfaces for eleven polynomial models, and a kernel density estimate for comparision.
+Note that the 4th- through 12th-degree polynomial models have the same color scale, while the 2nd- and 3rd-degree models and the kernel density estimate use different color scales for visibility.'>>=
 par(mfrow = c(4, 3), mar = c(1, 0, 1.2, 3), las = 2, cex = 1)
@@
 # to anomalies per acre.
-plot(43560 * ex_p2$estimate, zlim = c(0, 300), main = '2nd Degree')
-plot(43560 * ex_p3$estimate, zlim = c(0, 300), main = '3rd Degree')
+plot(43560 * ex_p2$estimate, main = '2nd Degree')
+plot(43560 * ex_p3$estimate, main = '3rd Degree')
 plot(43560 * ex_p4$estimate, zlim = c(0, 300), main = '4th Degree')
@@
 plot(43560 * ex_p12$estimate, zlim = c(0, 300), main = '12th Degree')
-plot(43560 * ex_kde, zlim = c(0, 500), main = 'Kernel Density')
+plot(43560 * ex_kde, main = 'Kernel Density')
 @
@@
 
-<<polyse1, fig.width = 6, fig.height = 4, fig.pos = 't', fig.cap = 'Log standard errors for six of the polynomial models.'>>=
+<<polyse1, fig.width = 6, fig.height = 4, fig.pos = 'p', fig.cap = 'Log standard errors for six of the polynomial models.'>>=
 par(mfrow = c(2, 3), mar = c(1, 0, 2, 3), las = 2, cex = 1)
@@
 
-\section{Difficulties with Implementation}
-
-\texttt{plot} doesn't work, and \texttt{predict} doesn't work on the original
-window
-
-<<dummy>>=
+<<dummy, fig.pos = 'p', fig.cap = 'The default dummy points selected by \\texttt{spatstat} (left) and the 500,000 random uniform dummy points that I use to fit the models.'>>=
 par(mfrow = c(1, 2), mar = c(1, 0, 1, 0), cex = 1)
@@
 
+\section{Difficulties with Implementation}
+
+I encountered many difficulties in fitting these models that suggest the
+methods and software are not yet developed well enough to make them useful for
+working with such sparsely sampled data. Most of these difficulties resulted
+from the implementation with \texttt{spatstat}, but several were computational
+issues.
+
+The troubles with \texttt{spatstat} largely come from how the package sets up
+grids of locations for dummy points and predictions. It finds a bounding box
+for the region and places grid cells on a lattice across the bounding box, but
+its grids are so coarse that most of my transects do not have cells centered
+on them. It attempts to create an \(80 \times 80\) grid of dummy points, but
+the only useable points are 160 points on two transects; all others are
+discarded~(Figure~\ref{fig:dummy}, left). Attempting to fit models in
+\texttt{ppm} using these dummy points results in an error.
+
+The \texttt{ppm} function fits models successfully when using the 500,000
+dummy points that I draw from a uniform distribution~(Figure~\ref{fig:dummy},
+right). However, the \texttt{plot} method gives an error about infinite
+values, while the \texttt{predict} method by default uses the same
+\(80 \times 80\) grid used for the default dummy points and makes predictions
+only in the 160 grid cells that happen to lie inside the observed region.
+Fortunately, the \texttt{predict} function behaves when the whole site is
+specified as the prediction window.
+
+The biggest disappoinment of using \texttt{spatstat} to fit these models is
+that models with splines cannot be plotted. The \texttt{ppm} functions
+estimates the parameters without any apparent trouble, but the \texttt{plot}
+and \texttt{predict} functions give errors because they do not know how to
+interpret the smooth terms in the model formula.
+
+All of the above difficulties are software issues that could be fixed. I also
+found two computational limitations that will not be remedied as easily.
+First, even after rescaling the coordinates, \(\mathbf{X}^{T}\mathbf{X}\)
+cannot be inverted because of numerical instabilities when \(p \geq 18\). Such
+models are more complicated than needed for my example site, but it is
+important to note that a limit on polynomial model complexity exists.
+
+Second, choosing dummy points is hard. I initially used 1,000 points, but that
+was not enough and the intensity values predicted by those models were in the
+tens of thousands. It is not apparent that the 500,000 randomly-selected dummy
+points I ultimately used are optimal, either. \textcite{wartonshepherd}
+discuss dummy point selection when the entire region of interest is observed,
+suggesting that the number of dummy points be increased until the maximized
+log-likelihood reaches approximate convergence. They use 86,227 points for
+their example. I used trial-and-error to increase the number of dummy points,
+fitting a couple of my models using 600,000 and then 1,000,000 dummy points;
+the maximized log-likelihood changed by thousands. I settled on 500,000
+because that was the smallest number I tried that yielded predicted
+intensities of the correct magnitude. With that many points, each model takes
+about a minute to fit; models with enough dummy points to accurately
+approximate the likelihood would likely take much longer to estimate.
+
+\pagebreak
 \section{Discussion and Conclusion}
 
-smallish SEs outside observed region because of faith in the model form --
-need model checking
-
-What geostatisticians and ordnance engineers will notice is that that SEs
-don't reflect added uncertainty about unobserved regions.
-Note kriging accounts for lack of model fit by autocorrelated errors --
-prediction SEs increase far from obs because of lack of info about local
-deviations from mean. For Poisson process, the events are independent given
-the intensity so local variation is not helpful for prediction. Thus we are
-heavily reliant on the correctness of the model form and the
-representativeness of the observed subregion, so model checking is crucial.
-If there is prior uncertainty about the form of the trend in a certain part of
-the site, that region must be observed!
-
-how much of the region do you need to observe to trust your model?
+Spatial point process intensity surface models are an intriguing option for
+modeling partially-observed inhomogeneous point processes, and polynomial
+models can approximate complex surfaces. However, these models are not
+yet ready for application to UXO data because the software is not quite there
+(yet) and they require heavy computing power. Several avenues for future work
+are available, such as seeking more efficient ways to approximate the
+likelihood and compute parameter estimates, and further exploring how to
+choose dummy points. Then, the next step might be to study whether familiar
+model selection tools like AIC or tests of predictive ability are suitable
+assuming the likelihood approximation is successful. Presently, it seems
+necessary either to observe more of the site area than covered by typical UXO
+sampling plans, or work with a model fit that is a rough approximation at
+best. Thus the biggest open question might be how much of the region do you
+need to observe to trust your model?
 

### 2017-04-04 project/paper.rnw
Spell check.

@@
 subregion to predict the intensity over the entire region of interest. This
-procedure would be relevant whenver it is known or suspected that some type of
+procedure would be relevant whenever it is known or suspected that some type of
 plant, animal, or other item occurs in the region, and the goal of the
@@
 are only observed when detection equipment passes directly over them, but the
-project leaders are concerened with finding UXO and do not want to waste
+project leaders are concerned with finding UXO and do not want to waste
 resources finding every inert fragment that could be found by metal detectors.
@@
 Poisson distribution with mean \(\mu=\int_{D}\lambda(\mathbf{s})d\mathbf{s}\).
-Given that \(n\) events occured, their locations \(\mathbf{s}_{1},\dots,
+Given that \(n\) events occurred, their locations \(\mathbf{s}_{1},\dots,
 \mathbf{s}_{n}\) are independent and identically distributed with density
@@
 partition contains at most one event, and then maximizing a weighted
-pseudolikelihood based upon a binomial or Poission distribution for the count
+pseudolikelihood based upon a binomial or Poisson distribution for the count
 of events in each subset. They comment that the covariance matrices produced
@@
 use generated as Poisson processes with bivariate Gaussian intensity. The
-total intesnsity at the center of each concentrated munitions use region is
+total intensity at the center of each concentrated munitions use region is
 300 anomalies per acre~(Figure~\ref{fig:truth}).
@@
 <<truth, cache = TRUE, dependson = 'uxoexampledata', fig.width = 4, fig.pos = 't', fig.cap = 'True intensity at the simulated UXO site.'>>=
-# Bivariate normal intensity function that I originalally wrote for my
+# Bivariate normal intensity function that I originally wrote for my
 # writing project, parameterized in terms of the major and minor axes scales
@@
 # Load the sample data used in my writing project.
-# The anomaly file contains the locations of the observed anamolies and the
+# The anomaly file contains the locations of the observed anomalies and the
 # cog file contains the course-over-ground waypoints on the transects.
@@
 would expect the kernel density estimate to show ridges of high intensity
-along the transects. As it turns out, the default bandwith is large enough to
+along the transects. As it turns out, the default bandwidth is large enough to
 smooth across multiple transects so the ridges do not appear. However, the
@@
 standard errors increase with the degree of the model. This is expected and
-resonable because locations in the interior of the region have the most nearby
+reasonable because locations in the interior of the region have the most nearby
 anomalies and dummy points, and predictions from more complex models depend on
@@
 on them. It attempts to create an \(80 \times 80\) grid of dummy points, but
-the only useable points are 160 points on two transects; all others are
+the only usable points are 160 points on two transects; all others are
 discarded~(Figure~\ref{fig:dummy}, left). Attempting to fit models in
@@
 
-The biggest disappoinment of using \texttt{spatstat} to fit these models is
+The biggest disappointment of using \texttt{spatstat} to fit these models is
 that models with splines cannot be plotted. The \texttt{ppm} functions

### 2017-04-07 hw09/hw09.rnw
Hw9

\documentclass[11pt]{article}
\usepackage{fullpage,graphicx,float,amsmath,enumitem,fancyhdr}

\setlist{parsep=5.5pt}
\setlength{\parindent}{0pt}

\newcommand\assignment{Stat 534 Homework 9}
\newcommand\myname{Kenny Flagg}
\newcommand\duedate{April 7, 2017}

\pagestyle{fancy}
\lhead{\assignment}
\chead{\duedate}
\rhead{\myname}
\setlength{\headheight}{18pt}
\setlength{\headsep}{2pt}

\title{\assignment}
\author{\myname}
\date{\duedate}

\begin{document}
\maketitle

<<setup, echo = FALSE, message = FALSE, cache = FALSE>>=
library(knitr)
library(extrafont)
opts_chunk$set(echo = TRUE, comment = NA, message = FALSE,
               show.signif.stars = FALSE,
               fig.align = 'center', fig.width = 6.5, fig.height = 4,
               fig.pos = 'H', size = 'footnotesize', dev = 'pdf',
               dev.args = list(family = 'CM Roman', pointsize = 11))
knit_theme$set('print')

library(nlme)
library(MASS)
library(geoR)

library(xtable)
options(xtable.floating = FALSE,
        xtable.sanitize.rownames.function = function(x) x,
        xtable.sanitize.colnames.function = function(x) x,
        width = 80, scipen = 2, show.signif.stars = FALSE)

set.seed(92387)
@

\begin{enumerate}

\item%1
{\it I would like you to try and reproduce some of the results in Table 9.10
on page 395 in Waller and Gotway. They fit several models to the Scottish
lipcancer data set. The ones I am interested in are models \(S+OD\) and the
\(MGLM\). I want you to use \texttt{glmmPQL} to see if you can get close to
their fitted results for \(S+OD\). I provide you with some code for fitting
the nonlinear least squares model to see what you can come up with as an
approximation to their \(MGLM\). We did something similar with the Virginia
lead level data earlier (take a look back at page 50 or thereabouts). I have
spent hours every time I teach this course trying to come close to their
results. So I have two possible outcomes in mind here: (1) you (or some of
you) are able to figure out where I am going wrong which I would love to see,
and/or (2) you get a birds-eye view of the issues related to trying to
incorporate spatial correlation structures into generalized linear models.}

{\it The data set they provided on their website is attached. The last 2
columns are the transformed spatial locations they discuss in the text. If you
look at the SAS code they used they divided these by 1000. I encourage you to
work together on this one. It will go much smoother. I would like for you all
to be ready to discuss this on Friday. I have been a bit easy on the exact
time homeworks are due, accepting several late over the course of the
semester. But I want to be able to talk about this in class on Friday so you
need to have them ready by class time.}

<<data, echo = FALSE>>=
lipcancer <- read.csv('lipcancer.WallerGotway.csv')

# I'm mostly sticking to the same variable names as in their SAS code:
#   http://web1.sph.emory.edu/users/lwaller/book/ch9/scotglms.sas
lipcancer$aff1 <- lipcancer$aff/10
lipcancer$logni <- log(lipcancer$exp)
lipcancer$x <- lipcancer$xcoord/1000
lipcancer$y <- lipcancer$ycoord/1000
lipcancer$logsmr <- log(lipcancer$smr + 1)
lipcancer$expct <- lipcancer$exp
@

I first spent some time trying to understand their semivariogram plot. They
say they computed it from the studentized residuals of a WLS regression of
\(\log(\texttt{SMR}+1)\) against \texttt{aff} using the expected counts as
weights. I translated their SAS code to get the plot on the next page. Two
counties are clear outliers with studentized residuals below \(-2\), and
removing them yields a very different semivariogram. Waller and Gotway don't
give any indication that they give these observations special handling, so
for the purpose of replicating their results I do not omit them.

<<data, eval = FALSE>>=
@
<<wls>>=
# Weighted least squares regression.
stat2 <- lm(logsmr ~ aff1, weights = expct, data = lipcancer)
lipcancer$resids <- rstudent(stat2)

plot(rstudent(stat2) ~ fitted(stat2), main = 'WLS Studentized Residuals vs Fitted Values')
@

Note that the smallest distance in the dataset is
\Sexpr{sprintf('%.1f', min(dist(lipcancer[,c('x', 'y')])))} km, but their
semivariogram has three points at distances smaller than that (the caption
claims that the distances are indeed in km). It would seem like either they
plotted the semivariogram of a different dataset or they rescaled the
distances somehow. The largest distance is
\Sexpr{sprintf('%.1f', max(dist(lipcancer[,c('x', 'y')])))} km,
so I took a wild guess that they estimated the semivariogram up to about half
the largest possible distance, and used units of 10 km for some reason. I
eyeballed the distances on their plot and set manual bins, resulting in the
plot on the next page. The vertical scales are slightly different, but the
shape is very similar. At the top of page 395 they say there is negative
dependence at 72 km to 105 km, but I don't know where they see that.

<<plots, message = FALSE>>=
# Trying to recreate their empirical semivariogram.
resid_geodata <- as.geodata(lipcancer, coords.col = c('x', 'y'), data.col = 'resids')

# uvec specifies the bin centers.
resid_vario <- variog(resid_geodata, uvec = seq(0, 275, 25))
plot(resid_vario, main = 'Empirical Semivariogram')
@

Next, I checked that I understood what I was doing by fitting their
non-spatial models. The R Poisson regression results below match their SAS
results.

<<poisson>>=
# Poisson regression without adjustment for overdispersion.
PR <- glm(observed ~ aff1, family = poisson, offset = logni, data = lipcancer)
summary(PR) # Agrees!
@

\pagebreak
Their Poisson regression with overdispersion results are matched by a
quasiPoisson regression in R, as seen below.

<<quasipoisson>>=
# Poisson regression with adjustment for overdispersion.
PR_OD <- glm(observed ~ aff1, family = quasipoisson, offset = logni, data = lipcancer)
summary(PR_OD) # Agrees!
@

The nonspatial model with uncorrelated random effects and ``adjusted for
overdispersion'' is run-of-the-mill GLMM, because \texttt{glmmPQL} estimates
a within-group residual variance parameter (even though that isn't strictly a
``Poisson'' model because of the overdispersion parameter). The coefficients
and standard errors match those in the book, and squaring the random effect
and residual standard deviations gives values that are off from the variances
in the book by only about 0.02.

<<glmmi, message = FALSE>>=
# Poisson GLMM adjusted for overdispersion.
GLMMI <- glmmPQL(observed ~ aff1 + offset(logni), random = ~1|county,
                 family = poisson, data = lipcancer)
summary(GLMMI) # Coefs/SEs match, variances are close.
@

Waller and Gotway appear to construct non-overdispersed spatial models in SAS
by fixing one of the covariance parameters, but I don't understand their
\texttt{parms} statement. The effect of their \texttt{parms} statement might
be to fix the nugget (within-county variance) at 1, but that is a wild guess.
If that guess is correct, and if SAS uses the nugget as a variance multiplier
(analogous to the quasiPoisson dispersion), then this should have the correct
effect: conditional on the random intercept, the observed count has a Poisson
distribution without overdispersion.

I would construct these models by placing the correlation structure on the
random effects, instead of placing it on the ``residuals'', and estimating the
nugget. Unfortunately, I don't know of any R functions that allow that. So
I'm skipping the non-overdispersion spatial mixed models.

The spatial GLMM with overdispersion is confusing. The nugget should not be
necessary because \texttt{glmmPQL} estimates a residual variance to account
for overdispersion. I initially tried to include both parameters anyway; I
played around with the \texttt{lmeControl} options and got a variety of
cryptic error messages, and never got the optimizer to happily converge.

When leaving out the nugget and using \texttt{optim} to do the optimization,
\texttt{glmmPQL} finds estimates in only 7 reweighting iterations. The BFGS
and L-BFGS algorithms both lead to the same estimates, but they are very
different from the SAS results.

With the \texttt{nlminb} optimizer, \texttt{glmmPQL} gets stuck in a feedback
loop where the working response changes a bit every iteration, so the initial
coefficient values change, then the final coefficient estimates for the
iteration are different from the previous iteration, and so the working
responses change again. One time I ran \texttt{glmmPQL} for 10,000 iterations
and it showed no sign of converging. It's strange that it converges so easily
using \texttt{optim}, so I am suspicious about the \texttt{optim} results and
curious if SAS has convergence issues too.

<<sod>>=
# Conditional spatial GLMM adjusted for overdispersion, spherical covariance.
#corr_S_OD <- corSpher(value = c(263, 0.38), form = ~x+y, nugget = TRUE)
corr_S_OD <- corSpher(value = 263, form = ~x+y)
corr_S_OD <- Initialize(corr_S_OD, lipcancer)
S_OD <- glmmPQL(observed ~ aff1 + offset(logni), random = ~1|county,
                correlation = corr_S_OD, family = poisson,
                control = lmeControl(opt = 'optim', 'BFGS'),
                data = lipcancer)
summary(S_OD) # Way off...
@

Finally, I think the marginal model with overdispersion is conceptually the
easiest to understand of their spatial models. It just allows nearby rates to
be similar to each other, without going to the trouble of predicting a latent
mean for each county. It's a relief than \texttt{gnls} gives the same
estimates as SAS.

<<mglm>>=
# R code for MGLM.
log.model <- function(x1, expct, b0, b1){exp(b0 + log(expct) + b1 * x1)}

corr_MGLM <- corSpher(value = 53, form = ~x+y, nugget = FALSE)
corr_MGLM <- Initialize(corr_MGLM, lipcancer)
MGLM <- gnls(observed ~ log.model(aff1, expct, b0, b1),
             data = lipcancer, start = c(b0 = -0.63, b1 = 0.74),
             correlation = corr_MGLM,
             weights = varPower(form = ~fitted(.), fixed = 0.5))
summary(MGLM) # Agrees!
@

\end{enumerate}

\end{document}

### 2017-04-17 hw10/hw10.rnw
Homework 10

\documentclass[11pt]{article}
\usepackage{fullpage,graphicx,float,amsmath,enumitem,fancyhdr}

\setlist{parsep=5.5pt}
\setlength{\parindent}{0pt}

\newcommand\assignment{Stat 534 Homework 10}
\newcommand\myname{Kenny Flagg}
\newcommand\duedate{April 17, 2017}

\pagestyle{fancy}
\lhead{\assignment}
\chead{\duedate}
\rhead{\myname}
\setlength{\headheight}{18pt}
\setlength{\headsep}{2pt}

\title{\assignment}
\author{\myname}
\date{\duedate}

\begin{document}
\maketitle

<<setup, echo = FALSE, message = FALSE, cache = FALSE>>=
library(knitr)
library(extrafont)
opts_chunk$set(echo = TRUE, comment = NA, message = FALSE,
               show.signif.stars = FALSE,
               fig.align = 'center', fig.width = 6.5, fig.height = 4,
               fig.pos = 'H', size = 'footnotesize', dev = 'pdf',
               dev.args = list(family = 'CM Roman', pointsize = 11))
knit_theme$set('print')

library(xtable)
options(xtable.floating = FALSE,
        xtable.sanitize.rownames.function = function(x) x,
        xtable.sanitize.colnames.function = function(x) x,
        width = 80, scipen = 2, show.signif.stars = FALSE)

set.seed(92387)
@

\begin{enumerate}

\item%1
{\it Empirical Bayes -- NC SIDS data: This is a well-known data set on the
incidence of SIDS (Sudden Infant Death Syndrome) deaths in North Carolina. The
data are available in the \texttt{nc.sids} data set in the \texttt{spdep}
package. We will use emprical Bayes' methods to produce smoothed disease maps
of SIDS incidence in the 1974-78 period. I have been letting you do most of
the coding yourself this semester but getting the plots out for a problem like
this is too much so you can use the R code below. Briefly discuss the results.
In addition to the maps you can look at the raw SMR values and the smoothed
estimates and summarize them in some meaningful way. I am particularly
interested in counties in which the data dominate and counties in which the
empirical Bayes' priors dominate. I will send an R script file with the code
included also. I do not know what all those commands mean myself but I know
they work.}

<<prob1_1>>=
library(maptools)
library(spdep)
library(rgdal)
nc.sids <- readOGR(system.file('etc/shapes/sids.shp', package = 'spdep'), verbose = FALSE)
proj4string(nc.sids) <- CRS('+proj=longlat +ellps=clrk66')
# The IDs are just numbers...
#row.names(nc.sids) <- sapply(slot(nc.sids, 'polygons'), function(x) slot(x, 'ID'))
row.names(nc.sids) <- as.character(nc.sids$NAME) # County names.

rn <- nc.sids$FIPSNO
r <- sum(nc.sids$SID74) / sum(nc.sids$BIR74) 
Expected <- nc.sids$BIR74 * r
SMR <- nc.sids$SID74 / Expected

require(DCluster)
EB.pg <- empbaysmooth(nc.sids$SID74, Expected)$smthrr
EB.LogN <- exp(lognormalEB(nc.sids$SID74, Expected)$smthrr)
EB.Global <- EBest(nc.sids$SID74, Expected)$estmm
ncCR85_nb <- read.gal(system.file('etc/weights/ncCR85.gal',
                                  package = 'spdep' )[1], region.id = rn)
EB.Local <- EBlocal(nc.sids$SID74, Expected, ncCR85_nb)$est

nc <- cbind(nc.sids, SMR, EB.pg, EB.LogN, EB.Global, EB.Local)
names(nc) <- c(names(nc.sids), 'SMR', 'PoissonGamma.EB', 'Lognormal.EB', 'Global.EB', 'Local.EB')
row.names(nc) <- as.character(nc$NAME)
spplot(nc, c('Global.EB', 'PoissonGamma.EB', 'Local.EB', 'Lognormal.EB', 'SMR'))
@

<<prob1_2, eval = FALSE>>=
# Select 15 counties to show.
counties <- sort(c('Anson', 'Dare', sample(as.character(nc$NAME), 13)))
allSMRs <- data.frame(SMR = nc$SMR, `Local EB` = nc$Local.EB,
                      `Lognormal EB` = nc$Lognormal.EB,
                      `Poisson-Gamma EB` = nc$PoissonGamma.EB,
                      `Global EB` = nc$Global.EB,
                      row.names = nc$NAME, check.names = FALSE)
xtable(allSMRs[counties,], digits = 4)
@
<<prob1_2, echo = FALSE, results = 'asis'>>=
@
<<prob1_3, eval = FALSE>>=
meansds <- sapply(allSMRs, function(x){return(c(Min = min(x), Mean = mean(x),
                                                Max = max(x), SD = sd(x)))})
xtable(meansds, digits = 4)
@
\begin{center}
<<prob1_3, echo = FALSE, results = 'asis'>>=
@
\end{center}

All of the smoothing methods clearly get the job done, as all four methods
result in adjecent counties generally having smoothed values that are more
similar than their raw values. The Global method does the most smoothing
(having the lowest standard deviation and the smallest range of smoothed
values), but the Poisson-Gamma results are similar. The Local method does the
least smoothing (biggest SD and range). Unsurprisingly, all the smoothing
methods result in smoothed values with similar means as they all pull the
observed SMRs toward the observed mean.

Anson county is an outlier in raw SMR, and it dominates in the posterior for
each method. For counties with low observed SMRs, the priors tend to dominate
and pull the values toward the mean. The Lognormal method allows large SMRs to
remain large, but still pulls low SMRs upward. The Local method behaves
uniquely, actually smoothing some values \emph{away} from the mean when
neighboring counties had more extreme observed SMRs (e.g. Davidson and Surrey
counties). Local smoothing actually assigns one county (Dare) a smoothed value
of 0 because it and its neighbors all have observed SMRs of zero.

If I knew enough about SIDS to expect spatial autocorrelation, I would use
Local smoothing to accomodate this. Otherwise, I would use either Lognormal or
Global smoothing depending on whether or not I have prior information
suggesting that large observed SMRs are real.

\item%2
{\it Continuing with the NC SIDS data. We assume the Poisson-Gamma model.}

\begin{enumerate}

\item%a
{\it Find the posterior distribution assuming a more informative Gamma prior
with \(\alpha = \beta = 4\).}

We assume \(Y_{i}|\theta \sim \mathrm{Poisson}(E_{i}\theta)\),
\(\theta \sim \mathrm{Gamma}(4, 4)\). We observe
\(\sum_{i=1}^{n}y_{i} = \Sexpr{sum(nc$SID74)}\) SIDS cases in 1974 across all
counties. I define the exposure to be the number of births in thousands, so
the response and the exposure are on similar scales. Thus
\(\sum_{i=1}^{n}E_{i} = \Sexpr{sum(nc$BIR74) / 1000}\).

Then the posterior distribution is \(\theta|\mathbf{y} \sim \mathrm{Gamma}(
\Sexpr{sum(nc$SID74) + 4},\, \Sexpr{sum(nc$BIR74) / 1000 + 4})\).

<<prob2a1, eval = FALSE, echo = FALSE>>=
hist(nc$SID74, breaks = 20, freq = FALSE, ylim = c(0, 1),
     xlab = 'SIDS Cases', main = 'Distribution of SIDS Counts')
curve(dgamma(x, 4, 4), n = 501, lty = 2, add = TRUE)
curve(dgamma(x, sum(nc$SID74) + 4, sum(nc$BIR74) / 1000 + 4), n = 501, lty = 1, add = TRUE)
legend('topright', lty = c(2, 1), legend = c('Prior', 'Posterior'))
@

\item%b
{\it Find the posterior distribution assuming a non-informative Gamma prior
with \(\alpha = \beta = 0.1\).}

The posterior distribution is \(\theta|\mathbf{y} \sim \mathrm{Gamma}(
\Sexpr{sum(nc$SID74) + 0.1}, \Sexpr{sum(nc$BIR74) / 1000 + 0.1})\).

<<prob2a2, eval = FALSE, echo = FALSE>>=
hist(nc$SID74, breaks = 20, freq = FALSE, ylim = c(0, 1),
     xlab = 'SIDS Cases', main = 'Distribution of SIDS Counts')
curve(dgamma(x, 0.1, 0.1), n = 501, from = 0.1, lty = 2, add = TRUE)
curve(dgamma(x, sum(nc$SID74) + 0.1, sum(nc$BIR74) / 1000 + 0.1), n = 501, lty = 1, add = TRUE)
legend('topright', lty = c(2, 1), legend = c('Prior', 'Posterior'))
@

\pagebreak
\item%c
{\it Summarize the role of the data and the prior in the posterior
distributions by comparing the observed SMR and the posterior means from the
2 Bayesian approaches. A plot would help and you should be able to modify the
code above to produce those plots. Just add the vectors of the poterior
distribution to \texttt{nc.sids} using \texttt{cbind} and then use
\texttt{spplot} to get plots of SMR, and the pure Bayes' posterior means.}

Both prior distributions have mean 1, and the observed SIDS counts have mean
\Sexpr{mean(nc$SID74)}. The informative prior results in a posterior mean of
\(\Sexpr{sum(nc$SID74) + 4} / \Sexpr{sum(nc$BIR74) / 1000 + 4}
= \Sexpr{sprintf('%.3f', (sum(nc$SID74) + 4) / (sum(nc$BIR74) / 1000 + 4))}\)
and the non-informative prior yields a posterior mean of
\(\Sexpr{sum(nc$SID74) + 0.1} / \Sexpr{sum(nc$BIR74) / 1000 + 0.1} =
\Sexpr{sprintf('%.3f', (sum(nc$SID74) + 0.1) / (sum(nc$BIR74) / 1000 + 0.1))}\).
The informative prior pulls the posterior mean toward 1, but the data dominate
over the non-informative prior resulting in a posterior mean close to the
data mean.

To make a smoothed spatial map, I use the hierarchical model where
\(Y_{i}|\gamma_{i} \sim \mathrm{Poisson}(E_{i}\gamma_{i})\) and
\(\gamma_{i} \sim \mathrm{Gamma}(\alpha, \beta)\). Then the posterior
distribution for each county's mean is
\(\gamma_{i}|y_{i} \sim \mathrm{Gamma}(\alpha + y_{i}, \beta + E_{i})\). The
result is that the informative prior dominates (doing a lot of smoothing)
while the non-informative prior allows the data to dominate (resulting in a
map that is barely distinguishable from map of the data). To create maps on
SMR scale, I multiplied the posterior means by the exposure and divided by the
expected counts.

<<prob2c2>>=
nc.bayes <- cbind(nc,
                  (nc$BIR74 / 1000) * (nc$SID74 + 0.1) / (nc$BIR74 / 1000 + 0.1) / Expected,
                  (nc$BIR74 / 1000) * (nc$SID74 + 4) / (nc$BIR74 / 1000 + 4) / Expected)
names(nc.bayes) <- c(names(nc), 'Non.Informative', 'Informative')
spplot(nc.bayes, c('Non.Informative', 'Informative', 'SMR'))
@

\end{enumerate}

\pagebreak
\item%3
{\it See the notes for the details on the Global Estimator (page 85). Marshall
assumed, with \(E_{i}\) being the expected count that}
\begin{equation*}
Z_{i}|\gamma_{i} \sim \mathrm{Poi}(E_{i}\gamma_{i})
\end{equation*}
{\it but made no assumption about the distribution of the spatially varying
relative risks or SMR values \(\gamma_{i}\). We denote the estimates of SMR by
\(r_{i} = Z_{i} / E_{i}\). We make no distributional assumptions about the
\(\gamma_{i}\) values but do assume that the prior means and variances exist.
Denote them \(m_{\gamma_{i}}\) and \(v_{\gamma_{i}}\), respectively. Marshall
found method-of-moments estimators of the marginal or unconditional mean and
variance of the \(\gamma_{i}\).}

\begin{enumerate}

\item%a
{\it Find the conditional mean and variance of \(r_{i}\). That is find
\(E(r_{i}|\gamma_{i})\) and \(Var(r_{i}|\gamma_{i})\).}

\vspace{-11pt}
\begin{equation*}
E(r_{i}|\gamma_{i}) = E\left(\left.\frac{Z_{i}}{E_{i}}\right|\gamma_{i}\right)
= \frac{E(Z_{i}|\gamma_{i})}{E_{i}} = \frac{E_{i}\gamma_{i}}{E_{i}}
= \gamma_{i}
\end{equation*}
\begin{equation*}
Var(r_{i}|\gamma_{i})
= Var\left(\left.\frac{Z_{i}}{E_{i}}\right|\gamma_{i}\right)
= \frac{Var(Z_{i}|\gamma_{i})}{E_{i}^{2}}
= \frac{E_{i}\gamma_{i}}{E_{i}^{2}}
= \frac{\gamma_{i}}{E_{i}}
\end{equation*}

\item%b
{\it Show that the marginal mean of \(r_{i}\) is equal to the prior mean of
\(\gamma_{i}\).}

\vspace{-11pt}
\begin{align*}
E(r_{i}) &= E(E(r_{i}|\gamma_{i})) \\
&= \int_{0}^{\infty} E(r_{i}|\gamma_{i}) f(\gamma_{i}) d\gamma_{i} \\
&= \int_{0}^{\infty} \gamma_{i} f(\gamma_{i}) d\gamma_{i} \\
&= E(\gamma_{i})
\end{align*}

\item%c
{\it Find the marginal variance of \(r_{i}\). (Hint: it will be a function of
the prior mean and variance of \(\gamma_{i}\).)}

\vspace{-11pt}
\begin{align*}
Var(r_{i}) &= E\left((r_{i} - E(r_{i}))^{2}\right) \\
&= E\left((r_{i} - E(\gamma_{i}))^{2}\right) \\
&= E\left(r_{i}^2\right) - (E(\gamma_{i}))^{2} \\
&= E\left(E\left(r_{i}^2|\gamma_{i}\right)\right) - (E(\gamma_{i}))^{2} \\
&= \int_{0}^{\infty} E\left(r_{i}^{2}|\gamma_{i}\right)
f(\gamma_{i}) d\gamma_{i} - (E(\gamma_{i}))^{2} \\
&= \int_{0}^{\infty} \left(Var\left(r_{i}|\gamma_{i}\right)
+ \left(E\left(r_{i}|\gamma_{i}\right)\right)^{2}\right)
f(\gamma_{i}) d\gamma_{i} - (E(\gamma_{i}))^{2} \\
&= \int_{0}^{\infty} \left(\frac{\gamma_{i}}{E_{i}}
+ \gamma_{i}^{2}\right)
f(\gamma_{i}) d\gamma_{i} - (E(\gamma_{i}))^{2} \\
&= E\left(\frac{\gamma_{i}}{E_{i}}\right) + E\left(\gamma_{i}^{2}\right)
- (E(\gamma_{i}))^{2} \\
&= \frac{E(\gamma_{i})}{E_{i}} + Var(\gamma_{i})
\end{align*}

\end{enumerate}

\end{enumerate}

\end{document}
